### 2019-08-13 paper.tex
Template

\documentclass[sigplan,10pt,review,anonymous,screen]{acmart}\settopmatter{printfolios=true}

\acmConference[VMIL'19]{ACM SIGPLAN International Workshop on Virtual Machines and Intermediate Languages}{October 22, 2019}{Athens, Greece}
\acmYear{2019}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

\setcopyright{none}
\bibliographystyle{ACM-Reference-Format}

\begin{document}
\title{How slow will my transient type checks go?}

\author{Isaac Oscar Gariano}
\affiliation{
  \position{PhD Student}
  \department{School of Engineering and Computer Science} % \department is recommended
  \institution{Victoria University of Wellington}
  \country{New Zealand}
}
\email{Isaac@ecs.vuw.ac.nz} % \email is recommended

\begin{abstract}
Text of abstract \ldots.
\end{abstract}

%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
\keywords{keyword1, keyword2, keyword3}\maketitle

Text of paper \ldots

\end{document}

### 2019-08-14 paper.tex
Setup skeleton and abstract

@@
-\documentclass[sigplan,10pt,review,anonymous,screen]{acmart}\settopmatter{printfolios=true}
+\documentclass[sigplan,10pt,review,screen]{acmart}\settopmatter{printfolios=true}
 
@@
 \begin{document}
-\title{How slow will my transient type checks go?}
+\title{How slow do my transient type checks go?}
 
@@
 \affiliation{
-  \position{PhD Student}
-  \department{School of Engineering and Computer Science} % \department is recommended
+  \department{Engineering and Computer Science} % \department is recommended
   \institution{Victoria University of Wellington}
@@
 
+\author{Richard Roberts}
+\affiliation{
+  \department{Computational Media Innovation Centre} % \department is recommended
+  \institution{Victoria University of Wellington}
+  \country{New Zealand}
+}
+\email{rykardo.r@gmail.com} % \email is recommended
+
+\author{Stefan Marr}
+\affiliation{
+  \department{School of Computing} % \department is recommended
+  \institution{University of Kent}
+  \country{s.marr@kent.ac.uk}
+}
+\email{mwh@ecs.vuw.ac.nz} % \email is recommended
+
+\author{Michael Homer}
+\affiliation{
+  \department{Engineering and Computer Science} % \department is recommended
+  \institution{Victoria University of Wellington}
+  \country{New Zealand}
+}
+\email{mwh@ecs.vuw.ac.nz} % \email is recommended
+
+\author{James Noble}
+\orcid{0000-0001-9036-5692}             %% \orcid is optional
+\affiliation{
+  \department{Engineering and Computer Science} % \department is recommended
+  \institution{Victoria University of Wellington}
+  \country{New Zealand}
+}
+\email{kjx@ecs.vuw.ac.nz} % \email is recommended
+
 \begin{abstract}
-Text of abstract \ldots.
+Just-in-time compilation and optimisation in virtual machines can
+eliminate much of the overhead of transient typechecks.  Unfortunately
+the improvement is not uniform: while most typechecks can be optimised
+away, some typechecks will will significantly decrease a program's
+performance.  In this paper we investigate how benchmark results can be
+indicate which typechecks cause the most problems. Programmers could
+use these techniques to optimise their programs by removing
+typechecks, and VM engineers to identify opportunities for further
+optimisations.   
 \end{abstract}
@@
 %% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+%%KJX XML IS BROKEN!!!
 \begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011008</concept_id>
-<concept_desc>Software and its engineering~General programming languages</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10003456.10003457.10003521.10003525</concept_id>
-<concept_desc>Social and professional topics~History of programming languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
 \end{CCSXML}
-\ccsdesc[500]{Software and its engineering~General programming languages}
-\ccsdesc[300]{Social and professional topics~History of programming languages}
-\keywords{keyword1, keyword2, keyword3}\maketitle
 
-Text of paper \ldots
+\ccsdesc[500]{Software and its engineering~Just-in-time compilers}
+\ccsdesc[300]{Software and its engineering~Object oriented languages}
+\ccsdesc[300]{Software and its engineering~Interpreters}
+
+\keywords{dynamic type checking, gradual types, optional types, Grace,
+Moth, object-oriented programming}
+\maketitle
+
+\section{Introduction}
+
+intro (short \& modified)
+
+\section{Background}
+
+BG: moth \& transient typing, probably has to inclde takikawa
+
+\section{Evaluating Transient Typechecks}
+
+test protocol
+
+\section{Results}
+
+core results - i.e. looking at one scatterplot per annotation - the original version
+   - is there a good benchmark that is small enough to fit on one page. and shows interesting results?
+
+\section{Individual Annotations}
+
+   trying-to-be-cleverer results
+  - single annotation overheads
+  - using them go back to scatterplots (i.e. what you do in the latest graphs)
+
+\section{Discussion and Conclusion}
+
+  discuccion / future work / conclusion (statistics, potentially doing this for other things)
 

### 2019-08-14 paper.tex
Copypasta intro, backgroud; conclusion draft

@@
 \begin{document}
-\title{How slow do my transient type checks go?}
+\title{How Slow do my Transient Typechecks Go?}
 
@@
 
-intro (short \& modified)
+Gradual Typing aims to add static type annotations to dynamic languages, increasing
+their safety while maintaining flexibility
+\citep{GiladPluggable2004,Siek2006,XXXSiek2015}. 
+--- or, complementarily,
+to permit dynamic type annotations within static languages, increasing
+flexibility while maintaining safety
+\citep{AbadiTOPLAS1991}.
+
+There is a spectrum of different approaches to gradual typing
+\cite{kafka18,bensurvey18icfp}.
+At one end\,---\,``pluggable types'' as in Strongtalk \cite{strongtalk} or ``erasure
+semantics'' as in 
+%% Two important approaches are optional\citep{}
+%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
+%% These are applied to dynamic languages to reap the benefits of typing, but
+%% unfortunately also have limitations.
+%With optional or pluggable approaches such as 
+%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
+%2018}{https://www.typescriptlang.org/}
+TypeScript\citep{typeScriptECOOP}\,---\,
+%
+all types are erased before the execution, limiting the benefit of
+types to the statically typed parts of programs, and preventing
+programs from depending on type checks at run time.  In the middle,
+``transient'' or ``type-tag'' checks as in Reticulated Python 
+% and ``concrete'' checks as in Thorn
+offer first-order semantics, checking
+whether an object's type constructor or supported methods match
+explicit type declarations
+%%%WRONG kjx
+%, and checking only when data flows through those declarations
+\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
+Reticulated Python also supports an alternative ``monotonic'' semantics
+which mutates an object to narrow its concrete type when it is passed
+into a more specific type context.
+At the other end of the spectrum, behavioral
+typechecks as in Typed Racket \cite{typedScheme08,takikawa2012},
+Gradualtalk \cite{gradualtalk14},
+and Reticulated Python's proxies,
+support higher-order semantics, retaining
+types until run 
+time, performing the checks eagerly, and giving detailed information
+about type violations as soon as possible via blame
+tracking \cite{blame2009,blameForAll2011}.
+Finally, Ductile typing
+dynamically interprets a static type system at runtime
+\cite{Ductile2011}.
+%
+%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
+%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
+%  after concrete and before behavioural. Does it matter?}
+%
+Unfortunately, any gradual system with run-time semantics
+(i.e.\ everything more complex than erasure) currently
+imposes a significant run-time performance overhead to provide those
+semantics.  This has lead to a significant body of research,
+on one hand developing techniques to optimise gradual typing 
+\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
+and on the other hand developing techniques to evaluate the
+resulting performance  \cite{takikawa2016,Greenman2019jfp}.
+
+This paper builds upon our recent work on optimising transient
+typechecks \cite{Roberts2017,roberts-and-co-ecoop-2019} to
+investigate how benchmark results can be repurposed to identify which
+particular transient typechecks are resistant to optimisation.
+The next section discusses dynamic typechecks and gradual typing in
+Grace, then section~\ref{s-eval} describes the benchmarking
+protocols. Section~\ref{s-overall} then presents the results of
+analysing overall benchmark results, while section~\ref{s-individual}
+looks at the results of benchmarking individual type checks.
+Finally~\ref{s-concl} discusses our results, briefly considers threats
+to validity, and concludes.
 
 \section{Background}
+\label{s-bg}
 
-BG: moth \& transient typing, probably has to inclde takikawa
+Our work is based on Moth virtual machine 
+\cite{Roberts2017,roberts-and-co-ecoop-2019}
+an implementation
+of the Grace programming language 
+\citep{graceOnward12,graceSigcse13},
+based on the Graal and Truffle toolchain
+\cite{Wurthinger:2017:PPE,Wurthinger2013}
+and developed from a Newspeak implementation based on the  Simple
+Object Machine \cite{Daloze2016,SOMns}.
+
+Grace is an object-oriented, imperative, educational programming
+language, with a focus on introductory programming
+courses, but also intended for more advanced study and research \citep{graceOnward12,graceSigcse13}.
+%
+While Grace's syntax draws
+from the so-called ``curly bracket'' tradition of C, Java, and
+JavaScript, the structure of the language
+is in many ways closer to Smalltalk:
+all computation is via dynamically dispatched  ``method requests''
+where the object receiving the request decides which code to run,
+and
+%
+returns within lambdas that are ``non-local'', returning to the method
+activation in which the block is instantiated \citep{bluebook}.  In
+other ways, Grace is closer to JavaScript than Smalltalk: Grace
+objects can be created from object literals, rather than by
+instantiating classes \citep{Black2007-emeraldHOPL,JonesECOOP2016} and
+objects and classes can be deeply nested within each 
+other \citep{betabook}.
+
+Critically, Grace's declarations and methods' arguments
+and results can be annotated with types, and those types can be  checked
+either statically or dynamically. This means the type system is
+intrinsically gradual:%
+%
+~type annotations should not affect the semantics of a correct
+program\citep{XXXSiek2015}, and the type system
+includes a distinguished ``{Unknown}'' type which matches any other type
+and is the implicit type for untyped program parts.
+
+The static core of Grace's type system is well described
+elsewhere\citep{TimJonesThesis};
+here we explain how
+these types can be understood 
+dynamically, from the Grace programmer's point of view.
+Grace's types are structural \citep{graceOnward12},
+that is, an object implements a type whenever it
+implements all the methods required by that type,
+rather than requiring classes or objects to declare types explicitly.
+Methods match when they have the same name and arity:
+argument and return types are ignored.
+A type thus expresses the requests an object can respond to,
+for example whether a particular accessor is available,
+rather than a nominal location in an explicit inheritance hierarchy.
+
+Grace then checks the types of values at run time:
+%
+\begin{itemize}
+\item the values of arguments are checked after a method is requested, 
+      but before the body of the method is executed;
+\item the value returned by a method is checked after its body is executed; and
+\item the values of variables are checked
+      whenever written or read by user code.
+\end{itemize}
+
+Moth is a virtual machine for Grace that we are developing as a
+research platform \cite{roberts-and-co-ecoop-2019}. Like other VMs
+based on the Truffle and Graal toolchain, Moth is a self-optimising
+AST interpreter \cite{Wurthinger:2012:SelfOptAST}. 
+The key idea is that an AST rewrites itself based on a program's run-time values
+to reflect the minimal set of operations needed to execute the program
+correctly. The rewritten AST is then compiled into efficient machine
+code. This rewriting often depends on the dyanamic types of the
+objects involved. In the simplest case, a ``self'' call (when one method
+on an object invokes a second method on the exact same object) will
+always result in executing the second method defined in same concrete
+class as the first method. The second method can then be inlined into
+the first method, avoiding the overheads of a machine-level subroutine
+invocation and of an object-oriented dynamic dispatch.
+
+Moth relies on a number of standard techniques for optimising
+object-oriented programs.
+''Shapes'' \citep{woss2014object} capture information about objects'
+structure and (runtime) 
+field types, allowing a just-in-time compiler to
+represent objects in memory similarly to C structs, and so to 
+generate highly efficient code.
+``Polymorphic inline caches''
+\citep{Hoelzle:91:PIC} uses object shapes to cache the results of
+method lookups, avoiding expensive class hierarchy searches or
+indirect jumps through virtual method table. 
+For languages built on the Truffle framework,
+Graal comes with  additional support for partial evaluation,
+which enables efficient native code generation for
+Truffle interpreters\citep{Wurthinger:2017:PPE}.
 
 \section{Evaluating Transient Typechecks}
+\label{s-eval}
 
@@
 \section{Results}
+\label{s-overall}
 
-core results - i.e. looking at one scatterplot per annotation - the original version
+core results - i.e.\ looking at one scatter plot per annotation - the original version
    - is there a good benchmark that is small enough to fit on one page. and shows interesting results?
@@
 \section{Individual Annotations}
+\label{s-individual}
 
@@
 \section{Discussion and Conclusion}
+\label{s-concl}
 
-  discuccion / future work / conclusion (statistics, potentially doing this for other things)
+In this paper we have investigated how benchmark results can be
+repurposed to determine precisely which transient typechecks are
+resistant to the optimisations provided by a just-in-time compiling
+virtual machine.  We observed that many of our benchmark results
+conducted under the Takikawa protocol 
+seemed to have a characteristic bimodal performance profile: some
+of the benchmark configurations ran significantly slower than
+the remaining configurations. By inspecting the benchmark graphs coded
+for each typecheck in a program's source code, we were often able to
+identify that just one or two of the typechecks were responsible for
+the bimodal performance.  Based on that result, we also investigated
+individual benchmark configurations for each individual typecheck:
+these results seem to identify problematic typechecks in some but not
+all cases.
+
+This is preliminary work: there are a number of threats to validity. 
+Regarding construct validity, 
+our underlying
+implementation may contain undetected bugs that affect the semantics
+or performance of the gradual typing checks. Regarding internal
+validity,
+our benchmarking harness run on the same implementation is
+subject to the same issues. Regarding external validity, 
+Moth is built
+on the Truffle and Graal toolchain, so we expect
+to other Graal
+VMs doing similar AST-based optimizations of transient typechecks.
+Because we rely on common techniques, 
+we expect our results should be transferable to JIT implementations
+more widely.
+
+Finally, it is not clear how out results would transfer
+to other gradually typed-languages or other semantics for gradual typing.
+Our benchmarks do not depend on any features of Grace
+that are not common in other gradually-typed object-oriented
+languages, but as Grace lacks a large corpus of programs the
+benchmarks are necessarily small and artificial.
+The advantage of Grace (and Moth) for this research is
+that their relative simplicity means we have been able to build an
+implementation that features competitive performance with significantly less
+effort than would be required for larger and more complex languages.
+
+In the future, we hope to investigate statistical techniques to
+determine the significance of each type annotation's contribution to
+programs overall performance.  We would also like to investigate
+
+\bibliography{references}
 

### 2019-08-14 paper.tex
Copypasta evaluation - just need results now

@@
 \startPage{1}
+
+\def\AWFY{Are\,We\,Fast\,Yet\xspace}
+
+\usepackage{listings}
+
+\lstdefinestyle{mystyle}{
+    language=SAS,
+    breakatwhitespace=true,
+    breaklines=true,
+    captionpos=b,
+    keepspaces=true,
+    numbers=left,
+    numbersep=5pt,
+    showspaces=false,
+    showstringspaces=false,
+    showtabs=false,
+    tabsize=2,
+    aboveskip=20pt,
+    belowskip=10pt,
+    xleftmargin=0.5cm,
+    basicstyle=\footnotesize\ttfamily,
+    commentstyle=\itshape\color{codegray},
+    numberstyle=\footnotesize\color{codegray},
+    stringstyle=\color{codegreen},
+    keywordstyle = {\color{codepurple}},
+    keywordstyle = [2]{\color{codered}},
+    keywordstyle = [3]{\color{codered}},
+    keywordstyle = [4]{\color{codegreen}},
+    keywords={method,object,interface,type,var,def,return,class,for,in,if},
+    otherkeywords = {:,->},
+    morekeywords = [2]{:},
+    morekeywords = [3]{->},
+    morekeywords = [4]{true,false},
+    morestring=*[d]{"},
+    backgroundcolor={}
+}
+
+\lstset{style=mystyle}
 
@@
 correctly. The rewritten AST is then compiled into efficient machine
-code. This rewriting often depends on the dyanamic types of the
+code. This rewriting often depends on the dynamic types of the
 objects involved. In the simplest case, a ``self'' call (when one method
@@
 
-test protocol
+In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the
+performance of transient typechecks using the \AWFY
+benchmark suite, plus other
+benchmarks from the gradual-typing literature.
+The goal was to complement our benchmarks with additional ones that are
+used for similar experiments and can be ported to Grace.
+To this end, we surveyed a number of papers\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
+and selected benchmarks that have been used by multiple papers.
+Some of these benchmarks overlapped with the \AWFY suite,
+or were available in different versions.
+While not always behaviorally equivalent,
+we chose the \AWFY versions since we already used them to
+establish the performance baseline.
+The selected benchmarks as well as the papers in which they were used are shown in
+\cref{tab:gradual-benchmarks}.
+
+\begin{table}[htb]
+  \caption{Benchmarks selected from literature.}
+  \label{tab:gradual-benchmarks}
+  \begin{center}
+    \begin{tabular}{l l r}
+      Fannkuch & \cite{Vitousek2017,Greenman2018} \\
+      Float & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
+      Go & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
+      NBody & \cite{Kuhlenschmidt:2018:preprint,Vitousek2017,Greenman2018} & used \cite{Marr2016} \\
+      Queens & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} & used \cite{Marr2016} \\
+      PyStone & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
+      Sieve & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} & used \cite{Marr2016} \\
+      Snake & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} \\
+      SpectralNorm & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
+    \end{tabular}
+  \end{center}
+\end{table}
+
+The benchmarks were modified to have complete type information.  To
+ensure correctness and completeness of these experiments, we added an
+additional check to Moth that reports absent type information to
+ensure each benchmark is completely typed.  To assess the performance
+overhead of type checking, we compared the execution of Moth with all
+checks disabled against an execution that has all checks enabled.  The
+results of this evaluation demonstrated that Moth run programs with
+transient typechecks enabled at an overhead of of 5\% (min. −13\%,
+max. 79\%) over peak performance without typechecks.
+
+We did not measure programs that mix typed and untyped code because
+with our implementation technique a fully typed program is expected to
+have the largest overhead. This is not the case for other kinds of
+gradual type checking, which have given rise to more complex
+evaluation techniques.  In particular, so-called ``Takikawa'' or
+``Takikawa-Greenman'' evaluation \cite{takikawa2016,Greenman2019jfp}
+benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
+benchmark. where a configuration is a particular mix of static and
+dynamically typed code.
+
+The Takikawa evaluation protocol was originally proposed for Typed
+Racket, where static vs dynamic typing is set per module, so $N$ is
+the number of modules. Grace, following other languages such as
+Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
+allows programmers to choose typechecks at the level of individual
+declarations: $N$ is the number of type annotations in the program.,
+so checking an entire lattice for a benchmark of any size
+is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
+these kinds of langauges by taking an approach based on sampling
+\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
+divides the number of type annotations in an fully-typed program into
+up to 100 intervals, and then randomly generates ten programs within
+each interval by erasing type annotations.
+
+To investigate Moth's support for transient typechecks further, we
+carried out a (small?) Takikawa-Vitousek evaluation using (some of?)
+our existing benchmarks.
+
+\textbf{
+!!!KJX: Need to update this paragraph with what we did this time around!!!.
+To account for the complex warmup behavior
+of modern systems\citep{Barrett:2017:VMW} as well as
+the non-determinism caused by \eg garbage collection and cache effects,
+we run each benchmark for \NumIterationsAll iterations in the same
+VM invocation.\footnote{
+For the Higgs VM, we only use \NumIterationsHiggs iterations,
+because of its lower performance.
+This is sufficient since Higgs's compilation approach induces less variation
+and leads to more stable measurements.}
+Afterwards, we inspected the run-time plots over the iterations
+and manually determined a cutoff of \WarmupCutOff iterations for warmup,
+\ie, we discard iterations with signs of compilation.
+As a result, we use a large number of data points to compute the average,
+but outliers, caused by \eg garbage collection, remain visible in the plots.
+% In this work, we do not consider startup performance,
+% because we want to assess the impact of gradual type checks
+% on the best possible performance.
+All reported averages use the geometric mean since they aggregate ratios.
+
+% Yuria
+%  - Ubuntu 16.04.4, Kernel 3.13
+%  - 24 hyperthreads
+%  - Intel Xeon E5-2620 v3 2.40GHz
+% Graal 0.33 Feb. 2018
+%
+% TODO: update these details when we have time... not a rejection criterion,
+% but needs to be correct in the final version
+All experiments were executed on a machine running Ubuntu Linux 16.04.4,
+with Kernel 3.13.
+The machine has two Intel Xeon E5-2620 v3 2.40GHz,
+with 6 cores each, for a total of 24 hyperthreads.
+We used ReBench 0.10.1\citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
+Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).
+Benchmarks were executed one by one to avoid interference between them.
+The analysis of the results was done with R 3.4.1,
+and plots are generated with ggplot 2.2.1 and tikzDevice 0.11.
+Our experimental setup is available online to enable reproductions.\footnote{
+  \smtodo{merge changes, and tag final version}
+\url{https://github.com/gracelang/moth-benchmarks}}
+}
 
@@
 
+give overall resuslts - one page of monochromatic scatterplots one for
+each benchmark.
+
+need to explKn how to read scatterplots by paraphrasing
+\cite{vitousek-transient-arXive-2019}
+
 core results - i.e.\ looking at one scatter plot per annotation - the original version
-   - is there a good benchmark that is small enough to fit on one page. and shows interesting results?
+   - is there a good benchmark that is small enough to fit on one
+   page. and shows interesting results?
 
@@
 
+We hypothesised that the overhead of an individual type annotation may
+be able to predict the performance of program configurations with that
+annotation. To test this hypothesis, we constructed a series of
+configurations, one for each type annotation, with only that
+one annotation included.
+
    trying-to-be-cleverer results
+
    - single annotation overheads
   - using them go back to scatterplots (i.e. what you do in the latest graphs)
+
+  - show that single annotation prediction doesn't reallty work\ldots
 

### 2019-08-14 paper.tex
Spellchecked and compiles

@@
 \def\AWFY{Are\,We\,Fast\,Yet\xspace}
+\newcommand{\code}[1]{\texttt{#1}}
+\newcommand{\ie}{i.e.\xspace}
+\newcommand{\eg}{e.g.\xspace}
 
 \usepackage{listings}
+\usepackage{xspace}
+
+\definecolor{codegreen}{rgb}{0,0.6,0}
+\definecolor{codegray}{rgb}{0.5,0.5,0.5}
+\definecolor{codepurple}{rgb}{0.58,0,0.82}
+\definecolor{codered}{rgb}{0.82,0.15,0.23}
+\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
 
@@
 and on the other hand developing techniques to evaluate the
-resulting performance  \cite{takikawa2016,Greenman2019jfp}.
+resulting performance  \cite{Takikawa2016,Greenman2019jfp}.
 
@@
 rather than requiring classes or objects to declare types explicitly.
-Methods match when they have the same name and arity:
+Grace's transient dynamic typechecks are first order, so methods match when they have
+the same name and arity: 
 argument and return types are ignored.
@@
 \end{itemize}
+
+To see how this works in practice, consider this code snippet, which
+will run in Grace.
+
+\begin{lstlisting}
+def o = object {
+   method three -> Number {3} }
+type ThreeString = interface {
+   three -> String
+}
+def t : ThreeString = o
+printNumber (t.three)
+\end{lstlisting}
+
+Dynamic typechecks will take place: on line 6, when object \code{o}
+initialises the variable \code{t}, we check it has the method
+\code{three}; on line 7, when the value of \code{t} is read, it must
+again have the method \code{three}; on line 2, when the \code{three}
+method requested by ``\code{t.three}'' returns, the return value must
+be a \code{Number}; and (presumably) within the definition of 
+%
+\code{printNumber(n :   Number)} 
+%
+(not shown) to check that its argument is a number. Note that we never check
+whether e.g.\ the result of requesting ``\code{t.three}'' is actually
+a \code{String} (as one may expect from line 4) because we check
+only first-order types (whether objects have conforming methods) not higher-order
+types (whether those methods' argument and result types conform), and
+we check only when values flow through explicit type annotations.
+This is why the type declared in lines 3-5 is checked only on line 6
+(where it is mentioned explicitly); and the check only requires the
+presence of a method called \code{three}, regardless of the method's
+declared return type.
 
@@
 The selected benchmarks as well as the papers in which they were used are shown in
-\cref{tab:gradual-benchmarks}.
+Table~\ref{tab:gradual-benchmarks}.
 
@@
 results of this evaluation demonstrated that Moth run programs with
-transient typechecks enabled at an overhead of of 5\% (min. −13\%,
+transient typechecks enabled at an overhead of of 5\% (min.\ -13\%,
 max. 79\%) over peak performance without typechecks.
@@
 evaluation techniques.  In particular, so-called ``Takikawa'' or
-``Takikawa-Greenman'' evaluation \cite{takikawa2016,Greenman2019jfp}
+``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
 benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
@@
 is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
-these kinds of langauges by taking an approach based on sampling
+these kinds of languages by taking an approach based on sampling
 \cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
@@
 
-\textbf{
-!!!KJX: Need to update this paragraph with what we did this time around!!!.
+\textbf{!!!KJX: Need to update this paragraph with what we did this time around!!!}
 To account for the complex warmup behavior
@@
 the non-determinism caused by \eg garbage collection and cache effects,
-we run each benchmark for \NumIterationsAll iterations in the same
+we run each benchmark for \textbf{$/$NumIterationsAll} iterations in the same
 VM invocation.\footnote{
-For the Higgs VM, we only use \NumIterationsHiggs iterations,
+For the Higgs VM, we only use \textbf{$/$NumIterationsHiggs} iterations,
 because of its lower performance.
@@
 Afterwards, we inspected the run-time plots over the iterations
-and manually determined a cutoff of \WarmupCutOff iterations for warmup,
+and manually determined a cutoff of \textbf{$/$WarmupCutOff} iterations for warmup,
 \ie, we discard iterations with signs of compilation.
@@
 % on the best possible performance.
-All reported averages use the geometric mean since they aggregate ratios.
+All reported averages use the geometric mean since they aggregate
+ratios.
 
+\textbf{
 % Yuria
@@
 Our experimental setup is available online to enable reproductions.\footnote{
-  \smtodo{merge changes, and tag final version}
 \url{https://github.com/gracelang/moth-benchmarks}}
@@
 
-give overall resuslts - one page of monochromatic scatterplots one for
+give overall results - one page of monochromatic scatterplots one for
 each benchmark.
@@
 
-  - show that single annotation prediction doesn't reallty work\ldots
+  - show that single annotation prediction doesn't really work\ldots
 

### 2019-08-14 paper.tex
Make various typo fixes

@@
 
-Our work is based on Moth virtual machine 
-\cite{Roberts2017,roberts-and-co-ecoop-2019}
+Our work is based on the Moth virtual machine 
+\cite{Roberts2017,roberts-and-co-ecoop-2019},
 an implementation
@@
 %
-returns within lambdas that are ``non-local'', returning to the method
+returns within lambdas are ``non-local'', returning to the method
 activation in which the block is instantiated \citep{bluebook}.  In
@@
 
-The static core of Grace's type system is well described
+The static core of Grace's type system is well-described
 elsewhere\citep{TimJonesThesis};
@@
 rather than requiring classes or objects to declare types explicitly.
-Grace's transient dynamic typechecks are first order, so methods match when they have
+% This should say Moth's, not Grace's, but I know there is objection to that
+Grace's transient dynamic typechecks are first-order, so methods match when they have
 the same name and arity: 
@@
 
+% Ditto
 Grace then checks the types of values at run time:
@@
 checks disabled against an execution that has all checks enabled.  The
-results of this evaluation demonstrated that Moth run programs with
+results of this evaluation demonstrated that Moth runs programs with
 transient typechecks enabled at an overhead of of 5\% (min.\ -13\%,
@@
 benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
-benchmark. where a configuration is a particular mix of static and
+benchmark, where a configuration is a particular mix of static and
 dynamically typed code.
@@
 The Takikawa evaluation protocol was originally proposed for Typed
-Racket, where static vs dynamic typing is set per module, so $N$ is
+Racket, where static vs dynamic typing is set per-module, so $N$ is
 the number of modules. Grace, following other languages such as
@@
 on the Truffle and Graal toolchain, so we expect
-to other Graal
+to resemble other Graal
 VMs doing similar AST-based optimizations of transient typechecks.

### 2019-08-14 paper.tex
Fix my country/email, added orcid
Signed-off-by: Stefan Marr <git@stefan-marr.de>
Added collab.sty to be able to comment conveniently
Signed-off-by: Stefan Marr <git@stefan-marr.de>
Edit and comment abstract
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 \usepackage{xspace}
+
+\usepackage{collab}
+
+\collabAuthor{rr}{green!60!black}{Richard}
+\collabAuthor{sm}{red}{Stefan}
+\collabAuthor{kjx}{orange}{James}
+\collabAuthor{mwh}{purple}{Michael}
 
@@
 \author{Stefan Marr}
+\orcid{0000-0001-9059-5180}
 \affiliation{
@@
   \institution{University of Kent}
-  \country{s.marr@kent.ac.uk}
+  \country{United Kingdom}
 }
-\email{mwh@ecs.vuw.ac.nz} % \email is recommended
+\email{s.marr@kent.ac.uk} % \email is recommended
 
@@
 the improvement is not uniform: while most typechecks can be optimised
-away, some typechecks will will significantly decrease a program's
-performance.  In this paper we investigate how benchmark results can be
-indicate which typechecks cause the most problems. Programmers could
+away, some typechecks will significantly decrease a program's
+performance.
+
+In this paper we investigate how benchmark results can be
+used to identify which typechecks cause performance issues. Programmers could
 use these techniques to optimise their programs by removing
@@
 optimisations.
+
+\sm{this is much to brief for me.
+Transient typechecks is a very specific thing.
+I believe, VM people need more context here.
+It's a form of gradual typing, which requires comparably simple checks
+at run time to check type violations.
+Then, how do we investigate the issues?
+we should give a hint what the solution is.
+otherwise, the sentence in using these techniques doesn't make sense to me, because we didn't say anything about them.}
 \end{abstract}

### 2019-08-14 paper.tex
Make spelling of `run time` consistent with rest of paper
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 Finally, Ductile typing
-dynamically interprets a static type system at runtime
+dynamically interprets a static type system at run time
 \cite{Ductile2011}.
@@
 ''Shapes'' \citep{woss2014object} capture information about objects'
-structure and (runtime) 
+structure and (run time) 
 field types, allowing a just-in-time compiler to

### 2019-08-14 paper.tex
Superfluous dots and missing spaces
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 their safety while maintaining flexibility
-\citep{GiladPluggable2004,Siek2006,XXXSiek2015}. 
+\citep{GiladPluggable2004,Siek2006,XXXSiek2015}
 --- or, complementarily,
@@
 %2018}{https://www.typescriptlang.org/}
-TypeScript\citep{typeScriptECOOP}\,---\,
+TypeScript
+\citep{typeScriptECOOP}\,---\,
 %

### 2019-08-14 paper.tex
Added structure to background
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
+\paragraph{Grace's Typeing}
 Critically, Grace's declarations and methods' arguments
@@
 
+\paragraph{Moth: A Grace Implementation}
 Moth is a virtual machine for Grace that we are developing as a

### 2019-08-14 paper.tex
Fix version of ReBench
Signed-off-by: Stefan Marr <git@stefan-marr.de>
Add spaces before \cite
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 ~type annotations should not affect the semantics of a correct
-program\citep{XXXSiek2015}, and the type system
+program \citep{XXXSiek2015}, and the type system
 includes a distinguished ``{Unknown}'' type which matches any other type
@@
 The static core of Grace's type system is well-described
-elsewhere\citep{TimJonesThesis};
+elsewhere \citep{TimJonesThesis};
 here we explain how
@@
 object-oriented programs.
-''Shapes'' \citep{woss2014object} capture information about objects'
+``Shapes'' \citep{woss2014object} capture information about objects'
 structure and (run time) 
@@
 which enables efficient native code generation for
-Truffle interpreters\citep{Wurthinger:2017:PPE}.
+Truffle interpreters \citep{Wurthinger:2017:PPE}.
 
@@
 used for similar experiments and can be ported to Grace.
-To this end, we surveyed a number of papers\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
+To this end, we surveyed a number of papers
+\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
 and selected benchmarks that have been used by multiple papers.
@@
 with 6 cores each, for a total of 24 hyperthreads.
-We used ReBench 0.10.1\citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
+We used ReBench 1.0\citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
 Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).

### 2019-08-14 paper.tex
Minor ‘typos’
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 allows programmers to choose typechecks at the level of individual
-declarations: $N$ is the number of type annotations in the program.,
+declarations: $N$ is the number of type annotations in the program,
 so checking an entire lattice for a benchmark of any size
-is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
+is infeasible.  Vitousek et al.\ modified the Takikawa protocol for
 these kinds of languages by taking an approach based on sampling
@@
 with 6 cores each, for a total of 24 hyperthreads.
-We used ReBench 1.0\citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
+We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
 Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).

### 2019-08-14 paper.tex
Minor phrasing changes and two comments

--- 2019-08-14 paper.tex
@@
 
+\newcommand{\RR}[1]{{\color{red}RR: #1}}
+
 \usepackage{listings}
 \usepackage{xspace}
-
-\usepackage{collab}
-
-\collabAuthor{rr}{green!60!black}{Richard}
-\collabAuthor{sm}{red}{Stefan}
-\collabAuthor{kjx}{orange}{James}
-\collabAuthor{mwh}{purple}{Michael}
 
@@
 \author{Stefan Marr}
-\orcid{0000-0001-9059-5180}
 \affiliation{
@@
   \institution{University of Kent}
-  \country{United Kingdom}
+  \country{s.marr@kent.ac.uk}
 }
-\email{s.marr@kent.ac.uk} % \email is recommended
+\email{mwh@ecs.vuw.ac.nz} % \email is recommended
 
@@
 away, some typechecks will significantly decrease a program's
-performance.
-
-In this paper we investigate how benchmark results can be
-used to identify which typechecks cause performance issues. Programmers could
-use these techniques to optimise their programs by removing
-typechecks, and VM engineers to identify opportunities for further
-optimisations.
-
-\sm{this is much to brief for me.
-Transient typechecks is a very specific thing.
-I believe, VM people need more context here.
-It's a form of gradual typing, which requires comparably simple checks
-at run time to check type violations.
-Then, how do we investigate the issues?
-we should give a hint what the solution is.
-otherwise, the sentence in using these techniques doesn't make sense to me, because we didn't say anything about them.}
+performance.  In this paper we investigate how benchmark results
+indicate which typechecks cause the most problems.
+Using these techniques,
+programmers can optimise programs by removing typechecks,
+and VM engineers can identify opportunities for compiler optimisation.   
 \end{abstract}
@@
 \citep{GiladPluggable2004,Siek2006,XXXSiek2015}
---- or, complementarily,
+or, complementarily,
 to permit dynamic type annotations within static languages, increasing
@@
 %2018}{https://www.typescriptlang.org/}
-TypeScript
-\citep{typeScriptECOOP}\,---\,
+TypeScript\citep{typeScriptECOOP}\,---\,
 %
@@
 Finally, Ductile typing
-dynamically interprets a static type system at run time
+dynamically interprets a static type system at runtime
 \cite{Ductile2011}.
@@
 
+\subsection{Grace's Transient Type Checking}
+
 Grace is an object-oriented, imperative, educational programming
@@
 other ways, Grace is closer to JavaScript than Smalltalk: Grace
-objects can be created from object literals, rather than by
+objects are created from object literals, rather than by
 instantiating classes \citep{Black2007-emeraldHOPL,JonesECOOP2016} and
@@
 
-\paragraph{Grace's Typeing}
-Critically, Grace's declarations and methods' arguments
+Grace's declarations and methods' arguments
 and results can be annotated with types, and those types can be  checked
@@
 program\citep{XXXSiek2015}, and the type system
-includes a distinguished ``{Unknown}'' type which matches any other type
-and is the implicit type for untyped program parts.
+includes a distinguished ``{Unknown}'' type which matches any type
+and is the implicit type for untyped parts of the program.
 
@@
 implements all the methods required by that type,
-rather than requiring classes or objects to declare types explicitly.
+rather than requiring classes or objects to declare types explicitly \RR{(explicitly or nominally)?}. 
 % This should say Moth's, not Grace's, but I know there is objection to that
 Grace's transient dynamic typechecks are first-order, so methods match when they have
-the same name and arity: 
+the same name and arity, while
 argument and return types are ignored.
@@
 
+\begin{minipage}{\linewidth}
 \begin{lstlisting}
 def o = object {
-   method three -> Number {3} }
+   method three -> Number {3}
+}
 type ThreeString = interface {
@@
 \end{lstlisting}
+\end{minipage}
 
-Dynamic typechecks will take place: on line 6, when object \code{o}
-initialises the variable \code{t}, we check it has the method
-\code{three}; on line 7, when the value of \code{t} is read, it must
-again have the method \code{three}; on line 2, when the \code{three}
-method requested by ``\code{t.three}'' returns, the return value must
-be a \code{Number}; and (presumably) within the definition of 
+Dynamic typechecks will take place:
+
+\begin{itemize}
+
+\item on line 7,
+when object \code{o} initialises the variable \code{t},
+we check it has the method \code{three};
+
+\item on line 8,
+when the value of \code{t} is read,
+it must again have the method \code{three};
+
+\item on line 2,
+when the \code{three} method requested by ``\code{t.three}'' returns,
+the return value must be a \code{Number};
+and (presumably) within the definition of
 %
@@
 %
-(not shown) to check that its argument is a number. Note that we never check
+(not shown) to check that its argument is a number. 
+\end{itemize}
+
+Note that we never check
 whether e.g.\ the result of requesting ``\code{t.three}'' is actually
-a \code{String} (as one may expect from line 4) because we check
+a \code{String} (as one may expect from line 5) because we check
 only first-order types (whether objects have conforming methods) not higher-order
@@
 we check only when values flow through explicit type annotations.
-This is why the type declared in lines 3-5 is checked only on line 6
+This is why the type declared in lines 4-6 is checked only on line 7
 (where it is mentioned explicitly); and the check only requires the
@@
 
-\paragraph{Moth: A Grace Implementation}
+\subsection{Moth}
+
 Moth is a virtual machine for Grace that we are developing as a
@@
 object-oriented programs.
-''Shapes'' \citep{woss2014object} capture information about objects'
-structure and (run time) 
+``Shapes'' \citep{woss2014object} capture information about objects'
+structure and (runtime) 
 field types, allowing a just-in-time compiler to
-represent objects in memory similarly to C structs, and so to 
-generate highly efficient code.
+represent objects in memory similarly to C structs and, consequently,
+can generate highly efficient code.
 ``Polymorphic inline caches''
@@
 method lookups, avoiding expensive class hierarchy searches or
-indirect jumps through virtual method table. 
+indirect jumps through virtual method tables. 
 For languages built on the Truffle framework,
@@
 Truffle interpreters\citep{Wurthinger:2017:PPE}.
+\RR{This last sentence seems out of place, does it belong here?}
 
@@
 performance of transient typechecks using the \AWFY
-benchmark suite, plus other
+benchmark suite \cite{Marr2016}, plus other
 benchmarks from the gradual-typing literature.
@@
 
-We did not measure programs that mix typed and untyped code because
+We did not measure programs that mixed typed and untyped code because
 with our implementation technique a fully typed program is expected to
@@
 gradual type checking, which have given rise to more complex
-evaluation techniques.  In particular, so-called ``Takikawa'' or
+evaluation techniques.  In particular, the so-called ``Takikawa'' or
 ``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
@@
 allows programmers to choose typechecks at the level of individual
-declarations: $N$ is the number of type annotations in the program.,
+declarations: $N$ is the number of type annotations in the program,
 so checking an entire lattice for a benchmark of any size
@@
 divides the number of type annotations in an fully-typed program into
-up to 100 intervals, and then randomly generates ten programs within
+a maximum of 100 intervals, and then randomly generates ten programs within
 each interval by erasing type annotations.
@@
 individual benchmark configurations for each individual typecheck:
-these results seem to identify problematic typechecks in some but not
-all cases.
+these results can identify problematic typechecks, but not all cases.
 
@@
 validity,
-our benchmarking harness run on the same implementation is
-subject to the same issues. Regarding external validity, 
+our benchmarking harness runs on the same implementation
+and therefore is subject to the same issues.
+Regarding external validity, 
+\RR{The next two sentences seem to be in the wrong place (they state an advantage rather than a threat); maybe move them into the previous paragraph?}
 Moth is built
@@
 benchmarks are necessarily small and artificial.
-The advantage of Grace (and Moth) for this research is
+The advantage of Grace and Moth for this research is
 that their relative simplicity means we have been able to build an
@@
 In the future, we hope to investigate statistical techniques to
-determine the significance of each type annotation's contribution to
+determine the significance of each type annotation's contribution to a
 programs overall performance. We would also like to investigate

### 2019-08-15 paper.tex

@@
 and VM engineers can identify opportunities for compiler optimisation.
+
+\sm{this is much to brief for me.
+Transient typechecks is a very specific thing.
+I believe, VM people need more context here.
+It's a form of gradual typing, which requires comparably simple checks
+at run time to check type violations.
+Then, how do we investigate the issues?
+we should give a hint what the solution is.
+otherwise, the sentence in using these techniques doesn't make sense to me, because we didn't say anything about them.}
 \end{abstract}
@@
 \citep{GiladPluggable2004,Siek2006,XXXSiek2015}
-or, complementarily,
+--- or, complementarily,
 to permit dynamic type annotations within static languages, increasing
@@
 
-Grace's declarations and methods' arguments
+\paragraph{Grace's Typing}
+Critically, Grace's declarations and methods' arguments
 and results can be annotated with types, and those types can be  checked
@@
 ~type annotations should not affect the semantics of a correct
-program\citep{XXXSiek2015}, and the type system
-includes a distinguished ``{Unknown}'' type which matches any type
+program \citep{XXXSiek2015}, and the type system
+includes a distinguished ``{Unknown}'' type which matches any other type
 and is the implicit type for untyped parts of the program.
@@
 
-\subsection{Moth}
-
+\paragraph{Moth: A Grace Implementation}
 Moth is a virtual machine for Grace that we are developing as a
@@
 ``Shapes'' \citep{woss2014object} capture information about objects'
-structure and (runtime) 
+structure and (run-time) 
 field types, allowing a just-in-time compiler to
@@
 so checking an entire lattice for a benchmark of any size
-is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
+is infeasible.  Vitousek et al.\ modified the Takikawa protocol for
 these kinds of languages by taking an approach based on sampling
@@
 with 6 cores each, for a total of 24 hyperthreads.
-We used ReBench 0.10.1\citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
+We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
 Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).

### 2019-08-15 paper.tex
Minor changes
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 ``Shapes'' \citep{woss2014object} capture information about objects'
-structure and (run-time) 
+structure and (run time) 
 field types, allowing a just-in-time compiler to
@@
 which enables efficient native code generation for
-Truffle interpreters\citep{Wurthinger:2017:PPE}.
+Truffle interpreters \citep{Wurthinger:2017:PPE}.
 \RR{This last sentence seems out of place, does it belong here?}

### 2019-08-16 graphs/full-scatter.py
Added Scatter Graphs

@@ -1,0 +1,30 @@
+#!/usr/bin/env python3
+from results import *
+
+res = []
+i = 0
+for b, r in sorted(RESULTS.items(), key=lambda kv: kv[0].name):
+	points = [((k.types/b.types)*100, v.times.m) for k, v in r.partial.items()]
+	ymax = max([y for _, y in points])*1.1
+	yrelmax = ymax/r.partial[variant("0-0")].times.m
+	
+	# This is a horrible way to manually get the left y axis labels only on the leftmost graphs,
+	# and so on for the x axis, and right y axis.
+	left = "Full Scatter Left"
+	if i == 18:
+		left += " Corner"
+	elif i == 19:
+		left += " Bottom"
+	elif i % 3 == 0:
+		left += " Edge"
+		
+	right = "Full Scatter Right"
+	if i % 3 == 2:
+		right += " Edge"	
+	
+	res.append(double_axis(f"{left} = <{ymax}>", f"{b.name} <\\small({b.types} Type Annotations)>", f"{right} = <{yrelmax}>", 
+			  plot("", *points)))
+	
+	i += 1
+	
+graphs(3, *res)

### 2019-08-16 graphs/results.py

@@ -1,0 +1,186 @@
+#!/usr/bin/env python3
+import sys
+from collections import defaultdict
+from statistics import mean, stdev, median
+from scipy.stats import norm
+from math import sqrt
+import pickle
+
+# Some variables
+W = 200 # Number of warmup iterations to skip
+Z = norm.ppf((1+0.95)/2) # Z score, representing a 95% confidence interval
+
+PICKLE_FILE = "results.pickle"
+DATA_FILE = "results.data"
+TYPEMASK_FILE = "results.typemasks"
+
+# Some usefull functions
+def eprint(*args, **kwargs):
+	print(*args, file=sys.stderr, **kwargs)
+def log(a, b): # For debuging, prints a, and then returns the result of evaluating b
+	eprint(a)
+	return b()
+def printl(arg): # Converts < to { and > to }, so that format strings don't get horrible messy with tons of doubled {{ and }}
+	print(arg.replace("<", "{").replace(">", "}"))
+
+class defaultlist(list): # Stolen from https://stackoverflow.com/questions/8749542/creating-a-defaultlist-in-python
+    def __init__(self, fx):
+        self._fx = fx
+    def _fill(self, index):
+        while len(self) <= index:
+            self.append(self._fx())
+    def __setitem__(self, index, value):
+        self._fill(index)
+        list.__setitem__(self, index, value)
+    def __getitem__(self, index):
+        self._fill(index)
+        return list.__getitem__(self, index)
+
+# A class to handle interval arrithmetic, Interval(l, m, u) represents a closed interval between l and u, whith m as specific point inbetween
+# (the idea is that "m" represents an approximate value, and l and u represent lower/upper limits of the value)
+class Interval:
+	def __init__(self, l, m=None, u=None, data=None):		
+		if (m == None): 
+			m = l
+			u = l;
+		self.l = l; self.m = m; self.u = u; self.data = data
+
+	def convert(other):
+		return other if isinstance(other, Interval) else Interval(other);
+	def width(self):
+		#assert self.midpoint = self.m
+		return self.u - self.l
+	def midpoint(self):
+		return (self.u + self.l) / 2
+	def __str__(self):
+		return f"[{self.l}, {self.m}, {self.u}]"
+	
+	def __add__(self, other):
[…]

### 2019-08-16 graphs/preamble.tex
Added missing file

\usepackage{pgfplots}
\pgfplotsset{compat=1.16}
%\usepgfplotslibrary{external} 
%\tikzexternalize[prefix=pgfplots-cache/]

% Available ACM colours: ACMBlue, ACMYellow, ACMOrange, ACMRed, ACMLightBlue, ACMGreen, ACMPurple and ACMDarkBlue
\colorlet{GraphNormal}{ACMRed} % Normal colour when I don't care about having multiple colours

\pgfplotsset{%
% WARNING: do not put empty lines here, and make sure each = ... is terminated with a comma!
	every axis/.append style = {% options for all axes/graphs
		cycle list={}, tick align=center, xtick pos = bottom, enlarge x limits=false, ymin=0, axis on top,%
		every axis plot/.style={GraphNormal},%
		every tick label/.append style={font=\tiny},%
		every axis label/.append style={font=\tiny},%
	},%
	%
	Full Left/.style = {%
		ymax={#1}, axis y line*=left, height=3.2cm, width=5.9cm, try min ticks=5},%
	Right Axis/.style = {%
		axis y line*=right, axis x line=none, xmin=0, xmax=100},%
	%
	Scatter/.style = {%
		every axis plot/.append style={only marks}, mark=o,
		xticklabel={$\pgfmathprintnumber{\tick}$\%}},%
	%
	Full Scatter Left/.style = {Scatter, Full Left={#1},
		enlarge x limits={abs=2pt}, every axis plot/.append style={mark size={1pt}}},%
	%
	Full Scatter Right/.style       = {Full Scatter Left={#1},        Right Axis},%
	% Variants of the above that contain axis labels
	Full Scatter Left Edge/.style   = {Full Scatter Left={#1},        ylabel={Average Time (ms)}},%
	Full Scatter Left Bottom/.style = {Full Scatter Left={#1},        xlabel={Type Annotations}},%
	Full Scatter Left Corner/.style = {Full Scatter Left Bottom={#1}, Full Scatter Left Edge={#1}},%
	Full Scatter Right Edge/.style  = {Full Scatter Right={#1},       ylabel={Relative to 0\% Typed}},%
	%
	%
	Column/.style n args = {2}{%
		const plot, every axis plot/.append style={draw=GraphNormal, no markers, fill=GraphNormal!40}},%
	%
	Full Column Left/.style 2 args        = {Column, Full Left={#1},           xtick distance={#2}},%
	Full Column Right/.style              = {Full Column Left={#1}{1},         Right Axis},%
	% Variants of the above that contain axis labels
	Full Column Left Edge/.style 2 args   = {Full Column Left={#1}{#2},        ylabel={Average Time (ms)}},%
	Full Column Left Bottom/.style 2 args = {Full Column Left={#1}{#2},        xlabel={Type Annotation Index}},%
	Full Column Left Corner/.style 2 args = {Full Column Left Bottom={#1}{#2}, Full Column Left Edge={#1}{#2}},%
	Full Column Right Edge/.style         = {Full Column Right={#1},           ylabel={Relative to Untyped}},%
}

\newcommand\DOUBLEAXIS[3]{%
	\begin{axis}[#1]#3\end{axis}%
	\begin{axis}[#2]\end{axis}}

\newcommand\GRAPHS[1]{\null\hfill\begin{tikzpicture}[ampersand replacement=\&]\matrix{#1\\};\end{tikzpicture}\null\hfill}

### 2019-08-16 paper.tex
Added Scatter Graphs
Started a related work section. First entry: Optimization Coaching
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
+\input{graphs/preamble}
 \setcopyright{none}
@@
 
+\begin{figure*}
+	\input{graphs/full-scatter}
+	\caption{The graphs!}
+\end{figure*}
+
 \section{Individual Annotations}
@@
   - show that single annotation prediction doesn't really work\ldots
+
+\section{Related Work}
+
+Optimization Coaching
+\citep{St-Amour:2012:OCO} is a related approach more
+directly integrated into the language runtime and development environment.
+It uses feedback from the runtime to guide developers to possible
+code changes that enable a compiler to generate a more optimal program.
+In spirit, we would eventually want to achieve the same.
+Though in our case,
+we need to run full experiments to get the necessary information.
+Furthermore, we still lack the seamless integration in a development environment.
 

### 2019-08-16 paper.tex
Some references for work on performance visualizations for parallel programming
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
+The high-performance computing community also has been investigating how
+tools and visualization can help developers to utilize their systems
+more efficiently \citep{Papenhausen:2016:IVT,daSilva:2019:PSV}.
+Their focus is typically on parallelisation opportunities,
+guided by runtime feedback, cost models, or heuristics. 
+The large body of work\citep{Isaacs:2014:PerfViz} uses various approaches,
+though, we are not aware of work has used an approach as ours.
+One may argue, this is largely a question of practicality.\sm{
+i put this in to hint at issues with usability.}
+
 \section{Discussion and Conclusion}
@@
 
+\sm{is our approach practical. Is it useful for day-to-day development?
+What's its context, use, general application?}
+
 \bibliography{references}

### 2019-08-17 graphs/preamble.tex
Reworded half the paper

@@
 
+\colorlet{GraphBase}{gray} % The colours for no type annotation
+\colorlet{GraphA}{ACMRed}
+\colorlet{GraphB}{ACMBlue}
+\colorlet{GraphAB}{ACMPurple}
+
 \pgfplotsset{%
@@
 	every axis/.append style = {% options for all axes/graphs
-		cycle list={}, tick align=center, xtick pos = bottom, enlarge x limits=false, ymin=0, axis on top,%
-		every axis plot/.style={GraphNormal},%
+		cycle list={}, tick align=center, xtick pos = bottom, ymin=0, axis on top,%
+		axis y line*=left, try min ticks=5,%
 		every tick label/.append style={font=\tiny},%
@@
 	%
-	Full Left/.style = {%
-		ymax={#1}, axis y line*=left, height=3.2cm, width=5.9cm, try min ticks=5},%
-	Right Axis/.style = {%
+	Full/.style = { % For the page of graphs for each benchmark
+		enlarge x limits={abs=2pt}, ymax={#1}, height=3.2cm, width=5.9cm,%
+		every axis plot/.style={mark size=1pt, draw=GraphNormal, fill=GraphNormal!40}},%
+	Right Axis/.style = { % For right axes
 		axis y line*=right, axis x line=none, xmin=0, xmax=100},% 
 	%
-	Scatter/.style = {%
-		every axis plot/.append style={only marks}, mark=o,
-		xticklabel={$\pgfmathprintnumber{\tick}$\%}},%
+	Scatter/.style = { % For the scatter plots
+		xticklabel={$\pgfmathprintnumber{\tick}$\%},%
+		every axis plot/.append style={only marks}, mark=o},%
 	%
-	Full Scatter Left/.style = {Scatter, Full Left={#1},
-		enlarge x limits={abs=2pt}, every axis plot/.append style={mark size={1pt}}},%
-	%
-	Full Scatter Right/.style       = {Full Scatter Left={#1},        Right Axis},%
 	% Variants of the above that contain axis labels
-	Full Scatter Left Edge/.style   = {Full Scatter Left={#1},        ylabel={Average Time (ms)}},%
-	Full Scatter Left Bottom/.style = {Full Scatter Left={#1},        xlabel={Type Annotations}},%
-	Full Scatter Left Corner/.style = {Full Scatter Left Bottom={#1}, Full Scatter Left Edge={#1}},%
-	Full Scatter Right Edge/.style  = {Full Scatter Right={#1},       ylabel={Relative to 0\% Typed}},%
+	Scatter Left Label/.style   = {ylabel={Average Time (ms)}},%
+	Scatter Bottom Label/.style = {xlabel={Type Annotations}},%
+	Scatter Right Label/.style  = {ylabel={Relative to 0\% Typed}},%
 	%
 	%
-	Column/.style n args = {2}{%
-		const plot, every axis plot/.append style={draw=GraphNormal, no markers, fill=GraphNormal!40}},%
+	Column/.style = {% For the column graphs
+		enlarge x limits=false, xtick distance={#1}, const plot,%
+		every axis plot/.append style={no markers}},%
 	%
-	Full Column Left/.style 2 args        = {Column, Full Left={#1},           xtick distance={#2}},%
-	Full Column Right/.style              = {Full Column Left={#1}{1},         Right Axis},%
 	% Variants of the above that contain axis labels
-	Full Column Left Edge/.style 2 args   = {Full Column Left={#1}{#2},        ylabel={Average Time (ms)}},%
-	Full Column Left Bottom/.style 2 args = {Full Column Left={#1}{#2},        xlabel={Type Annotation Index}},%
-	Full Column Left Corner/.style 2 args = {Full Column Left Bottom={#1}{#2}, Full Column Left Edge={#1}{#2}},%
-	Full Column Right Edge/.style         = {Full Column Right={#1},           ylabel={Relative to Untyped}},%
+	Column Left Label/.style   = {ylabel={Average Time (ms)}},%
+	Column Bottom Label/.style = {xlabel={Type Annotation Index}},%
+	Column Right Label/.style  = {ylabel={Relative to Untyped}},%
+	%
+	Pattern/.style = { % For the page of graphs for each benchmark
+		enlarge x limits={abs=1.5pt}, ymax={#1}, height=3.2cm, width=4.4cm, %
+		every axis plot/.style={mark size=0.75pt, draw=GraphBase, fill=GraphBase!40}},%
+	%
+	Pattern A/.style = {draw=GraphA, fill=GraphA!40, mark=*}, %
+	Pattern B/.style = {draw=GraphB, fill=GraphB!40, mark=*}, %
+	Pattern AB/.style = {draw=GraphAB, fill=GraphAB!40, mark=*}, %
+	%Pattern Left/.style = {Left Axis={#1}, height=3.2cm, width=7.9cm}, % Size for pattern graphs
+	%Full Scatter Left/.style = {Scatter, Full Left={#1},
+	%	enlarge x limits={abs=2pt}, every axis plot/.append style={mark size={1pt}}},%
+	%Full Scatter Right/.style       = {Full Scatter Left={#1},        Right Axis},%
+	%sleft = "Pattern Scatter Left"
+	%cleft = "Pattern Column Left"
+	%sright = "Pattern Scatter Right"
+	%cright = "Pattern Column Right"
+	%Pattern Scatter A, Pattern Scatter B
 }

### 2019-08-17 paper.tex
Reworded half the paper

@@
 \begin{abstract}
-Just-in-time compilation and optimisation in virtual machines can
-eliminate much of the overhead of transient typechecks.  Unfortunately
-the improvement is not uniform: while most typechecks can be optimised
-away, some typechecks will significantly decrease a program's
-performance.  In this paper we investigate how benchmark results
-indicate which typechecks cause the most problems.
+One form of type checking used in gradually typed language is \emph{transient type checking}: when ever an object 'flows' through code with a type annotation, the object is dynamically checked to have the methods required by said annotation.
+
+Though na\"ive implementations of this have a high runtime overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.  Unfortunately
+the improvement is not uniform: while most type checks can be optimised
+away, some will significantly decrease a program's
+performance, and some may even increase it.  In this paper we investigate how benchmark results can be used to identify which type checks have the greatest effects on performance.
 Using these techniques,
-programmers can optimise programs by removing typechecks,
-and VM engineers can identify opportunities for compiler optimisation.
+programmers can optimise programs by removing expensive type checks,
+and VM engineers can identify new opportunities for compiler optimisation.
 
@@
 %% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-%%KJX XML IS BROKEN!!!
 \begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10011006.10011041.10011044</concept_id>
+<concept_desc>Software and its engineering~Just-in-time compilers</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011008.10011009.10011011</concept_id>
+<concept_desc>Software and its engineering~Object oriented languages</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011041.10010943</concept_id>
+<concept_desc>Software and its engineering~Interpreters</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
 \end{CCSXML}
@@
 to permit dynamic type annotations within static languages, increasing
-flexibility while maintaining safety
+flexibility whilst maintaining safety
 \citep{AbadiTOPLAS1991}.
@@
 \cite{kafka18,bensurvey18icfp}.
-At one end\,---\,``pluggable types'' as in Strongtalk \cite{strongtalk} or ``erasure
-semantics'' as in 
+At one end is ``pluggable types'' (as in Strongtalk \cite{strongtalk}) or ``erasure
+semantics'' (as in 
 %% Two important approaches are optional\citep{}
@@
 TypeScript
-\citep{typeScriptECOOP}\,---\,
+\citep{typeScriptECOOP}), here
 %
-all types are erased before the execution, limiting the benefit of
-types to the statically typed parts of programs, and preventing
-programs from depending on type checks at run time.  In the middle,
-``transient'' or ``type-tag'' checks as in Reticulated Python 
+all types are erased before execution, limiting the benefit of
+types to statically typed portions of programs, whilst not providing 
+any for of run time checking.  In the middle there are
+``transient'' or ``type-tag'' checks (as in Reticulated Python)
 % and ``concrete'' checks as in Thorn
-offer first-order semantics, checking
-whether an object's type constructor or supported methods match
-explicit type declarations
+, these offer first-order semantics, by checking
+that an object's type constructor or supported methods match
+any explicit type annotations that the object flows through.
 %%%WRONG kjx
@@
 \cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
-Reticulated Python also supports an alternative ``monotonic'' semantics
-which mutates an object to narrow its concrete type when it is passed
-into a more specific type context.
-At the other end of the spectrum, behavioral
-typechecks as in Typed Racket \cite{typedScheme08,takikawa2012},
-Gradualtalk \cite{gradualtalk14},
-and Reticulated Python's proxies,
+Reticulated Python also supports alternative ``monotonic'' semantics
+which mutate objects by narrowing their concrete types when passed
+into contexts with more specific types.
+On the other end of the spectrum are behavioural
+typechecks (as in Typed Racket \cite{typedScheme08,takikawa2012},
+Gradualtalk \cite{gradualtalk14}),
+and ``proxies'' (provided by Reticulated Python), these
 support higher-order semantics, retaining
 types until run 
-time, performing the checks eagerly, and giving detailed information
-about type violations as soon as possible via blame
+time, and performing checks eagerly, thus detailed information
+about type violations can be given as soon as possible via ``blame''
 tracking \cite{blame2009,blameForAll2011}.
-Finally, Ductile typing
-dynamically interprets a static type system at run time
+Finally, there is also ductile typing,
+which dynamically interprets a static type system at run time
 \cite{Ductile2011}.
@@
 %
-Unfortunately, any gradual system with run-time semantics
-(i.e.\ everything more complex than erasure) currently
-imposes a significant run-time performance overhead to provide those
+% This said "any", but the ecoop paper specifically showed that 
+Unfortunately, most gradual system with run-time semantics
+(as opposed to type erasure as in TypeScript) currently
+imposes a significant run-time performance overhead to provide such
 semantics.  This has lead to a significant body of research,
@@
 \citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
-and on the other hand developing techniques to evaluate the
+and on the other, developing techniques to evaluate the
 resulting performance  \cite{Takikawa2016,Greenman2019jfp}.
@@
 This paper builds upon our recent work on optimising transient
-typechecks \cite{Roberts2017,roberts-and-co-ecoop-2019} to
-investigate how benchmark results can be repurposed to identify which
-particular transient typechecks are resistant to optimisation.
-The next section discusses dynamic typechecks and gradual typing in
-Grace, then section~\ref{s-eval} describes the benchmarking
-protocols. Section~\ref{s-overall} then presents the results of
-analysing overall benchmark results, while section~\ref{s-individual}
+type checks \cite{Roberts2017,roberts-and-co-ecoop-2019} by
+investigating how benchmark results can be repurposed to identify which
+particular transient type checks are resistant to optimisation.
+The next section discusses dynamic type checks and gradual typing in
+Moth (an implementation of the Grace language), then section~\ref{s-eval} describes our benchmarking protocol. Section~\ref{s-overall} then presents the overall results of
+our benchmarks, while section~\ref{s-individual}
 looks at the results of benchmarking individual type checks.
-Finally~\ref{s-concl} discusses our results, briefly considers threats
-to validity, and concludes.
+Section~\ref{s-rel} presents some additional related work, and finally \ref{s-concl} summarises our results, and briefly considers threats to validity.
 
@@
 of the Grace programming language 
-\citep{graceOnward12,graceSigcse13},
-based on the Graal and Truffle toolchain
-\cite{Wurthinger:2017:PPE,Wurthinger2013}
+\citep{graceOnward12,graceSigcse13}.
+Moth is based on the Graal and Truffle toolchain
+\cite{Wurthinger:2017:PPE,Wurthinger2013},
 and developed from a Newspeak implementation based on the  Simple
@@
 
-\subsection{Grace's Transient Type Checking}
+\subsection{Grace and Transient Type Checking}
 
@@
 While Grace's syntax draws
-from the so-called ``curly bracket'' tradition of C, Java, and
+from the so-called ``curly bracket'' traditions of C, Java, and
 JavaScript, the structure of the language
 is in many ways closer to Smalltalk:
-all computation is via dynamically dispatched  ``method requests''
-where the object receiving the request decides which code to run,
-and
-%
-returns within lambdas are ``non-local'', returning to the method
-activation in which the block is instantiated \citep{bluebook}.  In
+all computation is done via dynamically dispatched  ``method requests''
+where the object receiving the request decides which they return to the point where execution first encountered the lambda \citep{bluebook}.  In
 other ways, Grace is closer to JavaScript than Smalltalk: Grace
@@
 \paragraph{Grace's Typing}
-Critically, Grace's declarations and methods' arguments
-and results can be annotated with types, and those types can be  checked
-either statically or dynamically. This means the type system is
-intrinsically gradual:%
+Critically, in Grace, declarations, methods can be annotated with types.
+As Grace is designed to support a variety of teaching methods, implementation of Grace are free to check such type annotations  statically, dynamically, or not at all.
+ Thus the type system of Grace is intrinsically gradual:%
 %
@@
 program \citep{XXXSiek2015}, and the type system
-includes a distinguished ``{Unknown}'' type which matches any other type
-and is the implicit type for untyped parts of the program.
+includes a distinguished ``{Unknown}'' type which matches any other type,
+and is the default type when type annotations are omitted.
 
-The static core of Grace's type system is well-described
+Static typing for the core of Grace's type system has been described
 elsewhere \citep{TimJonesThesis};
@@
 Grace's types are structural \citep{graceOnward12},
-that is, an object implements a type whenever it
-implements all the methods required by that type,
-rather than requiring classes or objects to declare types explicitly \RR{(explicitly or nominally)?}. 
-% This should say Moth's, not Grace's, but I know there is objection to that
-Grace's transient dynamic typechecks are first-order, so methods match when they have
-the same name and arity, while
-argument and return types are ignored.
-A type thus expresses the requests an object can respond to,
-for example whether a particular accessor is available,
-rather than a nominal location in an explicit inheritance hierarchy.
+that is, an object conforms to a type whenever it conforms to the ''structural`` requirements of a type,
+rather than requiring classes or objects to explicitly declare their intended type.
 
-% Ditto
-Grace then checks the types of values at run time:
-%
+In Grace, types specify a set of method signatures that an object must provide.
+Thus a type expresses the requests an object can respond to, for example whether a particular accessor is available, rather than a location in a class hierarchy.
+
+Moth's implementation of transient type checks however are only only first-order, as such it only dynamically checks that an object has methods of the same name and arity as are required by a type, however any argument and return types of such methods are not checked.
+
+In particular, Moth performs such type checks are run time:
 \begin{itemize}
-\item the values of arguments are checked after a method is requested, 
-      but before the body of the method is executed;
-\item the value returned by a method is checked after its body is executed; and
-\item the values of variables are checked
-      whenever written or read by user code.
+\item when a method is requested, arguments that are passed are checked against the corresponding parameter type annotations of the called method, this is done before the body of the method is executed;
+\item when the body of a method has finished executing, but before it returns to its caller, the method's return value is checked against the return type annotation of the called method;
+\item whenever a variable is read or written to, its value is checked against the type specified by the variables declaration.
 \end{itemize}
 
-To see how this works in practice, consider this code snippet, which
-will run in Grace.
+To see how this works in practice, consider this piece of Grace code:
 
@@
 
-Dynamic typechecks will take place:
+Moth will perform dynamic type checks:
 
@@
 \item on line 7,
-when object \code{o} initialises the variable \code{t},
-we check it has the method \code{three};
+when the \code{o} object initialises the variable \code{t},
+Moth checks that \code{o} has a 0-argument method called \code{three};
 
@@
 when the value of \code{t} is read,
-it must again have the method \code{three};
+Moth checks that it's value (\code{o}) still has a \code{three} method;
 
@@
 when the \code{three} method requested by ``\code{t.three}'' returns,
-the return value must be a \code{Number};
+Moth checks that returned value conforms to the \code{Number} type;
 and (presumably) within the definition of
@@
 %
-(not shown) to check that its argument is a number. 
+(not shown), Moth will again check that the value is a \code{Number}.
 \end{itemize}
@@
 Note that we never check
-whether e.g.\ the result of requesting ``\code{t.three}'' is actually
-a \code{String} (as one may expect from line 5) because we check
-only first-order types (whether objects have conforming methods) not higher-order
-types (whether those methods' argument and result types conform), and
-we check only when values flow through explicit type annotations.
+whether the result of requesting ``\code{t.three}'' is actually
+a \code{String} (as one may expect from line 5) because Moth only performs first-order type checks
+(it checks whether objects have conforming methods) not higher-order
+checks (whether the argument and result types of methods' conform). In addition, Moth
+only checks when values flow through explicit type annotations.
 This is why the type declared in lines 4-6 is checked only on line 7
@@
 \section{Related Work}
+\label{s-rel}
 

### 2019-08-17 graphs/full-scatter.py

@@ -10,20 +10,17 @@ for b, r in sorted(RESULTS.items(), key=lambda kv: kv[0].name):
 	
 	# This is a horrible way to manually get the left y axis labels only on the leftmost graphs,
 	# and so on for the x axis, and right y axis.
-	left = "Full Scatter Left"
-	if i == 18:
-		left += " Corner"
-	elif i == 19:
-		left += " Bottom"
-	elif i % 3 == 0:
-		left += " Edge"
+	left = f"Full = <{ymax}>, Scatter"
+	if i % 3 == 0:
+		left += ", Scatter Left Label"
+	if i >= 18:
+		left += ", Scatter Bottom Label"
 		
-	right = "Full Scatter Right"
+	right = f"Full = <{yrelmax}>, Scatter, Right Axis"
 	if i % 3 == 2:
-		right += " Edge"	
+		right += ", Scatter Right Label"
 	
-	res.append(double_axis(f"{left} = <{ymax}>", f"{b.name} <\\small({b.types} Type Annotations)>", f"{right} = <{yrelmax}>", 
-			  plot("", *points)))
+	res.append(double_axis(left, f"{b.name} <\\small({b.types} Type Annotations)>", right,  plot("", *points)))
 	
 	i += 1
 	

### 2019-08-17 graphs/results.py

@@ -142,6 +142,8 @@ def variant(s):
 	vn = s.split("-")
 	return S(index = int(vn[0]), types = int(vn[1]))
 
+def floor_round(num, base):
+	return max(5, int(num//base)*base)
 
 # Have to use functions (and not lambdas) so they can be pickled
 def empty_result():

### 2019-08-17 paper.tex
Reworded up to section 4

@@
 
+\paragraph{Moth's Transient Type Checking}
 Moth's implementation of transient type checks however are only only first-order, as such it only dynamically checks that an object has methods of the same name and arity as are required by a type, however any argument and return types of such methods are not checked.
@@
 
-\paragraph{Moth: A Grace Implementation}
-Moth is a virtual machine for Grace that we are developing as a
+\paragraph{Moth's Optimisation}
+We are developing Moth as a
 research platform \cite{roberts-and-co-ecoop-2019}. Like other VMs
@@
 AST interpreter \cite{Wurthinger:2012:SelfOptAST}. 
-The key idea is that an AST rewrites itself based on a program's run-time values
+The key idea is that an AST rewrites itself based on a program's run time values
 to reflect the minimal set of operations needed to execute the program
@@
 on an object invokes a second method on the exact same object) will
-always result in executing the second method defined in same concrete
-class as the first method. The second method can then be inlined into
-the first method, avoiding the overheads of a machine-level subroutine
-invocation and of an object-oriented dynamic dispatch.
+always result in executing the exact same method. Thus the called method can be inlined into
+the callee, avoiding overhead of a machine-level subroutine
+invocation and an object-oriented dynamic dispatch.
 
@@
 ``Polymorphic inline caches''
-\citep{Hoelzle:91:PIC} uses object shapes to cache the results of
+\citep{Hoelzle:91:PIC} use object shapes to cache the results of
 method lookups, avoiding expensive class hierarchy searches or
 indirect jumps through virtual method tables. 
-For languages built on the Truffle framework,
+Since Moth is built on the Truffle framework,
 Graal comes with  additional support for partial evaluation,
@@
 In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the
-performance of transient typechecks using the \AWFY
-benchmark suite \cite{Marr2016}, plus other
-benchmarks from the gradual-typing literature.
-The goal was to complement our benchmarks with additional ones that are
-used for similar experiments and can be ported to Grace.
+performance of transient type checks using 21 benchmarks.
+Including the 14 benchmarks of the \AWFY
+benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.
+We also used benchmarks taken from the gradual-typing literature.
+Our goal was to complement our benchmarks with additional ones that are used for similar experiments and can be ported to Grace.
 To this end, we surveyed a number of papers
 \citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
-and selected benchmarks that have been used by multiple papers.
-Some of these benchmarks overlapped with the \AWFY suite,
+and selected 9, benchmarks that have been used by multiple papers.
+Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
 or were available in different versions.
-While not always behaviorally equivalent,
+While not always behaviourally equivalent,
 we chose the \AWFY versions since we already used them to
@@
 The benchmarks were modified to have complete type information.  To
-ensure correctness and completeness of these experiments, we added an
-additional check to Moth that reports absent type information to
-ensure each benchmark is completely typed.  To assess the performance
+ensure correctness and completeness of these experiments, we ensured that each benchmark is completely typed.  To assess the performance
 overhead of type checking, we compared the execution of Moth with all
 checks disabled against an execution that has all checks enabled.  The
-results of this evaluation demonstrated that Moth runs programs with
-transient typechecks enabled at an overhead of of 5\% (min.\ -13\%,
-max. 79\%) over peak performance without typechecks.
+results of this evaluation demonstrated that Moth's transient type checking has an overhead of 5\% (min.\ -13\%,
+max. 79\%) compared to Moth's peak performance with type checking disabled.
 
 We did not measure programs that mixed typed and untyped code because
-with our implementation technique a fully typed program is expected to
+with our implementation technique we expected fully typed programs to
 have the largest overhead. This is not the case for other kinds of
@@
 ``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
-benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
+benchmarks a \emph{lattice} of $2^N$ \emph{configurations} of each
 benchmark, where a configuration is a particular mix of static and
@@
 Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
-allows programmers to choose typechecks at the level of individual
-declarations: $N$ is the number of type annotations in the program,
-so checking an entire lattice for a benchmark of any size
-is infeasible.  Vitousek et al.\ modified the Takikawa protocol for
+allows programmers to choose whether each individual declaration should be type-checked: so here $N$ is the number of type annotations in the program,
+so checking an entire lattice for even a moderately sized benchmark would be infeasible.  Vitousek et al.\ modified the Takikawa protocol for
 these kinds of languages by taking an approach based on sampling
@@
 
-To investigate Moth's support for transient typechecks further, we
-carried out a (small?) Takikawa-Vitousek evaluation using (some of?)
-our existing benchmarks.
+To investigate Moth's support for transient type checks further, we carried out a similar approach to Takikawa-Vitousek evaluation: for each benchmark we generated 100 partially typed versions (or less if the benchmark was too small).
+We did an even split so that for each $i \ge 1$ and $i$ < $N$ (where $N$ is the total number of type annotation), we generated roughly the same number of configurations with $i$ type annotations.
+We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to randomly choose the type annotations each configuration contained, and we ensured that no duplicate configurations where generated.
+In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark, (or 97 in the case of Storage, since it only has 10 type annotations).
 
-\textbf{!!!KJX: Need to update this paragraph with what we did this time around!!!}
-To account for the complex warmup behavior
-of modern systems\citep{Barrett:2017:VMW} as well as
-the non-determinism caused by \eg garbage collection and cache effects,
-we run each benchmark for \textbf{$/$NumIterationsAll} iterations in the same
-VM invocation.\footnote{
-For the Higgs VM, we only use \textbf{$/$NumIterationsHiggs} iterations,
-because of its lower performance.
-This is sufficient since Higgs's compilation approach induces less variation
-and leads to more stable measurements.}
-Afterwards, we inspected the run-time plots over the iterations
-and manually determined a cutoff of \textbf{$/$WarmupCutOff} iterations for warmup,
-\ie, we discard iterations with signs of compilation.
-As a result, we use a large number of data points to compute the average,
-but outliers, caused by \eg garbage collection, remain visible in the plots.
-% In this work, we do not consider startup performance,
-% because we want to assess the impact of gradual type checks
-% on the best possible performance.
-All reported averages use the geometric mean since they aggregate
-ratios.
+As we did in our previous work \cite{roberts-and-co-ecoop-2019} which used the same set of benchmarks, to account for the complex warmup behaviour of modern systems\citep{Barrett:2017:VMW} as well as
+the non-determinism caused by \eg garbage collection and cache effects, we ran each benchmark for 1000 iterations in the same invocation of Moth, and used cut off of 350 iterations for warmup, \ie we discard the first 350 iterations, since they showed signs of compilation.
 
-\textbf{
-% Yuria
-%  - Ubuntu 16.04.4, Kernel 3.13
-%  - 24 hyperthreads
-%  - Intel Xeon E5-2620 v3 2.40GHz
-% Graal 0.33 Feb. 2018
-%
-% TODO: update these details when we have time... not a rejection criterion,
-% but needs to be correct in the final version
-All experiments were executed on a machine running Ubuntu Linux 16.04.4,
-with Kernel 3.13.
-The machine has two Intel Xeon E5-2620 v3 2.40GHz,
-with 6 cores each, for a total of 24 hyperthreads.
-We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
-Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).
+Though though outliers, caused by \eg garbage collection, remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained for the arithmetic mean of any of the results for any of our experiments was $\pm3.53\%$ (for the PyStone benchmark).
+
+All experiments were executed on a Windows 10 machine
+running Fedora 29 within Windows Subsystems for Linux (version 1). The machine has
+
+one Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
 Benchmarks were executed one by one to avoid interference between them.
-The analysis of the results was done with R 3.4.1,
-and plots are generated with ggplot 2.2.1 and tikzDevice 0.11.
-Our experimental setup is available online to enable reproductions.\footnote{
-\url{https://github.com/gracelang/moth-benchmarks}}
-}
+The analysis of the results and plots where generated using Python 3.7.3 and PGFPLOTS 1.16.
+To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, is available online.\footnote{
+\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}
 
@@
 The large body of work\citep{Isaacs:2014:PerfViz} uses various approaches,
-though, we are not aware of work has used an approach as ours.
+though, we are not aware of work that has used an approach similar to ours.
 One may argue, this is largely a question of practicality.\sm{

### 2019-08-17 graphs/results.py

@@ -7,7 +7,7 @@ from math import sqrt
 import pickle
 
 # Some variables
-W = 200 # Number of warmup iterations to skip
+W = 350 # Number of warmup iterations to skip
 Z = norm.ppf((1+0.95)/2) # Z score, representing a 95% confidence interval
 
 PICKLE_FILE = "results.pickle"

### 2019-08-17 paper.tex
James edits to abstract, sec1

@@
 \begin{abstract}
-One form of type checking used in gradually typed language is \emph{transient type checking}: when ever an object 'flows' through code with a type annotation, the object is dynamically checked to have the methods required by said annotation.
+One form of type checking used in gradually typed language is
+\emph{transient type checking}: when ever an object `flows' through
+code with a type annotation, the object is dynamically checked to
+ensure it has the methods required by the annotation.
 
-Though na\"ive implementations of this have a high runtime overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.  Unfortunately
+Although na\"ive implementations of transient type checks have a high runtime overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.
+Unfortunately
 the improvement is not uniform: while most type checks can be optimised
 away, some will significantly decrease a program's
-performance, and some may even increase it.  In this paper we investigate how benchmark results can be used to identify which type checks have the greatest effects on performance.
+performance, and some may even increase it.  In this paper we investigate techniques that interpret benchmark results to identify which type checks have the greatest effects on performance.
 Using these techniques,
@@
 otherwise, the sentence in using these techniques doesn't make sense to me, because we didn't say anything about them.}
+\kjx{still like brief abstracts. at least I defined ``techniques''}
 \end{abstract}
@@
 TypeScript
-\citep{typeScriptECOOP}), here
+\citep{typeScriptECOOP}), where
 %
 all types are erased before execution, limiting the benefit of
-types to statically typed portions of programs, whilst not providing 
+types to statically typed portions of programs, while not providing 
 any for of run time checking.  In the middle there are
-``transient'' or ``type-tag'' checks (as in Reticulated Python)
+``transient'' or ``type-tag'' checks (as in Reticulated Python),
 % and ``concrete'' checks as in Thorn
-, these offer first-order semantics, by checking
+which offer first-order semantics, by checking
 that an object's type constructor or supported methods match
-any explicit type annotations that the object flows through.
+% any explicit type annotations
+any static types
+that the object flows through.
 %%%WRONG kjx
@@
 Gradualtalk \cite{gradualtalk14}),
-and ``proxies'' (provided by Reticulated Python), these
+and ``proxies'' (provided by Reticulated Python), which
 support higher-order semantics, retaining
@@
 % This said "any", but the ecoop paper specifically showed that 
-Unfortunately, most gradual system with run-time semantics
-(as opposed to type erasure as in TypeScript) currently
-imposes a significant run-time performance overhead to provide such
+Unfortunately, most gradual systems with run-time semantics
+(as opposed to type erasure as in TypeScript) typically
+imposes significant run-time performance overheads to provide these
 semantics.  This has lead to a significant body of research,

### 2019-08-17 paper.tex
James edits to end sec 3

@@
 all computation is done via dynamically dispatched  ``method requests''
-where the object receiving the request decides which they return to the point where execution first encountered the lambda \citep{bluebook}.  In
+where the object receiving the request decides which
+%kjx reverted cos edited version made no sense.
+%for this paper, we could just delete the comments about blocks if you like.
+method to run, control structures are built out of blocks (essentialy lambda expressions), and return statements are ``non-local'', returning from the lexically enclosing method \citep{bluebook}.
+%
+In
 other ways, Grace is closer to JavaScript than Smalltalk: Grace
@@
 \paragraph{Grace's Typing}
-Critically, in Grace, declarations, methods can be annotated with types.
+In Grace, all declarations can be annotated with types.
 As Grace is designed to support a variety of teaching methods, implementation of Grace are free to check such type annotations  statically, dynamically, or not at all.
- Thus the type system of Grace is intrinsically gradual:%
+The type system of Grace is intrinsically gradual:%
 %
 ~type annotations should not affect the semantics of a correct
-program \citep{XXXSiek2015}, and the type system
-includes a distinguished ``{Unknown}'' type which matches any other type,
-and is the default type when type annotations are omitted.
+program \citep{XXXSiek2015}. The type system
+includes a distinguished ``{Unknown}'' type which matches any other type; this unknown type is the default when type annotations are omitted.
 
@@
 In Grace, types specify a set of method signatures that an object must provide.
-Thus a type expresses the requests an object can respond to, for example whether a particular accessor is available, rather than a location in a class hierarchy.
+A type expresses the requests an object can respond to, for example whether a particular accessor is available, rather than a location in a class hierarchy.
 
 \paragraph{Moth's Transient Type Checking}
-Moth's implementation of transient type checks however are only only first-order, as such it only dynamically checks that an object has methods of the same name and arity as are required by a type, however any argument and return types of such methods are not checked.
+Moth's implementation of transient type checks are only only first-order.
+Moth only checks dynamically that an object has methods of the same name and arity as are required by a type:  any argument and return types of such methods are not checked.
 
-In particular, Moth performs such type checks are run time:
+In particular, Moth performs the following type checks at run time:
 \begin{itemize}
@@
 when the value of \code{t} is read,
-Moth checks that it's value (\code{o}) still has a \code{three} method;
+Moth checks that its value (\code{o}) still has a \code{three} method;
 
 \item on line 2,
-when the \code{three} method requested by ``\code{t.three}'' returns,
+when the method requested by ``\code{t.three}'' returns,
 Moth checks that returned value conforms to the \code{Number} type;
@@
 objects involved. In the simplest case, a ``self'' call (when one method
-on an object invokes a second method on the exact same object) will
+on an object requests a second method on the exact same object) will
 always result in executing the exact same method. Thus the called method can be inlined into
@@
 ``Shapes'' \citep{woss2014object} capture information about objects'
-structure and (run time) 
+structures and (run time) 
 field types, allowing a just-in-time compiler to
@@
 \RR{This last sentence seems out of place, does it belong here?}
+\kjx{yes I think it's fine}
 
@@
 performance of transient type checks using 21 benchmarks.
-Including the 14 benchmarks of the \AWFY
+We started with the 14 benchmarks of the \AWFY
 benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.
-We also used benchmarks taken from the gradual-typing literature.
-Our goal was to complement our benchmarks with additional ones that are used for similar experiments and can be ported to Grace.
-To this end, we surveyed a number of papers
+We added another nine benchmarks taken from the gradual-typing literature.
+Our goal was to complement the \AWFY with additional ones that are used for similar experiments and can be ported to Grace.
+We surveyed a number of papers
 \citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
-and selected 9, benchmarks that have been used by multiple papers.
+and selected additional benchmarks that have been used by multiple papers.
 Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
@@
 we chose the \AWFY versions since we already used them to
-establish the performance baseline.
+establish a performance baseline.
 The selected benchmarks as well as the papers in which they were used are shown in
@@
 evaluation techniques.  In particular, the so-called ``Takikawa'' or
-``Takikawa-Greenman'' evaluation \cite{Takikawa2016,Greenman2019jfp}
+``Takikawa-Greenman'' evaluation protocol \cite{Takikawa2016,Greenman2019jfp}
 benchmarks a \emph{lattice} of $2^N$ \emph{configurations} of each
@@
 Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
-allows programmers to choose whether each individual declaration should be type-checked: so here $N$ is the number of type annotations in the program,
-so checking an entire lattice for even a moderately sized benchmark would be infeasible.  Vitousek et al.\ modified the Takikawa protocol for
+allows programmers to choose whether each individual declaration should be type-checked. This means $N$ in Grace is the number of type annotations in the program, and so
+checking an entire lattice for even a moderately sized benchmark would be infeasible.  Vitousek et al.\ modified the Takikawa protocol for
 these kinds of languages by taking an approach based on sampling
 \cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
-divides the number of type annotations in an fully-typed program into
+divides the number of type annotations in a fully-typed program into
 a maximum of 100 intervals, and then randomly generates ten programs within
@@
 
-To investigate Moth's support for transient type checks further, we carried out a similar approach to Takikawa-Vitousek evaluation: for each benchmark we generated 100 partially typed versions (or less if the benchmark was too small).
-We did an even split so that for each $i \ge 1$ and $i$ < $N$ (where $N$ is the total number of type annotation), we generated roughly the same number of configurations with $i$ type annotations.
-We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to randomly choose the type annotations each configuration contained, and we ensured that no duplicate configurations where generated.
+To investigate Moth's support for transient type checks further, we carried out a similar approach to the Takikawa-Vitousek protocol.
+For each benchmark we generated 100 partially typed versions (fewer if the benchmark was too small).
+We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations.
+We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to choose the type annotations each configuration contained randomly, and we ensured that no duplicate configurations were generated.
 In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark, (or 97 in the case of Storage, since it only has 10 type annotations).
 
-As we did in our previous work \cite{roberts-and-co-ecoop-2019} which used the same set of benchmarks, to account for the complex warmup behaviour of modern systems\citep{Barrett:2017:VMW} as well as
-the non-determinism caused by \eg garbage collection and cache effects, we ran each benchmark for 1000 iterations in the same invocation of Moth, and used cut off of 350 iterations for warmup, \ie we discard the first 350 iterations, since they showed signs of compilation.
+As in our previous work \cite{roberts-and-co-ecoop-2019} which used the same set of benchmarks, to account for the complex warmup behaviour of modern systems \citep{Barrett:2017:VMW} as well as
+the non-determinism caused by \eg garbage collection and cache effects, we ran each benchmark for 1000 iterations in the same invocation of Moth, and discard the first 350 iterations to ignore warmup JIT compilation.
 
-Though though outliers, caused by \eg garbage collection, remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained for the arithmetic mean of any of the results for any of our experiments was $\pm3.53\%$ (for the PyStone benchmark).
+Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained for the arithmetic mean of any of the results for any of our experiments was $\pm3.53\%$ (for the PyStone benchmark).
 
@@
 running Fedora 29 within Windows Subsystems for Linux (version 1). The machine has 
-
 one Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
@@
 The analysis of the results and plots where generated using Python 3.7.3 and PGFPLOTS 1.16.
-To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, is available online.\footnote{
+To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, are available online.\footnote{
 \url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}

### 2019-08-17 graphs/preamble.tex
Added Section 4

@@
 % Available ACM colours: ACMBlue, ACMYellow, ACMOrange, ACMRed, ACMLightBlue, ACMGreen, ACMPurple and ACMDarkBlue
-\colorlet{GraphNormal}{ACMRed} % Normal colour when I don't care about having multiple colours
+\colorlet{GraphNormal}{ACMOrange} % Normal colour when I don't care about having multiple colours
 
@@
 	Full/.style = { % For the page of graphs for each benchmark
-		enlarge x limits={abs=2pt}, ymax={#1}, height=3.2cm, width=5.9cm,%
-		every axis plot/.style={mark size=1pt, draw=GraphNormal, fill=GraphNormal!40}},%
+		enlarge x limits={abs=1.5pt}, ymax={#1}, height=3.2cm, width=5.9cm,%
+		every axis plot/.style={very thin, mark size=0.75pt, draw=GraphNormal, fill=GraphNormal!40}},%
 	Right Axis/.style = { % For right axes
@@
 	Scatter Left Label/.style   = {ylabel={Average Time (ms)}},%
-	Scatter Bottom Label/.style = {xlabel={Type Annotations}},%
+	Scatter Bottom Label/.style = {xlabel={Proportion of Type Annotations}},%
 	Scatter Right Label/.style  = {ylabel={Relative to 0\% Typed}},%
@@
 	Pattern/.style = { % For the page of graphs for each benchmark
-		enlarge x limits={abs=1.5pt}, ymax={#1}, height=3.2cm, width=4.4cm, %
-		every axis plot/.style={mark size=0.75pt, draw=GraphBase, fill=GraphBase!40}},%
+		enlarge x limits={abs=1pt}, ymax={#1}, height=3.2cm, width=4.4cm, % 4.4
+		every axis plot/.style={very thin, mark size=0.5pt, draw=GraphBase, fill=GraphBase!40}},%
 	%
@@
 	Pattern AB/.style = {draw=GraphAB, fill=GraphAB!40, mark=*}, %
-	%Pattern Left/.style = {Left Axis={#1}, height=3.2cm, width=7.9cm}, % Size for pattern graphs
-	%Full Scatter Left/.style = {Scatter, Full Left={#1},
-	%	enlarge x limits={abs=2pt}, every axis plot/.append style={mark size={1pt}}},%
-	%Full Scatter Right/.style       = {Full Scatter Left={#1},        Right Axis},%
-	%sleft = "Pattern Scatter Left"
-	%cleft = "Pattern Column Left"
-	%sright = "Pattern Scatter Right"
-	%cright = "Pattern Column Right"
-	%Pattern Scatter A, Pattern Scatter B
 }

### 2019-08-17 paper.tex
Added Section 4

--- 2019-08-17 paper.tex
@@
 
-Though though outliers, caused by \eg garbage collection, remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained for the arithmetic mean of any of the results for any of our experiments was $\pm3.53\%$ (for the PyStone benchmark).
+Though outliers, caused by \eg garbage collection, remain visible in the plots for each individual configuration, the largest 95\% confidence interval we obtained (over the mean time after warmup) for any of experiments was $\pm3.53\%$ (for the PyStone benchmark).
 
 All experiments were executed on a Windows 10 machine
-running Fedora 29 within Windows Subsystems for Linux (version 1). The machine has
-
-one Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
+running Fedora 30 within Windows Subsystems for Linux (version 1).
+The machine has an Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads.
+We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
 Benchmarks were executed one by one to avoid interference between them.
@@
 
-give overall results - one page of monochromatic scatterplots one for
-each benchmark.
-
-need to explKn how to read scatterplots by paraphrasing
-\cite{vitousek-transient-arXive-2019}
-
-core results - i.e.\ looking at one scatter plot per annotation - the original version
-   - is there a good benchmark that is small enough to fit on one
-   page. and shows interesting results?
-
 \begin{figure*}
 	\input{graphs/full-scatter}
-	\caption{The graphs!}
+	\caption{Graphs of (at most) 102 configurations in the typing latices for each benchmark.
+Time is measured as the mean of the 351\textsuperscript{st} to the 1,000\textsuperscript{th} benchmark iteration under a single invocation of Moth (lower is better).}
+	\label{f:full}
 \end{figure*}
+The results for the performance of our sample of typing lattice configurations are presented in figure~\ref{f:full}.
+Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution if each individual configuration.
+The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations (respective).
+The execution time (in milliseconds) is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.
+
+Most of the graphs are mostly horizontal lines: indicating that the performance effect of including type annotations is negligible.
+However the plots for CD and Richards show a roughly linear increase: i.e.
+adding more type annotations to the program increases performance linearly.
+The plots for Go, Permute, DeltaBlue, and Storage even show \emph{decreases}: i.e.
+adding more type annotations can actually \emph{increase} performance.
+Of particular note is that some of the graphs appear to show bimodal performance profiles (Permute, Storage, Snake, Towers, and Richards), or in the case of List, trimodal.
 
 \section{Individual Annotations}
+\begin{figure*}
+	\input{graphs/pattern}
+	\caption{Graphs of typing latices and performance with only an individual type annotation.}
+	\label{f:pattern}		
+\end{figure*}
+Figures
+
+We hypothesise that the tri/bi-modal graphs shown in figure~\ref{f:full}, are caused by individual type annotations: i.e.
+certain annotations have a significant effect on performance.
+In order to investigate this, we ran further experiments by testing configurations with each \emph{individual} type annotation enable, and all others removed, we then used this information to approximate and see if this appears to show a relation with the results across the typing lattice.
+
+Firgure X shows the patterns we were able to identify, in particular...
+
 \label{s-individual}

### 2019-08-17 graphs/pattern.py

@@ -1,0 +1,65 @@
+#!/usr/bin/env python3
+from results import *
+
+TODO = {
+	"List-23": [17, 18], "Permute-14": 8,
+	"CD-250": 107, 	"Richards-177": [172, 173], #24,
+	"Snake-70": 61,  "Towers-30": 7,
+	"DeltaBlue-243": [154, 155], "Go-230": [31, 47],
+	"SpectralNorm-39": 27, "Json-134": 83}
+
+res = []
+i = 0
+for bs, CA in TODO.items():
+	CB = 0
+	if isinstance(CA, list):
+		CB = CA[1]
+		CA = CA[0]
+
+	b = benchmark(bs)
+	r = RESULTS[b]		
+	
+	spoints0  = [((k.types/b.types)*100, v.times.m) for k, v in r.partial.items() if CA not in v.mask and CB not in v.mask]
+	spointsA  = [((k.types/b.types)*100, v.times.m) for k, v in r.partial.items() if CA     in v.mask and CB not in v.mask]
+	spointsB  = [((k.types/b.types)*100, v.times.m) for k, v in r.partial.items() if CA not in v.mask and CB     in v.mask]
+	spointsAB = [((k.types/b.types)*100, v.times.m) for k, v in r.partial.items() if CA     in v.mask and CB     in v.mask]
+	
+	cpoints  = [(i - 0.5, v.m) for i, v in enumerate(r.single)]
+	cpointA  = (CA - 0.5, r.single[CA].m)
+	if CB != 0:
+		cpointB  = (CB - 0.5, r.single[CB].m)
+
+	ymax = max([y for _, y in spoints0 + spointsA + spointsB + spointsAB + cpoints])*1.1
+	
+	srel = ymax/r.partial[variant("0-0")].times.m
+	crel = ymax/r.single[0].m
+
+	tick_offset = floor_round(b.types/6, 5)
+
+	sleft = f"Pattern = <{ymax}>, Scatter"
+	if i % 2 == 0:
+		sleft += ", Scatter Left Label"
+	if len(TODO) - i <= 2:
+		sleft += ", Scatter Bottom Label"
+	sright = f"Pattern = <{srel}>, Scatter, Right Axis"
+	
+	cleft = f"Pattern = <{ymax}>, Column = <{tick_offset}>"	
+	if len(TODO) - i <= 2:
+		cleft += ", Column Bottom Label"
[…]

### 2019-08-17 paper.tex

--- 2019-08-17 paper.tex
@@
 One form of type checking used in gradually typed language is
-\emph{transient type checking}: when ever an object `flows' through
+\emph{transient type checking}: whenever an object `flows' through
 code with a type annotation, the object is dynamically checked to
@@
 all computation is done via dynamically dispatched  ``method requests''
-where the object receiving the request decides which
-%kjx reverted cos edited version made no sense.
-%for this paper, we could just delete the comments about blocks if you like.
-method to run, control structures are built out of blocks (essentialy lambda expressions), and return statements are ``non-local'', returning from the lexically enclosing method \citep{bluebook}.
-%
-In
+where the object receiving the request decides what code to run,
+control structures are built out of lambda expressions support ``non-local'' returns, i.e. they can return to the point where execution first encountered the lambda \citep{bluebook}.  In
 other ways, Grace is closer to JavaScript than Smalltalk: Grace
@@
 
-Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained for the arithmetic mean of any of the results for any of our experiments was $\pm3.53\%$ (for the PyStone benchmark).
+Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained (over the mean time after warmup) for any of experiments was $\pm3.53\%$ (for the PyStone benchmark).
 
 All experiments were executed on a Windows 10 machine
-running Fedora 29 within Windows Subsystems for Linux (version 1). The machine has 
-one Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
+running Fedora 30 within Windows Subsystems for Linux (version 1). The machine has 
+an Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
 Benchmarks were executed one by one to avoid interference between them.

### 2019-08-17 paper.tex
J done 4

@@
 The results for the performance of our sample of typing lattice configurations are presented in figure~\ref{f:full}.
-Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution if each individual configuration.
-The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations (respective).
-The execution time (in milliseconds) is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.
+Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution of each individual configuration.
+The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations respectively.
+The execution time in milliseconds is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.
 
-Most of the graphs are mostly horizontal lines: indicating that the performance effect of including type annotations is negligible.
-However the plots for CD and Richards show a roughly linear increase: i.e.
-adding more type annotations to the program increases performance linearly.
-The plots for Go, Permute, DeltaBlue, and Storage even show \emph{decreases}: i.e.
-adding more type annotations can actually \emph{increase} performance.
-Of particular note is that some of the graphs appear to show bimodal performance profiles (Permute, Storage, Snake, Towers, and Richards), or in the case of List, trimodal.
+Most of the graphs are essentially horizontal lines, indicating that the overhead of including type annotations is negligible.
+The plots for CD and Richards show a roughly linear increase, however, i.e.\ for these two benchmarks, adding type annotations increases performance linearly.
+On the other hand, the plots for Go, Permute, DeltaBlue, and Storage show \emph{decreases}: i.e.
+adding more type annotations \emph{increases} performance of these benchmarks.
+
+Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly horizontal lines.
+Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away.
+List appears to have three performance modes: most of the configurations up to about 60\% annotations perform at roughly the same speed as untyped code; most configurations above 70\% perform 1.8 times more slowly than untyped code; and then across all annotation percentages there are configurations that perform roughly 1.5 times slower than the untyped baseline.
 

### 2019-08-17 paper.tex
Make small copyedits in S4

@@
 	\input{graphs/full-scatter}
-	\caption{Graphs of (at most) 102 configurations in the typing latices for each benchmark.
+	\caption{Graphs of (at most) 102 configurations in the typing lattices for each benchmark.
 Time is measured as the mean of the 351\textsuperscript{st} to the 1,000\textsuperscript{th} benchmark iteration under a single invocation of Moth (lower is better).}
@@
 Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution of each individual configuration.
-The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations respectively.
+The x-axis represents the proportion of type annotations for each configuration, with the left- and right-most points showing the times for the fully untyped and typed configurations respectively.
 The execution time in milliseconds is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.
@@
 Most of the graphs are essentially horizontal lines, indicating that the overhead of including type annotations is negligible.
-The plots for CD and Richards show a roughly linear increase, however, i.e.\ for these two benchmarks, adding type annotations increases performance linearly.
+The plots for CD and Richards show a roughly linear increase, however, i.e.\ for these two benchmarks, adding type annotations reduces performance linearly.
 On the other hand, the plots for Go, Permute, DeltaBlue, and Storage show \emph{decreases}: i.e.
-adding more type annotations \emph{increases} performance of these benchmarks.
+adding more type annotations \emph{improves} performance of these benchmarks.
 
@@
 \end{figure*}
-Figures
 
-We hypothesise that the tri/bi-modal graphs shown in figure~\ref{f:full}, are caused by individual type annotations: i.e.
-certain annotations have a significant effect on performance.
-In order to investigate this, we ran further experiments by testing configurations with each \emph{individual} type annotation enable, and all others removed, we then used this information to approximate and see if this appears to show a relation with the results across the typing lattice.
+We hypothesise that the tri/bi-modal graphs shown in figure~\ref{f:full} are caused by individual type annotations: that is, particular annotations have a significant effect on performance.
+In order to investigate this, we ran further experiments by testing configurations with each \emph{individual} type annotation enabled, and all others removed, and we then used this information to approximate and see whether this appears to show a relation with the results across the typing lattice.
 

### 2019-08-17 paper.tex
J out 6 7

@@
 directly integrated into the language runtime and development environment.
-It uses feedback from the runtime to guide developers to possible
-code changes that enable a compiler to generate a more optimal program.
-In spirit, we would eventually want to achieve the same.
-Though in our case,
+Optimizer Coaching uses feedback from the runtime to guide developers
+to insert or change type declarations 
+to enable a compiler to generate a more optimal program.
+In this spirit, we would eventually want to achieve the same, although
+in our case,
 we need to run full experiments to get the necessary information.
@@
 guided by runtime feedback, cost models, or heuristics. 
-The large body of work\citep{Isaacs:2014:PerfViz} uses various approaches,
+The large body of work \citep{Isaacs:2014:PerfViz} uses various approaches,
 though, we are not aware of work that has used an approach similar to ours.
@@
 i put this in to hint at issues with usability.}
+\kjx{Hmm not sure what the last sentence here means!}
 
@@
 individual benchmark configurations for each individual typecheck:
-these results can identify problematic typechecks, but not all cases.
+these results can identify problematic typechecks, in some but not all cases.
 
@@
 \RR{The next two sentences seem to be in the wrong place (they state an advantage rather than a threat); maybe move them into the previous paragraph?}
+\kjx{I'm happy with them here about external validity}
 Moth is built
@@
 
-Finally, it is not clear how out results would transfer
+Finally, it is not clear how our results would transfer
 to other gradually typed-languages or other semantics for gradual typing.
@@
 programs overall performance. We would also like to investigate 
+whether this approach can assist with optimsations for programmers' day-to-day  development, or help VM engineers identifying performance bugs in the underlying virtual machines.
 
@@
 What's its context, use, general application?}
+\kjx{filled out the last sentence}
 

### 2019-08-17 paper.tex
Horrible section 5

--- 2019-08-17 paper.tex
@@
 
-Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained (over the mean time after warmup) for any of experiments was $\pm3.53\%$ (for the PyStone benchmark).
+Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained (over the mean time after warmup) for any of experiments was $\pm3.36\%$ (for the PyStone benchmark).
 
 All experiments were executed on a Windows 10 machine
-running Fedora 30 within Windows Subsystems for Linux (version 1). The machine has 
-an Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads. We used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
+running Fedora 30 within Windows Subsystems for Linux (version 1).
+The machine has an Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads.
+This computer appears to produce comparable results to the one used in our previous work, although the absolute performance is higher, the relative overhead of transient type-checks is similar.
+
+We used the same version of Moth as previously, together with ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
 Benchmarks were executed one by one to avoid interference between them.
 The analysis of the results and plots where generated using Python 3.7.3 and PGFPLOTS 1.16.
-To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, are available online.\footnote{
-\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}
+To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, are available online.\footnote{\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}
 
@@
 The results for the performance of our sample of typing lattice configurations are presented in figure~\ref{f:full}.
-Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution of each individual configuration.
-The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations respectively.
-The execution time in milliseconds is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.
+Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution if each individual configuration.
+The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations (respective).
+The execution time (in milliseconds) is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.
 
-Most of the graphs are essentially horizontal lines, indicating that the overhead of including type annotations is negligible.
-The plots for CD and Richards show a roughly linear increase, however, i.e.\ for these two benchmarks, adding type annotations increases performance linearly.
-On the other hand, the plots for Go, Permute, DeltaBlue, and Storage show \emph{decreases}: i.e.
-adding more type annotations \emph{increases} performance of these benchmarks.
-
-Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly horizontal lines.
-Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away.
-List appears to have three performance modes: most of the configurations up to about 60\% annotations perform at roughly the same speed as untyped code; most configurations above 70\% perform 1.8 times more slowly than untyped code; and then across all annotation percentages there are configurations that perform roughly 1.5 times slower than the untyped baseline.
+Most of the graphs are mostly horizontal lines: indicating that the performance effect of including type annotations is negligible.
+However the plots for CD and Richards show a roughly linear increase: i.e.
+adding more type annotations to the program increases performance linearly.
+The plots for Go, Permute, DeltaBlue, and Storage even show \emph{decreases}: i.e.
+adding more type annotations can actually \emph{increase} performance.
+Of particular note is that some of the graphs appear to show bimodal performance profiles (Permute, Storage, Snake, Towers, and Richards), or in the case of List, trimodal.
 
 \section{Individual Annotations}
+\label{s-individual}
 \begin{figure*}
 	\input{graphs/pattern}
-	\caption{Graphs of typing latices and performance with only an individual type annotation.}
+	\caption{Pairs of colour coded scatter and column graphs.
+The scatter graphs represent the performance of a sample of typing lattice.
+The column graphs show the performance of every configuration with only one type annotation.
+The scatter plots and columns graphs are colour coded based on whether a particular type annotation or two are present in the source code.}
 	\label{f:pattern}		
 \end{figure*}
-Figures
-
 We hypothesise that the tri/bi-modal graphs shown in figure~\ref{f:full}, are caused by individual type annotations: i.e.
 certain annotations have a significant effect on performance.
-In order to investigate this, we ran further experiments by testing configurations with each \emph{individual} type annotation enable, and all others removed, we then used this information to approximate and see if this appears to show a relation with the results across the typing lattice.
+In order to identify such type annotations, for each benchmark and type annotation, we manually inspected the scatter plots for the benchmark's typing lattice, coloured based on whether the chosen type annotation was present or not.
+This resulted in 1,775 graphs, however most of these showed no discernible pattern.
 
-Firgure X shows the patterns we were able to identify, in particular...
+As an alternative, and less labour intensive approach to identifying potentially influential type annotations, for each type annotation we ran each benchmark with only that type annotation present.
+We used the same computer and number of iterations as for the typing lattice experiments, albeit these were done on different days.
 
-\label{s-individual}
+Figure \ref{f:pattern} shows the results of these experiments, as well as the patterns we manually identified.
 
-We hypothesised that the overhead of an individual type annotation may
-be able to predict the performance of program configurations with that
-annotation. To test this hypothesis, we constructed a series of
-configurations, one for each type annotation, with only that
-one annotation included.
+The column graphs show the execution time of configurations with with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the 1\textsubscript{st} annotation, and the last column represents the last one (in the order they appear in the source code).
+The y-axes for the column graphs are the same as the associated scatter graph.
+However, though the right y-axes for the column graphs appear to be identical the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments where run.
 
-   trying-to-be-cleverer results
+For each benchmark we highlighted one or two type annotations that appear to show a pattern for the typing lattice performance, or in the case of SpectralNorm and Json had higher than usual columns.
+The identified types are those represented by the red and blue columns.
+The scatter plots are colour coded accordingly: red/blue circle represents those with the given type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
 
-   - single annotation overheads
-  - using them go back to scatterplots (i.e. what you do in the latest graphs)
+As can bee seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
+For the Snake, Towers and CD benchmark, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
+For permute there is actually a signficant performance \emph{incrase}, indicated by the lower half of the lattice being highlighted in red.
+DeltaBlue and Go also show a slight incrase in performance, but here this is caused by two type annotations (in red and blue), however this does not appear to be comulative (as indicated by the purple dots being about the same height as the red and blue ones).
 
-  - show that single annotation prediction doesn't really work\ldots
+List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present.
+Richards is particularly od as it shows a performance decrease that appears to occur when the red type annotations is present, but not the blue one.
+
+Finally, SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a pattern in the typing lattice.
+
+Note however that in the case of DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively), however we found no apparent relation between these and the typing lattice.
+
+The remaining benchmarks, which we have not shown in \ref{f:pattern}, have relatively flat column graphs and we could not identify any patterns in the typing lattice.
+The benchmark with the greatest difference in single type configurations relative to the untyped configuration is Storage, with a configuration only $7.9\%$ faster than the untyped baseline.
+However, though Storage (as shown in \ref{f:full}) appears to be bimodal like Towers, we were unable to find any relationships between its typing lattice and any individual type annotations.
+%For example for Snake, the dots covered in red (corresponding to configurations with type annotation number 61) are noticeably higher than the grey ones (those without type annotation 61), thus it appears that the 61\textsuperscript{st} type annotation causes a noticeable slowdown, which is also exemplified by the higher than average red column.
+The graphs for CD (type \#107), Towers (type \#7) show similar patterns.
+The graph for Permute is similar, but it shows that type annotation 8 actually makes the benchmarks \emph{faster}.
+%In some cases, more than two type annotations have an effect, for DeltaBlue and Go it appears that types 154/155, and 31/47 (respectively) all cause an increase in performance, but this is not cumulative.
+%However with List, the performance identified by type annotations 17 and 18 is actually comulative
+
+%For most of the graphs, the identified type annotations are the same, however this is not always the case.
+For SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), there appears to be no relationship with the typing lattice.
+%For DeltaBlue and Richards, type annotations 62 and 24 on the column graphs appear to be significant, however we did not see any kind of pattern their.
 

### 2019-08-17 graphs/pattern.py

@@ -2,10 +2,10 @@
 from results import *
 
 TODO = {
-	"List-23": [17, 18], "Permute-14": 8,
-	"CD-250": 107, 	"Richards-177": [172, 173], #24,
-	"Snake-70": 61,  "Towers-30": 7,
+	"Snake-70": 61, "Towers-30": 7,
+	"CD-250": 107, "Permute-14": 8,
 	"DeltaBlue-243": [154, 155], "Go-230": [31, 47],
+	"List-23": [17, 18], "Richards-177": [172, 173], #24,
 	"SpectralNorm-39": 27, "Json-134": 83}
 
 res = []
@@ -59,7 +59,7 @@ for bs, CA in TODO.items():
 		
 	res.append(double_axis(f"{cleft} = <{ymax}><{floor_round(b.types/6, 5)}>", f"\\null\\hfill<\\small({b.types} Type Annotations)>", f"{cright} = <{crel}>", 
 		plot("", *cpoints[1:], (len(r.single) - 0.5, 0), (0.5, 0)), *columns))
-	
+
 	i += 1
 	
 graphs(4, *res)

### 2019-08-17 paper.tex
Add paragraph on 0-100 bounds to S4

@@
 The results for the performance of our sample of typing lattice configurations are presented in figure~\ref{f:full}.
-Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution if each individual configuration.
-The x-axis represents the proportion of type annotations for each configuration, with the left and right most points showing the times for the fully untyped and typed configurations (respective).
-The execution time (in milliseconds) is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.
+Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution of each individual configuration.
+The x-axis represents the proportion of type annotations for each configuration, with the left- and right-most points showing the times for the fully untyped and typed configurations respectively.
+The execution time in milliseconds is shown on the left y-axes, and time relative to the fully untyped configuration is shown on the right y-axes.
 
-Most of the graphs are mostly horizontal lines: indicating that the performance effect of including type annotations is negligible.
-However the plots for CD and Richards show a roughly linear increase: i.e.
-adding more type annotations to the program increases performance linearly.
-The plots for Go, Permute, DeltaBlue, and Storage even show \emph{decreases}: i.e.
-adding more type annotations can actually \emph{increase} performance.
-Of particular note is that some of the graphs appear to show bimodal performance profiles (Permute, Storage, Snake, Towers, and Richards), or in the case of List, trimodal.
+Most of the graphs are essentially horizontal lines, indicating that the overhead of including type annotations is negligible.
+The plots for CD and Richards show a roughly linear increase, however, i.e.\ for these two benchmarks, adding type annotations reduces performance linearly.
+On the other hand, the plots for Go, Permute, DeltaBlue, and Storage show \emph{decreases}: i.e.
+adding more type annotations \emph{improves} performance of these benchmarks.
+
+By inspecting the scatterplots, we observe that the performance of
+Moth in almost any configuration of a benchmark is bounded by
+the performance in the untyped and fully-typed configuration.
+That is, for these benchmarks on transient type checks,
+measuring just the untyped and fully-typed configurations
+would provide excellent estimates of a benchmark's
+performance bounds.  This is different to the experience
+of other kinds of gradual typing, where the best, and most
+importantly worst, configurations are not always those
+fully typed or fully untyped \cite{greenam-takikiaw-JFP}.
+%
+However, the Richards and Snake benchmarks \textit{do} have
+sections outside these bounds, and isolated executions of
+a few others (SpectralNorm, JSON, DeltaBlue) are also outliers.
+%
+We believe that the 0\% and 100\% bounds are nonetheless a
+reasonable heuristic estimator in most cases, and will
+examine further the outlying benchmarks.
+
+Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly-horizontal lines.
+Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away.
+List appears to have three performance modes: most of the configurations up to about 60\% annotations perform at roughly the same speed as untyped code; most configurations above 70\% perform 1.8 times more slowly than untyped code; and then across all annotation percentages there are configurations that perform roughly 1.5 times slower than the untyped baseline.
 

### 2019-08-17 paper.tex

@@
 
-By inspecting the scatterplots, we observe that the performance of
-Moth in almost any configuration of a benchmark is bounded by
-the performance in the untyped and fully-typed configuration.
-That is, for these benchmarks on transient type checks,
-measuring just the untyped and fully-typed configurations
-would provide excellent estimates of a benchmark's
-performance bounds.  This is different to the experience
-of other kinds of gradual typing, where the best, and most
-importantly worst, configurations are not always those
-fully typed or fully untyped \cite{greenam-takikiaw-JFP}.
-%
-However, the Richards and Snake benchmarks \textit{do} have
-sections outside these bounds, and isolated executions of
-a few others (SpectralNorm, JSON, DeltaBlue) are also outliers.
-%
-We believe that the 0\% and 100\% bounds are nonetheless a
-reasonable heuristic estimator in most cases, and will
-examine further the outlying benchmarks.
-
-Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly-horizontal lines.
+Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly horizontal lines.
 Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away.

### 2019-08-17 paper.tex
Fix cite in 0-100 paragraph

--- 2019-08-17 paper.tex
@@
 importantly worst, configurations are not always those
-fully typed or fully untyped \cite{greenam-takikiaw-JFP}.
+fully typed or fully untyped \cite{Greenman2019jfp}.
 %

### 2019-08-17 paper.tex
Edit pass through S5 Individual Annotations

@@
 The y-axes for the column graphs are the same as the associated scatter graph.
-However, though the right y-axes for the column graphs appear to be identical the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments where run.
+However, though the right y-axes for the column graphs appear to be identical the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.
 
@@
 
-As can bee seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
+As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
 For the Snake, Towers and CD benchmark, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
-For permute there is actually a signficant performance \emph{incrase}, indicated by the lower half of the lattice being highlighted in red.
-DeltaBlue and Go also show a slight incrase in performance, but here this is caused by two type annotations (in red and blue), however this does not appear to be comulative (as indicated by the purple dots being about the same height as the red and blue ones).
+For Permute there is actually a signficant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.
+DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this does not appear to be cumulative, however, as the purple dots are around the same height as the red and blue ones: \textit{either} annotation is sufficient for the full improvement.
 
 List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present.
-Richards is particularly od as it shows a performance decrease that appears to occur when the red type annotations is present, but not the blue one.
+Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotations is present, but not the blue one.
 
-Finally, SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a pattern in the typing lattice.
+Finally, for SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a pattern in the typing lattice.
 
-Note however that in the case of DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively), however we found no apparent relation between these and the typing lattice.
+Note however that in the case of DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively); nonetheless, we found no apparent relation between these and the typing lattice.
 

### 2019-08-17 paper.tex
Further edit pass through S5

@@
 \end{figure*}
-We hypothesise that the tri/bi-modal graphs shown in figure~\ref{f:full}, are caused by individual type annotations: i.e.
-certain annotations have a significant effect on performance.
-In order to identify such type annotations, for each benchmark and type annotation, we manually inspected the scatter plots for the benchmark's typing lattice, coloured based on whether the chosen type annotation was present or not.
+We hypothesise that the tri/bi-modal graphs, shown in figure~\ref{f:full}, are caused by individual type annotations: that is, certain annotations have a significant effect on performance.
+In order to identify such type annotations, for each benchmark and type annotation we manually inspected the scatter plots for the benchmark's typing lattice, coloured based on whether the chosen type annotation was present or not.
 This resulted in 1,775 graphs, however most of these showed no discernible pattern.
 
-As an alternative, and less labour intensive approach to identifying potentially influential type annotations, for each type annotation we ran each benchmark with only that type annotation present.
+As an alternative, and less labour-intensive approach to identifying potentially-influential type annotations, for each type annotation we ran each benchmark with \textit{only} that type annotation present.
 We used the same computer and number of iterations as for the typing lattice experiments, albeit these were done on different days.
@@
 List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present.
-Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotations is present, but not the blue one.
+Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotation is present, but not the blue one.
 
@@
 The remaining benchmarks, which we have not shown in \ref{f:pattern}, have relatively flat column graphs and we could not identify any patterns in the typing lattice.
-The benchmark with the greatest difference in single type configurations relative to the untyped configuration is Storage, with a configuration only $7.9\%$ faster than the untyped baseline.
+The benchmark with the greatest difference in single-type configurations relative to the untyped configuration is Storage, with a configuration only $7.9\%$ faster than the untyped baseline.
 However, though Storage (as shown in \ref{f:full}) appears to be bimodal like Towers, we were unable to find any relationships between its typing lattice and any individual type annotations.

### 2019-08-17 paper.tex
Conclusion!

--- 2019-08-17 paper.tex
@@
 away, some will significantly decrease a program's
-performance, and some may even increase it.  In this paper we investigate techniques that interpret benchmark results to identify which type checks have the greatest effects on performance.
+performance, and some may even increase it.  In this paper we show techniques where benchmark results can be used to help identify which type annotations have the greatest effects on performance.
 Using these techniques,
@@
 
-By inspecting the scatterplots, we observe that the performance of
-Moth in almost any configuration of a benchmark is bounded by
-the performance in the untyped and fully-typed configuration.
-That is, for these benchmarks on transient type checks,
-measuring just the untyped and fully-typed configurations
-would provide excellent estimates of a benchmark's
-performance bounds.  This is different to the experience
-of other kinds of gradual typing, where the best, and most
-importantly worst, configurations are not always those
-fully typed or fully untyped \cite{Greenman2019jfp}.
-%
-However, the Richards and Snake benchmarks \textit{do} have
-sections outside these bounds, and isolated executions of
-a few others (SpectralNorm, JSON, DeltaBlue) are also outliers.
-%
-We believe that the 0\% and 100\% bounds are nonetheless a
-reasonable heuristic estimator in most cases, and will
-examine further the outlying benchmarks.
-
-Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly-horizontal lines.
+Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly horizontal lines.
 Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away.
@@
 
-In this paper we have investigated how benchmark results can be
-repurposed to determine precisely which transient typechecks are
+In this paper we have investigated how benchmarks can be
+repurposed to determine precisely which transient type checks are
 resistant to the optimisations provided by a just-in-time compiling
@@
 of the benchmark configurations ran significantly slower than
-the remaining configurations. By inspecting the benchmark graphs coded
-for each typecheck in a program's source code, we were often able to
-identify that just one or two of the typechecks were responsible for
-the bimodal performance.  Based on that result, we also investigated
-individual benchmark configurations for each individual typecheck:
-these results can identify problematic typechecks, in some but not all cases.
+the remaining configurations. We also observed trimodal profiles, as well as performance increases when type annotations where added.
+By inspecting benchmark graphs colour coded for each individual type annotation in a program's source code, we were often able to
+identify that just one or two of the type checks were likely responsible for
+the bimodal performance, by observing patterns in the associated graph.
+However, this approach requires inspecting one graph per type annotation.
 
-This is preliminary work: there are a number of threats to validity. 
+We also investigated another approach to help identify such annotation, for an individual benchmark, we can produce a single graph of the performances of individual configurations for each type annotation.
+In most cases both approaches identified the same type annotations as likely causing a significant effect on performance.
+
+This is preliminary work, in particular we have not yet identified exactly why these type annotations appear to have an effect on performance.
+
+There are also a number of threats to validity. 
 Regarding construct validity, 
@@
 Regarding external validity, 
-\RR{The next two sentences seem to be in the wrong place (they state an advantage rather than a threat); maybe move them into the previous paragraph?}
-\kjx{I'm happy with them here about external validity}
+%\RR{The next two sentences seem to be in the wrong place (they state an advantage rather than a threat); maybe move them into the previous paragraph?}
+%\kjx{I'm happy with them here about external validity}
 Moth is built
@@
 to resemble other Graal
-VMs doing similar AST-based optimizations of transient typechecks.
+VMs doing similar AST-based optimizations of transient type checks.
 Because we rely on common techniques, 
-we expect our results should be transferable to JIT implementations
-more widely.
+we expect our results to be transferable to other JIT implementations as well.
 
@@
 Our benchmarks do not depend on any features of Grace
-that are not common in other gradually-typed object-oriented
+that are not common in other object-oriented
 languages, but as Grace lacks a large corpus of programs the
 benchmarks are necessarily small and artificial.
-The advantage of Grace and Moth for this research is
+The advantage of Grace for this research is
 that their relative simplicity means we have been able to build an
@@
 whether this approach can assist with optimsations for programmers' day-to-day  development, or help VM engineers identifying performance bugs in the underlying virtual machines.
-
 \sm{is our approach practical. Is it useful for day-to-day development?

### 2019-08-17 paper.tex
Manually merge kjx edits to S5

@@
         \input{graphs/pattern}
-	\caption{Pairs of colour coded scatter and column graphs.
+        \caption{Pairs of colour-coded scatter and column graphs.
 The scatter graphs represent the performance of a sample of typing lattice.
 The column graphs show the performance of every configuration with only one type annotation.
-The scatter plots and columns graphs are colour coded based on whether a particular type annotation or two are present in the source code.}
+The scatter plots and column graphs are colour-coded based on whether a particular type annotation or two are present in the source code.}
         \label{f:pattern}              
 \end{figure*}
-We hypothesise that the tri/bi-modal graphs, shown in figure~\ref{f:full}, are caused by individual type annotations: that is, certain annotations have a significant effect on performance.
-In order to identify such type annotations, for each benchmark and type annotation we manually inspected the scatter plots for the benchmark's typing lattice, coloured based on whether the chosen type annotation was present or not.
-This resulted in 1,775 graphs, however most of these showed no discernible pattern.
+We hypothesise that the bi/tri-modal graphs shown in figure~\ref{f:full} are caused by a few, highly significant type annotations: i.e.\ there are a very few annotaitons that determine each benchmark's performance.
 
-As an alternative, and less labour-intensive approach to identifying potentially-influential type annotations, for each type annotation we ran each benchmark with \textit{only} that type annotation present.
-We used the same computer and number of iterations as for the typing lattice experiments, albeit these were done on different days.
+In order to identify those type annotations, we generated coloured
+scatter plots for each benchmark based on whether each individual type
+annotation was present in that configuration or not. The position of
+the dots in each coloured plot are the same as Figure~\ref{f:full}, but
+they are coloured as to whether each annotation is there or not: This resulted in 1,775 scatterplots --- one for each type annotation for each benchmark.
+\kjx{if we say this, we should show a bunch for one benchmark.}
+We then manually inspected all these scatterplots,
+but unfortunately most of these showed no discernible pattern.
+\kjx{YEAH BUT SOME DID. Shouldn't we show at least one?}
 
-Figure \ref{f:pattern} shows the results of these experiments, as well as the patterns we manually identified.
+As an alternative, less labour intensive approach to identifying
+potentially influential type annotations, we measured one additional
+configuration for each type annotation in each benchmark, with only
+that single type annotation present. Our aim was to compare the overhead of
+each type annotation in isolation against the baseline without any
+type checks.
 
-The column graphs show the execution time of configurations with with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the 1\textsubscript{st} annotation, and the last column represents the last one (in the order they appear in the source code).
-The y-axes for the column graphs are the same as the associated scatter graph.
-However, though the right y-axes for the column graphs appear to be identical the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.
+% We used the same computer and number of iterations as for the typing
+% lattice experiments, albeit these were done on different days.
+
+Figure \ref{f:pattern} shows the results of these experiments.
+%, as well as the patterns we manually identified.
+%%KJX but you haven't talked about manual patterns!
+
+This figure shows a pair of graphs for a selection of ten benchmarks: a scatterplot and a column graph.
+\kjx{are they actualy histograpms?
+or just bar charts?}\mwh{It's a column graph, like Isaac said - histograms have a continuous axis and bar charts are spaced.}
+\kjx{argh. should figure do histograph first then scatterplot?}
+The column graphs show the execution times of each configurations for each individual annotation in a benchmark.
+The x-axis of the graphs indicates the index of this annotation in the order they appear in the source code.
+The first column represents the 1\textsubscript{st} annotation, and the last column represents the last one.
+The y-axes for the column graphs are the same as the associated scatter graph, showing the execution time of each configuration.\footnote{Although the right y-axes for the column graphs appear to be identical the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}\kjx{do we need the footnote?}
 
@@
 The identified types are those represented by the red and blue columns.
-The scatter plots are colour coded accordingly: red/blue circle represents those with the given type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
+The scatter plots are colour-coded accordingly: a red or blue circle represents those with the given type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
 
@@
 For Permute there is actually a signficant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.
-DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this does not appear to be cumulative, however, as the purple dots are around the same height as the red and blue ones: \textit{either} annotation is sufficient for the full improvement.
+DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue), but this does not appear to be cumulative: the purple dots are about the same height as the red and blue ones, so \textit{either} annotation is sufficient for the full benefit.
 
 List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present.
-Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotation is present, but not the blue one.
+Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotations is present, but not the blue one.
 
-Finally, for SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a pattern in the typing lattice.
-
-Note however that in the case of DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively); nonetheless, we found no apparent relation between these and the typing lattice.
+For SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a matching pattern in the configuration performance.
+Finally, for DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively), however we again found no apparent relation between these outliers and the overall performance
 

### 2019-08-17 paper.tex
Make new/old copyedits in S5

@@
 \end{figure*}
-We hypothesise that the bi/tri-modal graphs shown in figure~\ref{f:full} are caused by a few, highly significant type annotations: i.e.\ there are a very few annotaitons that determine each benchmark's performance.
+We hypothesise that the bi/tri-modal graphs shown in figure~\ref{f:full} are caused by a few, highly significant type annotations: i.e.\ there are a very few annotations that determine each benchmark's performance.
 
@@
 that single type annotation present. Our aim was to compare the overhead of
-each type annotation in isolation against the baseline without any
-type checks.
+each type annotation in isolation against the no-typecheck baseline.
 
@@
 The first column represents the 1\textsubscript{st} annotation, and the last column represents the last one.
-The y-axes for the column graphs are the same as the associated scatter graph, showing the execution time of each configuration.\footnote{Although the right y-axes for the column graphs appear to be identical the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}\kjx{do we need the footnote?}
+The y-axes for the column graphs are the same as the associated scatter graph, showing the execution time of each configuration.\footnote{Although the right y-axes for the column graphs appear to be identical to the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}\kjx{do we need the footnote?}
 
@@
 As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
-For the Snake, Towers and CD benchmark, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
+For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
 For Permute there is actually a signficant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.
-DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue), but this does not appear to be cumulative: the purple dots are about the same height as the red and blue ones, so \textit{either} annotation is sufficient for the full benefit.
+DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect does not appear to be cumulative, however: the purple dots are about the same height as the red and blue ones, so \textit{either} annotation is sufficient for the full benefit.
 

### 2019-08-17 paper.tex
Added a set of contributions to the intro
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 particular transient type checks are resistant to optimisation.
+
+Specifically, the contributions of this work are:
+
+\begin{itemize}
+  \item an approach to identify gradual type annotations
+        that cause significant performance effects
+  \item adapting the Takikawa protocol for languages and implementations
+        where performance effects are dominated by type annotations instead
+        of transitions between typed and untyped modules
+  \item observing that performance of small benchmarks
+        with 10--250 type annotations is dominated by only
+        one or two type annotations
+\end{itemize}
+
 The next section discusses dynamic type checks and gradual typing in

### 2019-08-17 paper.tex
Emphasise our contribuition more

--- 2019-08-17 paper.tex
@@
 
-\author{Isaac Oscar Gariano}
+\author[Isaac O. G.]{Isaac Oscar Gariano}
 \affiliation{
@@
 
-\author{Richard Roberts}
+\author[Richard R.]{Richard Roberts}
 \affiliation{
@@
 
-\author{Stefan Marr}
+\author[Stefan M.]{Stefan Marr}
 \orcid{0000-0001-9059-5180}
@@
 
-\author{Michael Homer}
+\author[Michael H.]{Michael Homer}
 \affiliation{
@@
 
-\author{James Noble}
+\author[James N.]{James Noble}
 \orcid{0000-0001-9036-5692}             %% \orcid is optional
@@
 away, some will significantly decrease a program's
-performance, and some may even increase it.  In this paper we show techniques where benchmark results can be used to help identify which type annotations have the greatest effects on performance.
-Using these techniques,
-programmers can optimise programs by removing expensive type checks,
-and VM engineers can identify new opportunities for compiler optimisation.
+performance, and some may even increase it.
 
-\sm{this is much to brief for me.
-Transient typechecks is a very specific thing.
-I believe, VM people need more context here.
-It's a form of gradual typing, which requires comparably simple checks
-at run time to check type violations.
-Then, how do we investigate the issues?
-we should give a hint what the solution is.
-otherwise, the sentence in using these techniques doesn't make sense to me, because we didn't say anything about them.}
-\kjx{still like brief abstracts. at least I defined ``techniques''}
+In this paper, we refine the so called ``Takikawa'' protocol, and use it to identify which type annotations have the greatest effects on performance.
+In particular, we show how graphing the performance of such benchmarks when varying which type annotations are present in the source code can be used to discern potential patterns in performance.
+We demonstrate our approach by testing the Moth virtual machine: for many of the benchmarks where Moth's transient type checking impacts performance, we have been able to identify one or two specific type annotations that are the likely cause.
+Without these type annotations, the performance impact of transient type checking becomes negligible.
+
+Using our technique programmers can optimise programs by removing expensive type checks, and VM engineers can identify new opportunities for compiler optimisation.
 \end{abstract}
@@
 \keywords{dynamic type checking, gradual types, optional types, Grace,
-Moth, object-oriented programming}
+object-oriented programming}
 \maketitle
@@
 
-Gradual Typing aims to add static type annotations to dynamic languages, increasing
+Gradual typing aims to add static type annotations to dynamic languages, increasing
 their safety while maintaining flexibility
@@
 
-\section{Evaluating Transient Typechecks}
-\label{s-eval}
+\section{Experimental Methodology}
+Our methodology is built upon the so-called ``Takikawa'' or ``Takikawa-Greenman'' evaluation protocol \cite{Takikawa2016,Greenman2019jfp} benchmarks a \emph{lattice} of $2^N$ \emph{configurations} of each benchmark, where a configuration is a particular mix of static and dynamically typed code, however here we only test a relatively small sample of this lattice.
+Here we apply this methodology to our previous work \cite{roberts-and-co-ecoop-2019}, where we evaluated the performance of transient type checks in Moth by using a suite of benchmarks.
 
-In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the
-performance of transient type checks using 21 benchmarks.
-We started with the 14 benchmarks of the \AWFY
+\subsection{The Takikawa Protocol}
+The Takikawa evaluation protocol was originally proposed for Typed
+Racket, where static vs dynamic typing is set per-module, so $N$ is
+the number of modules. Grace, following other languages such as
+Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
+allows programmers to choose whether each individual declaration should be type-checked. This means $N$ in Grace is the number of type annotations in the program, and so
+checking an entire lattice for even a moderately sized benchmark would be infeasible.  Vitousek et al.\ modified the Takikawa protocol for
+these kinds of languages by taking an approach based on sampling
+\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
+divides the number of type annotations in a fully-typed program into
+a maximum of 100 intervals, and then randomly generates ten programs within
+each interval by erasing type annotations.
+
+To investigate Moth's support for transient type checks further, we carried out a similar approach to the Takikawa-Vitousek protocol.
+For each benchmark we generated 100 partially typed versions (fewer if the benchmark was too small).
+We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations.
+We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to choose the type annotations each configuration contained randomly, and we ensured that no duplicate configurations were generated.
+In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark (or 97 in the case of our Storage benchmark, since it only has 10 type annotations).
+
+\subsection{The Benchmarks}
+In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the performance of transient type checks using 21 benchmarks. We started with the 14 benchmarks of the \AWFY
 benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.
@@
 with our implementation technique we expected fully typed programs to
-have the largest overhead. This is not the case for other kinds of
-gradual type checking, which have given rise to more complex
-evaluation techniques.  In particular, the so-called ``Takikawa'' or
-``Takikawa-Greenman'' evaluation protocol \cite{Takikawa2016,Greenman2019jfp}
-benchmarks a \emph{lattice} of $2^N$ \emph{configurations} of each
-benchmark, where a configuration is a particular mix of static and
-dynamically typed code.
+have the largest overhead, which is not the case for other kinds of
+gradual type checking.
 
-The Takikawa evaluation protocol was originally proposed for Typed
-Racket, where static vs dynamic typing is set per-module, so $N$ is
-the number of modules. Grace, following other languages such as
-Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
-allows programmers to choose whether each individual declaration should be type-checked. This means $N$ in Grace is the number of type annotations in the program, and so
-checking an entire lattice for even a moderately sized benchmark would be infeasible.  Vitousek et al.\ modified the Takikawa protocol for
-these kinds of languages by taking an approach based on sampling
-\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
-divides the number of type annotations in a fully-typed program into
-a maximum of 100 intervals, and then randomly generates ten programs within
-each interval by erasing type annotations.
-
-To investigate Moth's support for transient type checks further, we carried out a similar approach to the Takikawa-Vitousek protocol.
-For each benchmark we generated 100 partially typed versions (fewer if the benchmark was too small).
-We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations.
-We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to choose the type annotations each configuration contained randomly, and we ensured that no duplicate configurations were generated.
-In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark, (or 97 in the case of Storage, since it only has 10 type annotations).
-
+\subsection{Experimental Set Up}
 As in our previous work \cite{roberts-and-co-ecoop-2019} which used the same set of benchmarks, to account for the complex warmup behaviour of modern systems \citep{Barrett:2017:VMW} as well as
@@
 	\input{graphs/pattern}
-        \caption{Pairs of colour-coded scatter and column graphs.
-The scatter graphs represent the performance of a sample of typing lattice.
+	\caption{Pairs of colour coded scatter and column graphs.
+The scatter graphs represent the performance of a sample of the typing lattices.
 The column graphs show the performance of every configuration with only one type annotation.
-The scatter plots and column graphs are colour-coded based on whether a particular type annotation or two are present in the source code.}
+The scatter plots and column graphs are colour coded based on whether a particular type annotation or two are present in the source code.}
 	\label{f:pattern}		
 \end{figure*}
-We hypothesise that the bi/tri-modal graphs shown in figure~\ref{f:full} are caused by a few, highly significant type annotations: i.e.\ there are a very few annotaitons that determine each benchmark's performance.
+We hypothesise that the tri/bi-modal graphs, shown in figure~\ref{f:full}, are caused by individual type annotations: that is, certain annotations have a significant effect on performance.
+In order to identify such type annotations, for each benchmark and type annotation we manually inspected the scatter plots for the benchmark's typing lattice, coloured based on whether the chosen type annotation was present or not.
+This resulted in 1,775 graphs, however most of these showed no discernible pattern.
 
-In order to identify those type annotations, we generated coloured
-scatter plots for each benchmark based on whether each individual type
-annotation was present in that configuration or not. The position of
-the dots in each coloured plot are the same as Figure~\ref{f:full}, but
-they are coloured as to whether each annotation is there or not: This resulted in 1,775 scatterplots --- one for each type annotation for each benchmark.
-\kjx{if we say this, we should show a bunch for one benchmark.}
-We then manually inspected all these scatterplots,
-but unfortunately most of these showed no discernible pattern.
-\kjx{YEAH BUT SOME DID. Shouldn't we show at least one?}
+As an alternative, and less labour intensive approach to identifying potentially influential type annotations, for each type annotation we ran each benchmark with \textit{only} that type annotation present.
+We used the same computer and number of iterations as for the typing lattice experiments, albeit these were done on different days.
 
-As an alternative, less labour intensive approach to identifying
-potentially influential type annotations, we measured one additional
-configuration for each type annotation in each benchmark, with only
-that single type annotation present. Our aim was to compare the overhead of
-each type annotation in isolation against the baseline without any
-type checks.
+Figure \ref{f:pattern} shows the results of these experiments, as well as the patterns we manually identified.
 
-% We used the same computer and number of iterations as for the typing
-% lattice experiments, albeit these were done on different days.
-
-Figure \ref{f:pattern} shows the results of these experiments.
-%, as well as the patterns we manually identified.
-%%KJX but you haven't talked about manual patterns!
-
-This figure shows a pair of graphs for a selection of ten benchmarks: a scatterplot and a column graph.
-\kjx{are they actualy histograpms?
-or just bar charts?}\mwh{It's a column graph, like Isaac said - histograms have a continuous axis and bar charts are spaced.}
-\kjx{argh. should figure do histograph first then scatterplot?}
-The column graphs show the execution times of each configurations for each individual annotation in a benchmark.
-The x-axis of the graphs indicates the index of this annotation in the order they appear in the source code.
-The first column represents the 1\textsubscript{st} annotation, and the last column represents the last one.
-The y-axes for the column graphs are the same as the associated scatter graph, showing the execution time of each configuration.\footnote{Although the right y-axes for the column graphs appear to be identical the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}\kjx{do we need the footnote?}
+The column graphs show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
+The y-axes for the column graphs are the same as the associated scatter graphs.
+However, though the right y-axes for the column graphs appear to be identical to the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.
 
@@
 The identified types are those represented by the red and blue columns.
-The scatter plots are colour-coded accordingly: a red or blue circle represents those with the given type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
+The scatter plots are colour coded accordingly: red/blue circles represent configurations with the red/blue type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
 
 As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
-For the Snake, Towers and CD benchmark, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
+For the Snake, Towers, and CD benchmark, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
 For Permute there is actually a signficant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.
-DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue), but this does not appear to be cumulative: the purple dots are about the same height as the red and blue ones, so \textit{either} annotation is sufficient for the full benefit.
+DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this does not appear to be cumulative since the purple dots are around the same height as the red and blue ones: thus \textit{either} annotation is sufficient for the full improvement.
 
 List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present.
-Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotations is present, but not the blue one.
+Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotation is present, but not the blue one.
+We had previously identified this benchmark and the afformentioned type annotations \cite{roberts-and-co-ecoop-2019} as being the worst case for Moth.
 
-For SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a matching pattern in the configuration performance.
-Finally, for DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively), however we again found no apparent relation between these outliers and the overall performance
+Finally, for SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a pattern in the typing lattice.
 
-The remaining benchmarks, which we have not shown in \ref{f:pattern}, have relatively flat column graphs and we could not identify any patterns in the typing lattice.
-The benchmark with the greatest difference in single-type configurations relative to the untyped configuration is Storage, with a configuration only $7.9\%$ faster than the untyped baseline.
+Note however that in the case of DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively); nonetheless, we found no apparent relation between these and the typing lattice.
+
+In particular, observe that the grey dots for Snake, Towers, Permute, DeltaBlue, Go, and List are mostly flat horizontal lines, this indicates that by simply deleting the red and blue type annotations, the performance impact of transient type checking becomes negligible.
+However for CD and Richards, the transient type checking overhead of the grey dots appears linear, albeit still less than with the red and blue type annotations present.
+Thus we can observe that usually, only a couple of type annotations are responsible for the overhead caused by Moth's transient typed checking, whereas the rest are ``free''.
+
+The remaining benchmarks, which we have not shown in figure \ref{f:pattern}, have relatively flat column graphs and we could not identify any patterns in the typing latticse.
+The benchmark with the greatest difference in single-type configurations relative to the untyped configuration is Storage, with a configuration $7.9\%$ faster than the untyped baseline.
 However, though Storage (as shown in \ref{f:full}) appears to be bimodal like Towers, we were unable to find any relationships between its typing lattice and any individual type annotations.
@@
 
-The high-performance computing community also has been investigating how
+The high-performance computing community has been investigating how
 tools and visualization can help developers to utilize their systems
@@
 guided by runtime feedback, cost models, or heuristics. 
-The large body of work \citep{Isaacs:2014:PerfViz} uses various approaches,
+Their large body of work \citep{Isaacs:2014:PerfViz} uses various approaches,
 though, we are not aware of work that has used an approach similar to ours.
@@
 the remaining configurations. We also observed trimodal profiles, as well as performance increases when type annotations where added.
-By inspecting benchmark graphs colour coded for each individual type annotation in a program's source code, we were often able to
-identify that just one or two of the type checks were likely responsible for
-the bimodal performance, by observing patterns in the associated graph.
+By inspecting benchmark graphs colour coded for each individual type annotation in a program's source code, we were often able to notice patterns which suggested
+that just one or two of the type annotations were likely responsible for
+the bimodal performance.
 However, this approach requires inspecting one graph per type annotation.
 
-We also investigated another approach to help identify such annotation, for an individual benchmark, we can produce a single graph of the performances of individual configurations for each type annotation.
+We also investigated another approach to help identify such annotations, for an individual benchmark, we can produce a single graph of the performances of individual configurations for each type annotation.
 In most cases both approaches identified the same type annotations as likely causing a significant effect on performance.
+In particular these type annotations seemed to be responsible for most of the overhead caused by Moth's transient type checking.
 
 This is preliminary work, in particular we have not yet identified exactly why these type annotations appear to have an effect on performance.
-
 There are also a number of threats to validity. 

### 2019-08-17 paper.tex
Removing Takikawa protocol contribution, with comment documenting reasoning
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 \end{figure*}
-We hypothesise that the tri/bi-modal graphs, shown in figure~\ref{f:full}, are caused by individual type annotations: that is, certain annotations have a significant effect on performance.
-In order to identify such type annotations, for each benchmark and type annotation we manually inspected the scatter plots for the benchmark's typing lattice, coloured based on whether the chosen type annotation was present or not.
-This resulted in 1,775 graphs, however most of these showed no discernible pattern.
+We hypothesise that the bi/tri-modal graphs shown in figure~\ref{f:full} are caused by a few, highly significant type annotations: i.e.\ there are a very few annotations that determine each benchmark's performance.
+
+In order to identify those type annotations, we generated coloured
+scatter plots for each benchmark based on whether each individual type
+annotation was present in that configuration or not. The position of
+the dots in each coloured plot are the same as Figure~\ref{f:full}, but
+they are coloured as to whether each annotation is there or not: This resulted in 1,775 scatterplots --- one for each type annotation for each benchmark.
+\kjx{if we say this, we should show a bunch for one benchmark.}
+We then manually inspected all these scatterplots,
+but unfortunately most of these showed no discernible pattern.
+\kjx{YEAH BUT SOME DID. Shouldn't we show at least one?}
+
+As an alternative, less labour intensive approach to identifying
+potentially influential type annotations, we measured one additional
+configuration for each type annotation in each benchmark, with only
+that single type annotation present. Our aim was to compare the overhead of
+each type annotation in isolation against the no-typecheck baseline.
 
@@
 
-Figure \ref{f:pattern} shows the results of these experiments, as well as the patterns we manually identified.
+% We used the same computer and number of iterations as for the typing
+% lattice experiments, albeit these were done on different days.
 
-The column graphs show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
-The y-axes for the column graphs are the same as the associated scatter graphs.
-However, though the right y-axes for the column graphs appear to be identical to the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.
+Figure \ref{f:pattern} shows the results of these experiments.
+%, as well as the patterns we manually identified.
+%%KJX but you haven't talked about manual patterns!
+
+This figure shows a pair of graphs for a selection of ten benchmarks: a scatterplot and a column graph.
+
+The column graphs show the execution times of each configuration for each individual annotation in a benchmark.
+The x-axis of the graphs indicates the index of this annotation in the order they appear in the source code.
+The first column represents the first annotation, and the last column represents the last one.
+The y-axes for the column graphs are the same as the associated scatter graphs, showing the execution time of each configuration.\footnote{Although the right y-axes for the column graphs appear to be identical to the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}
 
@@
 The identified types are those represented by the red and blue columns.
-The scatter plots are colour coded accordingly: red/blue circles represent configurations with the red/blue type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
+The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
 
 As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
-For the Snake, Towers, and CD benchmark, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
+For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
 For Permute there is actually a signficant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.
-DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this does not appear to be cumulative since the purple dots are around the same height as the red and blue ones: thus \textit{either} annotation is sufficient for the full improvement.
+DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect does not appear to be cumulative, however: the purple dots are about the same height as the red and blue ones, so \textit{either} annotation is sufficient for the full benefit.
 
@@
 
-Finally, for SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a pattern in the typing lattice.
-
-Note however that in the case of DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively); nonetheless, we found no apparent relation between these and the typing lattice.
+For SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a matching pattern in the configuration performance.
+Finally, for DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively), however we again found no apparent relation between these outliers and the overall performance
 

### 2019-08-17 paper.tex
Correct abbreviation of everyone but Isaac’s name
Signed-off-by: Stefan Marr <git@stefan-marr.de>
Edit intro to section 3, incl. sec. 3.1
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
-\author[Richard R.]{Richard Roberts}
+\author[R. Roberts]{Richard Roberts}
 \affiliation{
@@
 
-\author[Stefan M.]{Stefan Marr}
+\author[S. Marr]{Stefan Marr}
 \orcid{0000-0001-9059-5180}
@@
 
-\author[Michael H.]{Michael Homer}
+\author[M. Homer]{Michael Homer}
 \affiliation{
@@
 
-\author[James N.]{James Noble}
+\author[J. Noble]{James Noble}
 \orcid{0000-0001-9036-5692}             %% \orcid is optional
@@
         that cause significant performance effects
-  \item adapting the Takikawa protocol for languages and implementations
-        where performance effects are dominated by type annotations instead
-        of transitions between typed and untyped modules
+  % SM: after discussion with Isaac, this was done
+  %     by some Reticulated Python paper before
+  % and Isaac's adaptation is restricted to using rounding rules
+  %    (>= 0.5 is round up) that give useful results (at least 1 type check)
+  %    for selecting the number of type checks
+  % \item adapting the Takikawa protocol for languages and implementations
+  %       where performance effects are dominated by type annotations instead
+  %       of transitions between typed and untyped modules
   \item observing that performance of small benchmarks
@@
 Truffle interpreters \citep{Wurthinger:2017:PPE}.
-\RR{This last sentence seems out of place, does it belong here?}
-\kjx{yes I think it's fine}
+% SM: kjx says it's fine, so, drop for now
+% \RR{This last sentence seems out of place, does it belong here?}
+% \kjx{yes I think it's fine}
 
 \section{Experimental Methodology}
-Our methodology is built upon the so-called ``Takikawa'' or ``Takikawa-Greenman'' evaluation protocol \cite{Takikawa2016,Greenman2019jfp} benchmarks a \emph{lattice} of $2^N$ \emph{configurations} of each benchmark, where a configuration is a particular mix of static and dynamically typed code, however here we only test a relatively small sample of this lattice.
-Here we apply this methodology to our previous work \cite{roberts-and-co-ecoop-2019}, where we evaluated the performance of transient type checks in Moth by using a suite of benchmarks.
+
+Our goal is to identify which type annotations in Grace programs
+cause performance effects.
+To this end, we built upon the so-called ``Takikawa'' or ``Takikawa-Greenman'' evaluation protocol \cite{Takikawa2016,Greenman2019jfp}.
+It uses $2^N$ \emph{configurations} of each benchmark.
+A configuration is a particular mix of static and dynamically typed code, forming a lattice of configurations.
+We only test a relatively small sample of this lattice,
+which is in our experience sufficient to pinpoint performance anomalies caused
+by type annotations.
 
@@
 
-To investigate Moth's support for transient type checks further, we carried out a similar approach to the Takikawa-Vitousek protocol.
-For each benchmark we generated 100 partially typed versions (fewer if the benchmark was too small).
+ % Here we apply this methodology to our previous work \cite{roberts-and-co-ecoop-2019}, where we evaluated the performance of transient type checks in Moth by using a suite of benchmarks.
+
+To identify type annotations causing anomalies,
+% To investigate Moth's support for transient type checks further, 
+we adapted the Takikawa-Vitousek protocol. For each benchmark we generated 100 partially typed versions,
+or fewer if the benchmark has at most 10 types.
 We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations.
@@
 \subsection{The Benchmarks}
-In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the performance of transient type checks using 21 benchmarks. We started with the 14 benchmarks of the \AWFY
+In our previous work \cite{roberts-and-co-ecoop-2019}, we evaluated the performance of transient type checks using 21 benchmarks. We started with the 14 benchmarks of the \AWFY
 benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.

### 2019-08-17 paper.tex
Benchmark section needs revision, need input on this though.
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.
-We added another nine benchmarks taken from the gradual-typing literature.
+We added another \ugh{nine}\sm{14 plus 9 isn't 21} benchmarks taken from the gradual-typing literature.\sm{it's also unclear whether this is about previous work or what we do now.}
 Our goal was to complement the \AWFY with additional ones that are used for similar experiments and can be ported to Grace.
@@
 Table~\ref{tab:gradual-benchmarks}.
+\sm{this paragraph needs to be cleaned up.
+We need to say what we use now. Clearly, and not intermingled with previous work.
+And then, afterwards we can say whether this is the same we did before. or how it differs.
+}
 

### 2019-08-17 paper.tex
Hopefully recovered from merge hell...

--- 2019-08-17 paper.tex
@@
 \begin{document}
-\title{How Slow do my Transient Typechecks Go?}
+\title{How Slow do my Transient Type checks Go?}
 
@@
 On the other end of the spectrum are behavioural
-typechecks (as in Typed Racket \cite{typedScheme08,takikawa2012},
+type checks (as in Typed Racket \cite{typedScheme08,takikawa2012},
 Gradualtalk \cite{gradualtalk14}),
@@
 \begin{itemize}
-  \item an approach to identify gradual type annotations
+  \item an approach to identifying gradual type annotations
         that cause significant performance effects
@@
   %       of transitions between typed and untyped modules
-  \item observing that performance of small benchmarks
-        with 10--250 type annotations is dominated by only
-        one or two type annotations
+  \item an observation that the overhead of Moth's transient type checking on small benchmarks (with 10--250 type annotations) is usually caused by only one or two type annotations
 \end{itemize}
@@
 we adapted the Takikawa-Vitousek protocol. For each benchmark we generated 100 partially typed versions,
-or fewer if the benchmark has at most 10 types.
+or fewer if the benchmark has less than 11 types.
 We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations.
@@
 
-As an alternative, and less labour intensive approach to identifying potentially influential type annotations, for each type annotation we ran each benchmark with \textit{only} that type annotation present.
-We used the same computer and number of iterations as for the typing lattice experiments, albeit these were done on different days.
+Figure \ref{f:pattern} shows the results of these experiments, together with patterns we manually identified.
 
-% We used the same computer and number of iterations as for the typing
-% lattice experiments, albeit these were done on different days.
-
-Figure \ref{f:pattern} shows the results of these experiments.
-%, as well as the patterns we manually identified.
-%%KJX but you haven't talked about manual patterns!
+The column graphs show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
+The y-axes for the column graphs are the same as the associated scatter graphs.
+However, though the right y-axes for the column graphs appear to be identical to the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.
 

### 2019-08-17 paper.tex
Revise The Benchmarks section. Cut it down to what’s needed for the paper
Signed-off-by: Stefan Marr <git@stefan-marr.de>

--- 2019-08-17 paper.tex
@@
 \begin{document}
-\title{How Slow do my Transient Type checks Go?}
+\title{How Slow do my Transient Type Checks Go?}
 
@@
 \subsection{The Benchmarks}
-In our previous work \cite{roberts-and-co-ecoop-2019}, we evaluated the performance of transient type checks using 21 benchmarks. We started with the 14 benchmarks of the \AWFY
-benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.
-We added another nine benchmarks taken from the gradual-typing literature.
-Our goal was to complement the \AWFY with additional ones that are used for similar experiments and can be ported to Grace.
-We surveyed a number of papers
-\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
-and selected additional benchmarks that have been used by multiple papers.
-Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
-or were available in different versions.
-While not always behaviourally equivalent,
-we chose the \AWFY versions since we already used them to
-establish a performance baseline.
-The selected benchmarks as well as the papers in which they were used are shown in
-Table~\ref{tab:gradual-benchmarks}.
 
-\begin{table}[htb]
-  \caption{Benchmarks selected from literature.}
-  \label{tab:gradual-benchmarks}
-  \begin{center}
-    \begin{tabular}{l l r}
-      Fannkuch & \cite{Vitousek2017,Greenman2018} \\
-      Float & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
-      Go & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
-      NBody & \cite{Kuhlenschmidt:2018:preprint,Vitousek2017,Greenman2018} & used \cite{Marr2016} \\
-      Queens & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} & used \cite{Marr2016} \\
-      PyStone & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
-      Sieve & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} & used \cite{Marr2016} \\
-      Snake & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} \\
-      SpectralNorm & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
-    \end{tabular}
-  \end{center}
-\end{table}
+For this work, we rely on the benchmark suite compiled for previous work
+\cite{roberts-and-co-ecoop-2019}.
+It is a collection of 21 benchmarks in total,
+derived from the \AWFY
+benchmark suite \cite{Marr2016} and other benchmarks
+from the gradual-typing literature.
 
-The benchmarks were modified to have complete type information.  To
-ensure correctness and completeness of these experiments, we ensured that each benchmark is completely typed.  To assess the performance
-overhead of type checking, we compared the execution of Moth with all
-checks disabled against an execution that has all checks enabled.  The
-results of this evaluation demonstrated that Moth's transient type checking has an overhead of 5\% (min.\ -13\%,
-max. 79\%) compared to Moth's peak performance with type checking disabled.
+% In our previous work , we evaluated the performance of transient type checks using 21 benchmarks.
+We started with the 14 benchmarks of the : Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.
+We added another \ugh{nine}\sm{14 plus 9 isn't 21} benchmarks taken from the gradual-typing literature.\sm{it's also unclear whether this is about previous work or what we do now.}
+% Our goal was to complement the \AWFY with additional ones that are used for similar experiments and can be ported to Grace.
+% We surveyed a number of papers
+% \citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
+% and selected additional benchmarks that have been used by multiple papers.
+% Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
+% or were available in different versions.
+% While not always behaviourally equivalent,
+% we chose the \AWFY versions since we already used them to
+% establish a performance baseline.
+% The selected benchmarks as well as the papers in which they were used are shown in
+% Table~\ref{tab:gradual-benchmarks}.
+% \sm{this paragraph needs to be cleaned up.
+% We need to say what we use now. Clearly, and not intermingled with previous work.
+% And then, afterwards we can say whether this is the same we did before. or how it differs.
+% }
+%
+% \begin{table}[htb]
+%   \caption{Benchmarks selected from literature.}
+%   \label{tab:gradual-benchmarks}
+%   \begin{center}
+%     \begin{tabular}{l l r}
+%       Fannkuch & \cite{Vitousek2017,Greenman2018} \\
+%       Float & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
+%       Go & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
+%       NBody & \cite{Kuhlenschmidt:2018:preprint,Vitousek2017,Greenman2018} & used \cite{Marr2016} \\
+%       Queens & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} & used \cite{Marr2016} \\
+%       PyStone & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
+%       Sieve & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} & used \cite{Marr2016} \\
+%       Snake & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} \\
+%       SpectralNorm & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
+%     \end{tabular}
+%   \end{center}
+% \end{table}
 
-We did not measure programs that mixed typed and untyped code because
-with our implementation technique we expected fully typed programs to
-have the largest overhead, which is not the case for other kinds of
-gradual type checking.
+Each benchmark has a complete set of type annotations,
+\ie, all possible type annotations are present,
+which is verified in the parser.
+
+In our previous work, we \cite{roberts-and-co-ecoop-2019}
+we determined that the overhead of type checking on Moth is on average
+of 5\% (min.\ -13\%, max. 79\%).
+This compares the peak performance of Moth with all
+checks disabled against an execution that has all checks enabled. 
+%  The
+% results of this evaluation demonstrated that Moth's transient type checking has an overhead  compared to Moth's peak performance with type checking disabled.
+%
+% We did not measure programs that mixed typed and untyped code because
+% with our implementation technique we expected fully typed programs to
+% have the largest overhead, which is not the case for other kinds of
+% gradual type checking.
 
@@
 they are coloured as to whether each annotation is there or not: This resulted in 1,775 scatterplots --- one for each type annotation for each benchmark.
-\kjx{if we say this, we should show a bunch for one benchmark.}
-We then manually inspected all these scatterplots,
-but unfortunately most of these showed no discernible pattern.
-\kjx{YEAH BUT SOME DID. Shouldn't we show at least one?}
+We then manually inspected all these scatterplots, though almost all of these graphs showed no discernible pattern, we were able to identify some for about a third of the benchmarks.
 
@@
 though, we are not aware of work that has used an approach similar to ours.
-One may argue, this is largely a question of practicality.\sm{
-i put this in to hint at issues with usability.}
-\kjx{Hmm not sure what the last sentence here means!}
+One may argue, this is largely a question of practicality.
 
@@
 whether this approach can assist with optimsations for programmers' day-to-day  development, or help VM engineers identifying performance bugs in the underlying virtual machines.
-\sm{is our approach practical. Is it useful for day-to-day development?
-What's its context, use, general application?}
-\kjx{filled out the last sentence}
 

### 2019-08-17 paper.tex
intro of sec. 4
Signed-off-by: Stefan Marr <git@stefan-marr.de>

--- 2019-08-17 paper.tex
@@
 
-\section{Results}
+\section{Performance of Benchmark Configurations}
 \label{s-overall}
@@
 \end{figure*}
-The results for the performance of our sample of typing lattice configurations are presented in figure~\ref{f:full}.
-Following \cite{vitousek-transient-arXive-2019}, the points on each graph show the average execution of each individual configuration.
+
+Before we start to investigate specific type annotations,
+we present the performance measurements of our sample of the typing lattice configurations in figure~\ref{f:full}.
+These results are the foundation for a more details analysis.
+
+Following \cite{vitousek-transient-arXive-2019}, the points on each graph in figure~\ref{f:full} show the average execution of each individual configuration.
 The x-axis represents the proportion of type annotations for each configuration, with the left- and right-most points showing the times for the fully untyped and typed configurations respectively.

### 2019-08-17 paper.tex
Say we start from Takikawa, and are inspired by T-V
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 % To investigate Moth's support for transient type checks further, 
-we adapted the Takikawa-Vitousek protocol. For each benchmark we generated 100 partially typed versions,
+we adapted the Takikawa protocol
+and take inspiration from the Takikawa-Vitousek variant.
+For each benchmark, we generated 100 partially typed versions,
 or fewer if the benchmark has less than 11 types.

### 2019-08-17 paper.tex
Fixed english

@@
 we adapted the Takikawa protocol
-and take inspiration from the Takikawa-Vitousek variant.
+and took inspiration from the Takikawa-Vitousek variant.
 For each benchmark, we generated 100 partially typed versions,
@@
 we present the performance measurements of our sample of the typing lattice configurations in figure~\ref{f:full}.
-These results are the foundation for a more details analysis.
+These results are the foundation for a more detailed analysis.
 

### 2019-08-17 paper.tex
Revise sec. 5
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
-\section{Individual Annotations}
+\section{Identifying Type Annotations With Signification Performance Impact}
 \label{s-individual}
@@
 \end{figure*}
-We hypothesise that the bi/tri-modal graphs shown in figure~\ref{f:full} are caused by a few, highly significant type annotations: i.e.\ there are a very few annotations that determine each benchmark's performance.
 
-In order to identify those type annotations, we generated coloured
-scatter plots for each benchmark based on whether each individual type
-annotation was present in that configuration or not. The position of
-the dots in each coloured plot are the same as Figure~\ref{f:full}, but
-they are coloured as to whether each annotation is there or not: This resulted in 1,775 scatterplots --- one for each type annotation for each benchmark.
+We hypothesise that the previously identified bi/tri-modal performance behaviour
+as seen in the graphs for Permute and others (cf. figure~\ref{f:full}) are caused by only a few type annotations: i.e.\ there are a very few annotations that determine each benchmark's performance.
+
+% In order to identify those type annotations, we generated coloured
+% scatter plots for each benchmark based on whether each individual type
+% annotation was present in that configuration or not. The position of
+% the dots in each coloured plot are the same as Figure~\ref{f:full}, but
+% they are coloured as to whether each annotation is there or not: This resulted in 1,775 scatterplots --- one for each type annotation for each benchmark.
 We then manually inspected all these scatterplots, though almost all of these graphs showed no discernible pattern, we were able to identify some for about a third of the benchmarks.
 
-As an alternative, less labour intensive approach to identifying
-potentially influential type annotations, we measured one additional
-configuration for each type annotation in each benchmark, with only
-that single type annotation present. Our aim was to compare the overhead of
+To verify this hypothesis 
+% As an alternative, less labour intensive approach to identifying
+% potentially influential type annotations,
+we measured for each type annotation one additional
+configuration, with only that single type annotation present.
+Our aim was to compare the overhead of
 each type annotation in isolation against the no-typecheck baseline.
 
-Figure \ref{f:pattern} shows the results of these experiments, together with patterns we manually identified.
+Figure \ref{f:pattern} shows the results of these experiments.
+% , together with patterns we manually identified.
+It shows a pair of graphs for a selection of ten benchmarks: a scatterplot and a column graph.
 
-The column graphs show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
+The column graphs, right of the corresponding scatter plot, show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
 The y-axes for the column graphs are the same as the associated scatter graphs.
@@
 
-This figure shows a pair of graphs for a selection of ten benchmarks: a scatterplot and a column graph.
-
-The column graphs show the execution times of each configuration for each individual annotation in a benchmark.
+% SM: already said before
+% The column graphs show the execution times of each configuration for each individual annotation in a benchmark.
 The x-axis of the graphs indicates the index of this annotation in the order they appear in the source code.
 The first column represents the first annotation, and the last column represents the last one.
-The y-axes for the column graphs are the same as the associated scatter graphs, showing the execution time of each configuration.\footnote{Although the right y-axes for the column graphs appear to be identical to the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}
+The y-axes for the column graphs are the same as the associated scatter graphs, showing the execution time of each configuration.
+% \footnote{Although the right y-axes for the column graphs appear to be identical to the scatter plot axes, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.}
 
@@
 The identified types are those represented by the red and blue columns.
-The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
+The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation present, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
 
+\here
 As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
 For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
-For Permute there is actually a signficant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.
-DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect does not appear to be cumulative, however: the purple dots are about the same height as the red and blue ones, so \textit{either} annotation is sufficient for the full benefit.
+For Permute there is actually a significant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.
+DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect does not appear to be cumulative, however: the purple dots are about the same height as the red and blue ones, so \emph{either} annotation is sufficient for the full benefit.
 
@@
 Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotation is present, but not the blue one.
-We had previously identified this benchmark and the afformentioned type annotations \cite{roberts-and-co-ecoop-2019} as being the worst case for Moth.
+We had previously identified this benchmark and the aforementioned type annotations \cite{roberts-and-co-ecoop-2019} as being the worst case for Moth.
 

### 2019-08-17 paper.tex
Remove marker
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
-\here
 As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
-For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
+For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing configurations.
 For Permute there is actually a significant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.

### 2019-08-17 paper.tex
Reorder related work, to flow better
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
-Optimization Coaching
-\citep{St-Amour:2012:OCO} is a related approach more
-directly integrated into the language runtime and development environment.
-Optimizer Coaching uses feedback from the runtime to guide developers
+The high-performance computing community has been investigating how
+tools and visualisations can help developers to utilise their systems
+more efficiently \citep{Papenhausen:2016:IVT,daSilva:2019:PSV}.
+Their focus is typically on parallelisation opportunities,
+guided by runtime feedback, cost models, or heuristics. 
+Their large body of work \citep{Isaacs:2014:PerfViz} uses various approaches,
+though, we are not aware of work that has used an approach similar to ours.
+
+At the moment, our approach to identifing type annotations that case
+performance anomalies is not integrated into a development environment.
+Though, for instance Optimization Coaching
+\citep{St-Amour:2012:OCO} is a promising direction.
+Optimization Coaching uses feedback from the runtime to guide developers
 to insert or change type declarations 
@@
 we need to run full experiments to get the necessary information.
-Furthermore, we still lack the seamless integration in a development environment.
-
-The high-performance computing community has been investigating how
-tools and visualization can help developers to utilize their systems
-more efficiently \citep{Papenhausen:2016:IVT,daSilva:2019:PSV}.
-Their focus is typically on parallelisation opportunities,
-guided by runtime feedback, cost models, or heuristics. 
-Their large body of work \citep{Isaacs:2014:PerfViz} uses various approaches,
-though, we are not aware of work that has used an approach similar to ours.
-One may argue, this is largely a question of practicality.
 

### 2019-08-17 paper.tex
Aded a sentence about not finding other patterns

--- 2019-08-17 paper.tex
@@
 % potentially influential type annotations,
-we measured for each type annotation one additional
+for each type annotation we measured one additional
 configuration, with only that single type annotation present.
-Our aim was to compare the overhead of
+We did this to compare the overhead of
 each type annotation in isolation against the no-typecheck baseline.
@@
 % , together with patterns we manually identified.
-It shows a pair of graphs for a selection of ten benchmarks: a scatterplot and a column graph.
+It shows a pair of graphs for a selection of ten benchmarks: it's associated typing lattice scatter plot and a column graph showing the results of these single type annotation experiments.
 
-The column graphs, right of the corresponding scatter plot, show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
+The column graphs, right of the corresponding scatter plot, shows the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
 The y-axes for the column graphs are the same as the associated scatter graphs.
@@
 
-For each benchmark we highlighted one or two type annotations that appear to show a pattern for the typing lattice performance, or in the case of SpectralNorm and Json had higher than usual columns.
+For each benchmark we highlighted one or two type annotations that appear to show a pattern for the typing lattice performance scatter plots, or in the case of SpectralNorm and Json had higher than usual columns.
 The identified types are those represented by the red and blue columns.
 The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation present, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
+Though we exhaustively inspected such colour-coded scatter plot for all 1,775 type annotations across all 21 benchmarks, the only patterns we noticed are those shown in \ref{f:pattern}.
 
-\here
-As can be seen, most of the time the patterns we identified in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
+As can be seen, most of the time the patterns we found in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
 For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.

### 2019-08-17 paper.tex
Fixed conclusion

@@
 The scatter plots are colour-coded accordingly: a red or blue circle represents configurations with the given type annotation present, but not both, purple circles represent configurations with both type annotations present, and grey circles represent those with neither.
-Though we exhaustively inspected such colour-coded scatter plot for all 1,775 type annotations across all 21 benchmarks, the only patterns we noticed are those shown in \ref{f:pattern}.
+Though we exhaustively inspected such colour coded scatter plots for all 1,775 type annotations across all 21 benchmarks, the only patterns we noticed are those shown in \ref{f:pattern}.
 
@@
 the remaining configurations. We also observed trimodal profiles, as well as performance increases when type annotations where added.
-By inspecting benchmark graphs colour coded for each individual type annotation in a program's source code, we were often able to notice patterns which suggested
-that just one or two of the type annotations were likely responsible for
-the bimodal performance.
-However, this approach requires inspecting one graph per type annotation.
 
-We also investigated another approach to help identify such annotations, for an individual benchmark, we can produce a single graph of the performances of individual configurations for each type annotation.
-In most cases both approaches identified the same type annotations as likely causing a significant effect on performance.
-In particular these type annotations seemed to be responsible for most of the overhead caused by Moth's transient type checking.
+By inspecting graphs of the performances of individual configurations for each type annotation, we can see type annotations that likely have a significant effect on performance.
+By then inspecting the typing lattice graphs colour coded for such type annotations, we were often able to notice patterns which suggested that just one or two of the type annotations were likely responsible for the bimodal performance.
+Though every type annotation that appeared to have a significant effect across the typing lattice also showed a singificant performance effect when no other typing annotations where present, the converse did not hold.
+In particular, the type annotations we identified appear to be responsible for most of the overhead caused by Moth's transient type checking.
 

### 2019-08-17 paper.tex
Reworded Fixed conclusion

@@
 
-By inspecting graphs of the performances of individual configurations for each type annotation, we can see type annotations that likely have a significant effect on performance.
-By then inspecting the typing lattice graphs colour coded for such type annotations, we were often able to notice patterns which suggested that just one or two of the type annotations were likely responsible for the bimodal performance.
-Though every type annotation that appeared to have a significant effect across the typing lattice also showed a singificant performance effect when no other typing annotations where present, the converse did not hold.
-In particular, the type annotations we identified appear to be responsible for most of the overhead caused by Moth's transient type checking.
+By inspecting graphs of the performance of where only one type annotation is present, we can easily identify type annotations that likely have a significant effect on performance.
+By then inspecting the typing lattice graphs colour coded based on whether such type annotation was present or not, we were often able to notice patterns.
+In particular, these patterns suggested the just one or type type annotations are likely responsible for the bimodal performance and most of the overhead caused by Moth's transient type checking.
+Though every type annotation that appeared to have a significant effect across the typing lattice also showed a significant performance effect when no other typing annotations where present, the converse did not hold.
 

### 2019-08-17 paper.tex
Fix type type to two type
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 By then inspecting the typing lattice graphs colour coded based on whether such type annotation was present or not, we were often able to notice patterns.
-In particular, these patterns suggested the just one or type type annotations are likely responsible for the bimodal performance and most of the overhead caused by Moth's transient type checking.
+In particular, these patterns suggested the just one or two type annotations are likely responsible for the bimodal performance and most of the overhead caused by Moth's transient type checking.
 Though every type annotation that appeared to have a significant effect across the typing lattice also showed a significant performance effect when no other typing annotations where present, the converse did not hold.

### 2019-09-10 paper.tex
Changed Title
Updated to new data

@@
-\documentclass[sigplan,10pt,review,screen]{acmart}\settopmatter{printfolios=true}
+\documentclass[sigplan,10pt,screen]{acmart}\settopmatter{printfolios=true}
 
-\acmConference[VMIL'19]{ACM SIGPLAN International Workshop on Virtual Machines and Intermediate Languages}{October 22, 2019}{Athens, Greece}
+\setcopyright{acmlicensed}
+\acmPrice{}
+\acmDOI{10.1145/3358504.3361232}
 \acmYear{2019}
-\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
-\acmDOI{10.1145/nnnnnnn.nnnnnnn}
-\startPage{1}
+\copyrightyear{2019}
+\acmISBN{978-1-4503-6987-9/19/10}
+\acmConference[VMIL '19]{Proceedings of the 11th ACM SIGPLAN International Workshop on Virtual Machines and Intermediate Languages}{October 22, 2019}{Athens, Greece}
+\acmBooktitle{Proceedings of the 11th ACM SIGPLAN International Workshop on Virtual Machines and Intermediate Languages (VMIL '19), October 22, 2019, Athens, Greece}
 
@@
 \collabAuthor{kjx}{orange}{James}
+\collabAuthor{isaac}{blue}{Isaac}
 \collabAuthor{mwh}{purple}{Michael}
@@
 \begin{document}
-\title{How Slow do my Transient Type Checks Go?}
+\title{Which of my Transient Type Checks are not (Almost) Free?}
 
@@
 The next section discusses dynamic type checks and gradual typing in
-Moth (an implementation of the Grace language), then section~\ref{s-eval} describes our benchmarking protocol. Section~\ref{s-overall} then presents the overall results of
+Moth (an implementation of the Grace language), then section~\ref{s-meth} describes our benchmarking protocol. Section~\ref{s-overall} then presents the overall results of
 our benchmarks, while section~\ref{s-individual}
@@
 \section{Experimental Methodology}
+\label{s-meth}
 
@@
 Racket, where static vs dynamic typing is set per-module, so $N$ is
-the number of modules. Grace, following other languages such as
+the number of modules.
+The original Takikawa protocol also suggested a sampling strategy, where $10N$ configurations are randomly chosen from the lattice, however the lattice is a binomial distribution, meaning the majority of chosen benchmarks will have around $N/2$ type annotations
+Grace, following other languages such as
 Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
@@
 each interval by erasing type annotations.
+However, this approach was designed for benchmarks with large numbers of type annotations, as well as for a larger sample size than our work.
 
@@
 or fewer if the benchmark has less than 11 types.
-We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations.
-We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to choose the type annotations each configuration contained randomly, and we ensured that no duplicate configurations were generated.
+We did an even split so that for each $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations.
+We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to randomly choose the type annotations each configuration contained, and we ensured that no duplicate configurations were generated.
 In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark (or 97 in the case of our Storage benchmark, since it only has 10 type annotations).
@@
 \ie, all possible type annotations are present,
-which is verified in the parser.
+which is verified in the parser. \isaac{This is misleading, in a few cases, some of the type annotations are in fact "Unknown", which is exactly what my "removing" of type annotations places}
 
@@
 
-Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained (over the mean time after warmup) for any of experiments was $\pm3.36\%$ (for the PyStone benchmark).
+Though outliers remain visible in the plots for each individual benchmark, the largest 95\% confidence interval we obtained (over the mean time after warmup) for any of experiments was $\pm8.3\%$ (for the PyStone benchmark).
 
-All experiments were executed on a Windows 10 machine
-running Fedora 30 within Windows Subsystems for Linux (version 1).
-The machine has an Intel Core i7-6800K @3.40GHz, with 6 cores, for a total of 12 hyperthreads.
-This computer appears to produce comparable results to the one used in our previous work, although the absolute performance is higher, the relative overhead of transient type-checks is similar.
-
-We used the same version of Moth as previously, together with ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
+We used the same machine, version of Moth and Graal as previously, specifically all experiments were executed on a machine running Ubuntu Linux 16.04.4, with Kernel 3.13.
+The machine has two Intel Xeon E5-2620 v3 2.40GHz, with 6 cores each, for a total of 24 hyperthreads.
+We also used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
 Benchmarks were executed one by one to avoid interference between them.
@@
 Most of the graphs are essentially horizontal lines, indicating that the overhead of including type annotations is negligible.
-The plots for CD and Richards show a roughly linear increase, however, i.e.\ for these two benchmarks, adding type annotations reduces performance linearly.
+The plots for CD and Richards show a roughly linear increase, i.e.\ for these two benchmarks, adding type annotations reduces performance linearly.
 On the other hand, the plots for Go, Permute, DeltaBlue, and Storage show \emph{decreases}: i.e.
@@
 %
-However, the Richards and Snake benchmarks \textit{do} have
+However, the Richards benchmark \textit{does} have
 sections outside these bounds, and isolated executions of
-a few others (SpectralNorm, JSON, DeltaBlue) are also outliers.
+a couple of others (Fannkuch and DeltaBlue) are also outliers.
 %
 We believe that the 0\% and 100\% bounds are nonetheless a
-reasonable heuristic estimator in most cases, and will
+reasonable heuristic estimator in most cases, and we will
 examine further the outlying benchmarks.
 
-Of particular note is that some of the graphs (Permute, Storage, Snake, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly-horizontal lines.
+Of particular note is that some of the graphs (Permute, Storage, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly-horizontal lines.
 Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away.
-List appears to have three performance modes: most of the configurations up to about 60\% annotations perform at roughly the same speed as untyped code; most configurations above 70\% perform 1.8 times more slowly than untyped code; and then across all annotation percentages there are configurations that perform roughly 1.5 times slower than the untyped baseline.
+List appears to have three performance modes: the graph consists of three mostly flat lines, at about 1 times slower, 1.5 times slower, and 1.8 times slower than untyped code.
 
@@
 
-For SpectralNorm and Json, although we found outliers in the column graphs (types 27 and 83, highlighted in red), we were unable to find a matching pattern in the configuration performance.
-Finally, for DeltaBlue and Richards, there are two noticeable spikes in the column graphs (at \#62 and \#24, respectively), however we again found no apparent relation between these outliers and the overall performance
+For Json and PyStone, although we found outliers in the column graphs (such as types \#112 and \#16, highlighted in red), we were unable to find a matching pattern in the configuration performance.
+Finally, for DeltaBlue there is a noticeable spike in the column graphs (at \#198), however we again found no apparent relation between this outlier and the overall performance.
 
-In particular, observe that the grey dots for Snake, Towers, Permute, DeltaBlue, Go, and List are mostly flat horizontal lines, this indicates that by simply deleting the red and blue type annotations, the performance impact of transient type checking becomes negligible.
+In particular, observe that the grey dots for Snake, Towers, DeltaBlue, Go, and List are mostly flat horizontal lines, this indicates that by simply deleting the red and blue type annotations, the performance impact of transient type checking becomes negligible.
 However for CD and Richards, the transient type checking overhead of the grey dots appears linear, albeit still less than with the red and blue type annotations present.
@@
 
-The remaining benchmarks, which we have not shown in figure \ref{f:pattern}, have relatively flat column graphs and we could not identify any patterns in the typing latticse.
-The benchmark with the greatest difference in single-type configurations relative to the untyped configuration is Storage, with a configuration $7.9\%$ faster than the untyped baseline.
-However, though Storage (as shown in \ref{f:full}) appears to be bimodal like Towers, we were unable to find any relationships between its typing lattice and any individual type annotations.
+The remaining benchmarks, which we have not shown in figure \ref{f:pattern}, have even flatter column graphs than PyStone, and we could not identify any patterns in the typing lattices.
+Of those we did not show, the greatest difference in single-type configurations relative to the untyped configuration was $5.07\%$ (for the Float benchmark).
 %For example for Snake, the dots covered in red (corresponding to configurations with type annotation number 61) are noticeably higher than the grey ones (those without type annotation 61), thus it appears that the 61\textsuperscript{st} type annotation causes a noticeable slowdown, which is also exemplified by the higher than average red column.

### 2019-09-10 graphs/pattern.py

@@ -2,11 +2,11 @@
 from results import *
 
 TODO = {
-	"Snake-70": 61, "Towers-30": 7,
-	"CD-250": 107, "Permute-14": 8,
+	"Snake-70": 61, "CD-250": 107, 
+        "Towers-30": 7, "Permute-14": 8,
 	"DeltaBlue-243": [154, 155], "Go-230": [31, 47],
-	"List-23": [17, 18], "Richards-177": [172, 173], #24,
-	"SpectralNorm-39": 27, "Json-134": 83}
+	"List-23": [17, 18], "Richards-177": [172, 173],
+        "Json-134": 112, "PyStone-85": 16}
 
 res = []
 i = 0

### 2019-09-10 paper.tex
Update CCS and keywords

@@
-\documentclass[sigplan,10pt,screen]{acmart}\settopmatter{printfolios=true}
+\documentclass[sigplan,screen]{acmart}
 
@@
 %% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
-\begin{CCSXML}
-<ccs2012>
-<concept>
-<concept_id>10011007.10011006.10011041.10011044</concept_id>
-<concept_desc>Software and its engineering~Just-in-time compilers</concept_desc>
-<concept_significance>500</concept_significance>
-</concept>
-<concept>
-<concept_id>10011007.10011006.10011008.10011009.10011011</concept_id>
-<concept_desc>Software and its engineering~Object oriented languages</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-<concept>
-<concept_id>10011007.10011006.10011041.10010943</concept_id>
-<concept_desc>Software and its engineering~Interpreters</concept_desc>
-<concept_significance>300</concept_significance>
-</concept>
-</ccs2012>
-\end{CCSXML}
-
-\ccsdesc[500]{Software and its engineering~Just-in-time compilers}
+\ccsdesc[500]{Software and its engineering~Software performance}
 \ccsdesc[300]{Software and its engineering~Object oriented languages}
-\ccsdesc[300]{Software and its engineering~Interpreters}
+\ccsdesc[300]{Software and its engineering~Just-in-time compilers}
 
 \keywords{dynamic type checking, gradual types, optional types, Grace,
-object-oriented programming}
+performance evaluation, benchmarking, object-oriented programming}
 \maketitle
@@
 \bibliography{references}
-
 \end{document}

### 2019-09-11 paper.tex
More changes

@@
 \setcopyright{acmlicensed}
-\acmPrice{}
+\acmPrice{15.00}
 \acmDOI{10.1145/3358504.3361232}
@@
 \section{Introduction}
+Gradual typing aims to add static type annotations to dynamic languages, increasing their safety while maintaining flexibility \citep{GiladPluggable2004,Siek2006,XXXSiek2015}, and/or, permitting dynamic type annotations within static languages, increasing flexibility whilst maintaining some safety \citep{AbadiTOPLAS1991}.
 
-Gradual typing aims to add static type annotations to dynamic languages, increasing
-their safety while maintaining flexibility
-\citep{GiladPluggable2004,Siek2006,XXXSiek2015}
---- or, complementarily,
-to permit dynamic type annotations within static languages, increasing
-flexibility whilst maintaining safety
-\citep{AbadiTOPLAS1991}.
+There is a vast spectrum of different approaches to gradual typing \cite{kafka18,bensurvey18icfp}.
+Here we measure the performance of ``transient'' or ``type-tag'' checks (as in Reticulated Python), which offer first-order semantics: they check that an object's type constructor or names of supported methods match any static types that the object flows through, but not the return/argument types of such methods \cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
 
-There is a spectrum of different approaches to gradual typing
-\cite{kafka18,bensurvey18icfp}.
-At one end is ``pluggable types'' (as in Strongtalk \cite{strongtalk}) or ``erasure
-semantics'' (as in 
-%% Two important approaches are optional\citep{}
-%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
-%% These are applied to dynamic languages to reap the benefits of typing, but
-%% unfortunately also have limitations.
-%With optional or pluggable approaches such as 
-%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
-%2018}{https://www.typescriptlang.org/}
-TypeScript
-\citep{typeScriptECOOP}), where
-%
-all types are erased before execution, limiting the benefit of
-types to statically typed portions of programs, while not providing 
-any for of run time checking.  In the middle there are
-``transient'' or ``type-tag'' checks (as in Reticulated Python),
-% and ``concrete'' checks as in Thorn
-which offer first-order semantics, by checking
-that an object's type constructor or supported methods match
-% any explicit type annotations
-any static types
-that the object flows through.
-%%%WRONG kjx
-%, and checking only when data flows through those declarations
-\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
-Reticulated Python also supports alternative ``monotonic'' semantics
-which mutate objects by narrowing their concrete types when passed
-into contexts with more specific types.
-On the other end of the spectrum are behavioural
-type checks (as in Typed Racket \cite{typedScheme08,takikawa2012},
-Gradualtalk \cite{gradualtalk14}),
-and ``proxies'' (provided by Reticulated Python), which
-support higher-order semantics, retaining
-types until run 
-time, and performing checks eagerly, thus detailed information
-about type violations can be given as soon as possible via ``blame''
-tracking \cite{blame2009,blameForAll2011}.
-Finally, there is also ductile typing,
-which dynamically interprets a static type system at run time
-\cite{Ductile2011}.
-%
-%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
-%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
-%  after concrete and before behavioural. Does it matter?}
-%
-% This said "any", but the ecoop paper specifically showed that 
-Unfortunately, most gradual systems with run-time semantics
-(as opposed to type erasure as in TypeScript) typically
-imposes significant run-time performance overheads to provide these
-semantics.  This has lead to a significant body of research,
-on one hand developing techniques to optimise gradual typing 
-\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
-and on the other, developing techniques to evaluate the
-resulting performance  \cite{Takikawa2016,Greenman2019jfp}.
+Unfortunately most gradual systems with run-time semantics (as opposed to type erasure as in TypeScript) typically impose significant run-time performance overheads.
+This has lead to a significant body of research to develop techniques to optimise gradual typing \citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}.
 
-This paper builds upon our recent work on optimising transient
-type checks \cite{Roberts2017,roberts-and-co-ecoop-2019} by
-investigating how benchmark results can be repurposed to identify which
-particular transient type checks are resistant to optimisation.
+This has also lead to a technique to evaluate the performance of gradual typing, such as the so called so-called ``Takikawa'' protocol \cite{Takikawa2016,Greenman2019jfp,vitousek-transient-arXive-2019}.
+The Takikawa protocol was created to measure the runtime cost of gradual typing by testing various configurations of typed/untyped code.
+This approach was designed to investigate the the performance effect of the amount of typed vs untyped code, in particular it shows that how simply adding more type annotations does not always produce a uniform effect on performance.
+Here we adapt their approach in order to identify individual type annotations that may be responsible for performance effects.
 
-Specifically, the contributions of this work are:
-
+This paper builds upon our recent work on optimising transient type checks \cite{Roberts2017,roberts-and-co-ecoop-2019} to make the following contributions:
 \begin{itemize}
-  \item an approach to identifying gradual type annotations
+  \item an approach to identifying individual gradual type annotations
         that cause significant performance effects
@@
 
-To identify type annotations causing anomalies,
+Unlike such prior work, we wish to identify which type annotations cause anomalies,
 % To investigate Moth's support for transient type checks further, 
-we adapted the Takikawa protocol
+thus we adapted the Takikawa protocol
 and took inspiration from the Takikawa-Vitousek variant.
@@
 % In our previous work , we evaluated the performance of transient type checks using 21 benchmarks.
-We started with the 14 benchmarks of the : Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.
+We started with %the 14 benchmarks of the : Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks.
 We added another \ugh{nine}\sm{14 plus 9 isn't 21} benchmarks taken from the gradual-typing literature.\sm{it's also unclear whether this is about previous work or what we do now.}
@@
 
-Each benchmark has a complete set of type annotations,
-\ie, all possible type annotations are present,
-which is verified in the parser. \isaac{This is misleading, in a few cases, some of the type annotations are in fact "Unknown", which is exactly what my "removing" of type annotations places}
+Each benchmark has an almost complete set of type annotations,
+\ie, static type annotations are present at almost all possible points.
+%ISAAC: I changed this as it was misleading, in a few cases, some of the type annotations are in fact "Unknown", which is exactly what my "removing" of type annotations places
 
@@
 
-We used the same machine, version of Moth and Graal as previously, specifically all experiments were executed on a machine running Ubuntu Linux 16.04.4, with Kernel 3.13.
-The machine has two Intel Xeon E5-2620 v3 2.40GHz, with 6 cores each, for a total of 24 hyperthreads.
-We also used ReBench 1.0 \citep{ReBench:2018}, Java 1.8.0\_191 Graal 0.43.
+All our experiments used the same machine, Graal, and Moth as previously; the machine has two Intel Xeon E5-2620 v3 2.40GHz, with 6 cores each, for a total of 24 hyperthreads.
+The machine was running Ubuntu Linux 16.04.6, with Kernel 4.4, and we used ReBench 1.0 \citep{ReBench:2018} and Java 1.8.0\_191 Graal 0.43.
 Benchmarks were executed one by one to avoid interference between them.
@@
 To enable reproductions, the scripts we used to generate and run our experiments, including the source code for all the configurations tested, are available online.\footnote{\url{https://gitlab.ecs.vuw.ac.nz/isaac/Moth-Takikawa}}
+
+In our previous work we also compared the performance of untyped code on Moth against state-of-the-art VMs: Java, Node.js using the V8 JavaScript VM, and the Higgs JavaScript VM.
+Java was the fastest of these, and on average V8 was about 1.8x slower than Java, Moth was 2.3x slower, and Higgs was 10.4x slower.
+Hence we continue to argue that ``Moth is a suitable platform to assess the impact of our approach to gradual type checking, because its performance is close enough to state-of-the-art VMs, and run-time overhead is not hidden by slow baseline performance.''
 
@@
 
-Of particular note is that some of the graphs (Permute, Storage, Towers, and Richards) appear to show bimodal performance profiles, that is, two separate roughly-horizontal lines.
+Of particular note is that some of the graphs (Permute, Storage, Towers, and Richards) show bimodal performance profiles, that is, two separate roughly-horizontal lines.
 Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away.
-List appears to have three performance modes: the graph consists of three mostly flat lines, at about 1 times slower, 1.5 times slower, and 1.8 times slower than untyped code.
+List shows three performance modes: the graph consists of three mostly flat lines, at about 1 times slower, 1.5 times slower, and 1.8 times slower than untyped code.
 
@@
 The y-axes for the column graphs are the same as the associated scatter graphs.
-However, though the right y-axes for the column graphs appear to be identical to the scatter plot ones, they are relative to performance results collected on different days: this was to ensure that the performance of the test computer had not changed significantly from when the typing lattice experiments were run.
 
@@
 
-For each benchmark we highlighted one or two type annotations that appear to show a pattern for the typing lattice performance scatter plots, or in the case of SpectralNorm and Json had higher than usual columns.
+For each benchmark we highlighted one or two type annotations that seem to show a pattern for the typing lattice performance scatter plots, or in the case of SpectralNorm and Json had higher than usual columns.
 The identified types are those represented by the red and blue columns.
@@
 As can be seen, most of the time the patterns we found in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
-For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears to be the cause of the slowest/upper half of the typing lattice graphs.
+For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears responsible for the upper/slowest half of the typing lattice graphs.
 For Permute there is actually a significant performance \emph{increase}, indicated by the lower half of the lattice being highlighted in red.
-DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect does not appear to be cumulative, however: the purple dots are about the same height as the red and blue ones, so \emph{either} annotation is sufficient for the full benefit.
+DeltaBlue and Go also show a slight increase in performance, but here this is caused by two type annotations (in red and blue); this effect is not cumulative, however: the purple dots are about the same height as the red and blue ones, so \emph{either} annotation is sufficient for the full benefit.
 
 List is interesting as it demonstrates that the red and blue type annotations both cause a significant decrease in performance, which is even greater when both are present.
-Richards is particularly odd as it shows a performance decrease that appears to occur when the red type annotation is present, but not the blue one.
+Richards is particularly odd as it shows a performance decrease that occurs when the red type annotation is present, but not the blue one.
 We had previously identified this benchmark and the aforementioned type annotations \cite{roberts-and-co-ecoop-2019} as being the worst case for Moth.
@@
 In particular, observe that the grey dots for Snake, Towers, DeltaBlue, Go, and List are mostly flat horizontal lines, this indicates that by simply deleting the red and blue type annotations, the performance impact of transient type checking becomes negligible.
-However for CD and Richards, the transient type checking overhead of the grey dots appears linear, albeit still less than with the red and blue type annotations present.
+However for CD and Richards, the transient type checking overhead of the grey dots is roughly linear, albeit still less than with the red and blue type annotations present.
 Thus we can observe that usually, only a couple of type annotations are responsible for the overhead caused by Moth's transient typed checking, whereas the rest are ``free''.
@@
 repurposed to determine precisely which transient type checks are
-resistant to the optimisations provided by a just-in-time compiling
-virtual machine.  We observed that many of our benchmark results
+likely resistant to the optimisations provided by a just-in-time compiling
+virtual machine.
+However, detailed analysis will be needed in order to identify exactly what causes such performance effects, such as that we previously did for the List benchmark \cite{roberts-and-co-ecoop-2019}.
+We observed that many of our benchmark results
 conducted under the Takikawa protocol 
-seemed to have a characteristic bimodal performance profile: some
+showed a characteristic bimodal performance profile: some
 of the benchmark configurations ran significantly slower than
@@
 In particular, these patterns suggested the just one or two type annotations are likely responsible for the bimodal performance and most of the overhead caused by Moth's transient type checking.
-Though every type annotation that appeared to have a significant effect across the typing lattice also showed a significant performance effect when no other typing annotations where present, the converse did not hold.
+Though every type annotation that showed a significant effect across the typing lattice also showed a significant performance effect when no other typing annotations where present, the converse did not hold.
 
-This is preliminary work, in particular we have not yet identified exactly why these type annotations appear to have an effect on performance.
+This is preliminary work, in particular we have not yet identified exactly why these type annotations show an effect on performance.
 There are also a number of threats to validity. 
@@
 
+\begin{acks}
+The authors would like to thank the anonymous reviewers for their
+valuable comments and helpful suggestions. This work is supported
+in part by the \grantsponsor{Marsden Fund}{Royal Society of New Zealand (Te Ap\={a}rangi) Marsden Fund (Te P\={u}tea Rangahau a Marsden)}{https://royalsociety.org.nz/what-we-do/funds-and-opportunities/marsden/} under grant \grantnum{Marsden Fund}{VUW1815}.
+\end{acks}
+
 \bibliography{references}

### 2019-09-11 paper.tex
Run-time (at the time of running). Runtime (the engine supporting execution)
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
-Although na\"ive implementations of transient type checks have a high runtime overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.
+Although na\"ive implementations of transient type checks have a high run-time overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.
 Unfortunately
@@
 This has also lead to a technique to evaluate the performance of gradual typing, such as the so called so-called ``Takikawa'' protocol \cite{Takikawa2016,Greenman2019jfp,vitousek-transient-arXive-2019}.
-The Takikawa protocol was created to measure the runtime cost of gradual typing by testing various configurations of typed/untyped code.
+The Takikawa protocol was created to measure the run-time cost of gradual typing by testing various configurations of typed/untyped code.
 This approach was designed to investigate the the performance effect of the amount of typed vs untyped code, in particular it shows that how simply adding more type annotations does not always produce a uniform effect on performance.
@@
 Their focus is typically on parallelisation opportunities,
-guided by runtime feedback, cost models, or heuristics. 
+guided by run-time feedback, cost models, or heuristics. 
 Their large body of work \citep{Isaacs:2014:PerfViz} uses various approaches,
@@
 
-At the moment, our approach to identifing type annotations that case
+At the moment, our approach to identifying type annotations that case
 performance anomalies is not integrated into a development environment.
@@
 programs overall performance. We would also like to investigate 
-whether this approach can assist with optimsations for programmers' day-to-day  development, or help VM engineers identifying performance bugs in the underlying virtual machines.
+whether this approach can assist with optimisations for programmers' day-to-day  development, or help VM engineers identifying performance bugs in the underlying virtual machines.
 

### 2019-09-11 paper.tex
Minor edits in introduction
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
-Unfortunately most gradual systems with run-time semantics (as opposed to type erasure as in TypeScript) typically impose significant run-time performance overheads.
+Unfortunately most gradual systems with run-time semantics, as opposed to type erasure as in TypeScript, impose significant run-time performance overheads.
 This has lead to a significant body of research to develop techniques to optimise gradual typing \citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}.
@@
 The Takikawa protocol was created to measure the run-time cost of gradual typing by testing various configurations of typed/untyped code.
-This approach was designed to investigate the the performance effect of the amount of typed vs untyped code, in particular it shows that how simply adding more type annotations does not always produce a uniform effect on performance.
+This approach was designed to characterise how the amount of typed and untyped code influences performance.
+In particular it shows that how simply adding more type annotations does not always produce a uniform effect on performance.
 Here we adapt their approach in order to identify individual type annotations that may be responsible for performance effects.

### 2019-09-11 paper.tex
Some edits sec. 3.1
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 the number of modules.
-The original Takikawa protocol also suggested a sampling strategy, where $10N$ configurations are randomly chosen from the lattice, however the lattice is a binomial distribution, meaning the majority of chosen benchmarks will have around $N/2$ type annotations
-Grace, following other languages such as
-Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
-allows programmers to choose whether each individual declaration should be type-checked. This means $N$ in Grace is the number of type annotations in the program, and so
-checking an entire lattice for even a moderately sized benchmark would be infeasible.  Vitousek et al.\ modified the Takikawa protocol for
+The original Takikawa protocol also suggested a sampling strategy, where $10N$ configurations are randomly chosen from the lattice, however the lattice is a binomial distribution, meaning the majority of chosen benchmarks will have around $N/2$ type annotations.
+
+Grace allows programmers to choose for each individual declaration whether it should be type-checked, and thus follows languages such as
+Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017}.
+This means in Grace $N$ is the number of type annotations in the program.
+This makes it infeasible to checking an entire lattice, even for a moderately sized benchmark.
+Vitousek et al.\ therefore modified the Takikawa protocol for
 these kinds of languages by taking an approach based on sampling
@@
 
-Unlike such prior work, we wish to identify which type annotations cause anomalies,
+Unlike prior work, we wish to identify which type annotations cause anomalies,
 % To investigate Moth's support for transient type checks further, 
-thus we adapted the Takikawa protocol
+and thus we adapted the Takikawa protocol
 and took inspiration from the Takikawa-Vitousek variant.

### 2019-09-11 paper.tex
Edits sec. 3.3
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 \subsection{Experimental Set Up}
-As in our previous work \cite{roberts-and-co-ecoop-2019} which used the same set of benchmarks, to account for the complex warmup behaviour of modern systems \citep{Barrett:2017:VMW} as well as
+To account for the complex warmup behaviour of modern systems \citep{Barrett:2017:VMW} as well as
 the non-determinism caused by \eg garbage collection and cache effects, we ran each benchmark for 1000 iterations in the same invocation of Moth, and discard the first 350 iterations to ignore warmup JIT compilation.
@@
 
-In our previous work we also compared the performance of untyped code on Moth against state-of-the-art VMs: Java, Node.js using the V8 JavaScript VM, and the Higgs JavaScript VM.
+In previous work \cite{roberts-and-co-ecoop-2019}, we also compared the performance of untyped code on Moth against state-of-the-art VMs: Java, Node.js using the V8 JavaScript VM, and the Higgs JavaScript VM.
 Java was the fastest of these, and on average V8 was about 1.8x slower than Java, Moth was 2.3x slower, and Higgs was 10.4x slower.
-Hence we continue to argue that ``Moth is a suitable platform to assess the impact of our approach to gradual type checking, because its performance is close enough to state-of-the-art VMs, and run-time overhead is not hidden by slow baseline performance.''
+We believe, this makes Moth suitable for assessing the impact of type checking,
+because Moth's performance is close enough to state-of-the-art VMs,
+which avoids hiding of type check overheads in a slow baseline.
 

### 2019-09-11 paper.tex
Added paragraph heading to enable reviewer 2 to find the refined protocol
Signed-off-by: Stefan Marr <git@stefan-marr.de>
Trying to be a bit more correct
Signed-off-by: Stefan Marr <git@stefan-marr.de>

@@
 
+\paragraph{Refined Takikawa Protocol.}
 Unlike prior work, we wish to identify which type annotations cause anomalies,
@@
 Of particular note is that some of the graphs (Permute, Storage, Towers, and Richards) show bimodal performance profiles, that is, two separate roughly-horizontal lines.
-Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that cannot be optimised away.
+Presumably Moth can remove all the overhead from some configurations, but in others there must be one or more type checks that could not be optimised away.
 List shows three performance modes: the graph consists of three mostly flat lines, at about 1 times slower, 1.5 times slower, and 1.8 times slower than untyped code.
@@
 repurposed to determine precisely which transient type checks are
-likely resistant to the optimisations provided by a just-in-time compiling
-virtual machine.
+likely to cause performance effects,
+and for instance are currently not optimised by the just-in-time compiler.
 However, detailed analysis will be needed in order to identify exactly what causes such performance effects, such as that we previously did for the List benchmark \cite{roberts-and-co-ecoop-2019}.

### 2019-09-12 paper.tex
J's minor tweaks

@@
 
-There is a vast spectrum of different approaches to gradual typing \cite{kafka18,bensurvey18icfp}.
-Here we measure the performance of ``transient'' or ``type-tag'' checks (as in Reticulated Python), which offer first-order semantics: they check that an object's type constructor or names of supported methods match any static types that the object flows through, but not the return/argument types of such methods \cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
+There is a vast spectrum of different approaches to gradual typing
+\cite{kafka18,bensurvey18icfp}. Here we measure the performance of
+``transient'' or ``type-tag'' checks (as in Reticulated Python), which
+offer first-order semantics: they check that an object's type
+constructor or names of supported methods match any static types that
+the object flows through, but not the return types or argument types of those methods \cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
 
@@
 
-This has also lead to a technique to evaluate the performance of gradual typing, such as the so called so-called ``Takikawa'' protocol \cite{Takikawa2016,Greenman2019jfp,vitousek-transient-arXive-2019}.
-The Takikawa protocol was created to measure the run-time cost of gradual typing by testing various configurations of typed/untyped code.
+This has also lead to a technique to evaluate the performance of gradual typing, the ``Takikawa'' protocol \cite{Takikawa2016,Greenman2019jfp,vitousek-transient-arXive-2019}.
+The Takikawa protocol was created to measure the run-time cost of gradual typing by testing various configurations of typed and untyped code.
 This approach was designed to characterise how the amount of typed and untyped code influences performance.
-In particular it shows that how simply adding more type annotations does not always produce a uniform effect on performance.
+In particular, Takikawa protocol evaluations often show that how simply adding more type annotations does not always produce a uniform effect on performance.
 Here we adapt their approach in order to identify individual type annotations that may be responsible for performance effects.
@@
 
-Grace allows programmers to choose for each individual declaration whether it should be type-checked, and thus follows languages such as
+Grace allows programmers to choose whether each individual declaration should be type-checked, and thus follows languages such as
 Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017}.
-This means in Grace $N$ is the number of type annotations in the program.
-This makes it infeasible to checking an entire lattice, even for a moderately sized benchmark.
+This means in Grace $N$ is the number of type annotations in the program, so it is
+infeasible to check an entire lattice, even for a moderately sized benchmark.
 Vitousek et al.\ therefore modified the Takikawa protocol for
@@
 
-Each benchmark has an almost complete set of type annotations,
-\ie, static type annotations are present at almost all possible points.
+%Each benchmark has an almost complete set of type annotations,
+%\ie, static type annotations are present at almost all possible points.
 %ISAAC: I changed this as it was misleading, in a few cases, some of the type annotations are in fact "Unknown", which is exactly what my "removing" of type annotations places
+
+We added type annotations to every benchmark, aiming to use the most appropriate (most specific) annotation for each declaration.
+%KJX: I changed it too because I though that was also misleading.
 
@@
 Java was the fastest of these, and on average V8 was about 1.8x slower than Java, Moth was 2.3x slower, and Higgs was 10.4x slower.
-We believe, this makes Moth suitable for assessing the impact of type checking,
+We believe this makes Moth suitable for assessing the impact of type checking,
 because Moth's performance is close enough to state-of-the-art VMs,
-which avoids hiding of type check overheads in a slow baseline.
+which should make it harder to hide type checking overheads in a slow baseline.
 
@@
 reasonable heuristic estimator in most cases, and we will
-examine further the outlying benchmarks.
+examine the outlying benchmarks further.
 
@@
 
-The column graphs, right of the corresponding scatter plot, shows the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
+The column graphs, to the right of the corresponding scatter plot, show the execution time of configurations with only a single type annotation, the x-axis indicates the index of this annotation, thus the first column represents the first annotation, and the last column represents the last one (in the order they appear in the source code).
 The y-axes for the column graphs are the same as the associated scatter graphs.
@@
 
-As can be seen, most of the time the patterns we found in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
+As can be seen, most of the patterns we found in the typing lattice graphs correspond to outliers in the single type annotation column graphs.
 For the Snake, Towers and CD benchmarks, there is clearly a pattern where an individual type annotation (highlighted in red) appears responsible for the upper/slowest half of the typing lattice graphs.
@@
 and for instance are currently not optimised by the just-in-time compiler.
-However, detailed analysis will be needed in order to identify exactly what causes such performance effects, such as that we previously did for the List benchmark \cite{roberts-and-co-ecoop-2019}.
+However, detailed analysis will be needed in order to identify exactly what causes such performance effects: we previously undertook such an analysis for the List benchmark \cite{roberts-and-co-ecoop-2019}.
+
 We observed that many of our benchmark results

### 2019-09-12 paper.tex
Camera Ready

@@
 
-\usepackage{collab}
+%\usepackage{collab}
 
-\collabAuthor{rr}{green!60!black}{Richard}
-\collabAuthor{sm}{red}{Stefan}
-\collabAuthor{kjx}{orange}{James}
-\collabAuthor{isaac}{blue}{Isaac}
-\collabAuthor{mwh}{purple}{Michael}
+%\collabAuthor{rr}{green!60!black}{Richard}
+%\collabAuthor{sm}{red}{Stefan}
+%\collabAuthor{kjx}{orange}{James}
+%\collabAuthor{isaac}{blue}{Isaac}
+%\collabAuthor{mwh}{purple}{Michael}
 
@@
 
-\input{graphs/preamble}
-\setcopyright{none}
+\input{preamble}
 \bibliographystyle{ACM-Reference-Format}
@@
 ensure it has the methods required by the annotation. 
-
-Although na\"ive implementations of transient type checks have a high run-time overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.
-Unfortunately
-the improvement is not uniform: while most type checks can be optimised
-away, some will significantly decrease a program's
-performance, and some may even increase it.
+Just-in-time compilation and optimisation in virtual machines can eliminate much of the overhead of run-time transient type checks. 
+Unfortunately this optimisation is not uniform: some type checks will significantly decrease, or even increase, a program's performance.
 
@@
 
-Using our technique programmers can optimise programs by removing expensive type checks, and VM engineers can identify new opportunities for compiler optimisation.
+Using our technique programmers can optimise programs by removing expensive type checks, and VM engineers can identify new opportunities for compiler optimisation.\pagebreak
 \end{abstract}
@@
 %% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
+\begin{CCSXML}
+<ccs2012>
+<concept>
+<concept_id>10011007.10010940.10011003.10011002</concept_id>
+<concept_desc>Software and its engineering~Software performance</concept_desc>
+<concept_significance>500</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011008.10011009.10011011</concept_id>
+<concept_desc>Software and its engineering~Object oriented languages</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+<concept>
+<concept_id>10011007.10011006.10011041.10011044</concept_id>
+<concept_desc>Software and its engineering~Just-in-time compilers</concept_desc>
+<concept_significance>300</concept_significance>
+</concept>
+</ccs2012>
+\end{CCSXML}
+
 \ccsdesc[500]{Software and its engineering~Software performance}
@@
 This approach was designed to characterise how the amount of typed and untyped code influences performance.
-In particular, Takikawa protocol evaluations often show that how simply adding more type annotations does not always produce a uniform effect on performance.
+In particular, Takikawa protocol evaluations often show that simply adding more type annotations does not always produce a uniform effect on performance.
 Here we adapt their approach in order to identify individual type annotations that may be responsible for performance effects.
@@
 Vitousek et al.\ therefore modified the Takikawa protocol for
-these kinds of languages by taking an approach based on sampling
+these kinds of languages by using a different form of sampling
 \cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
@@
 
-We added type annotations to every benchmark, aiming to use the most appropriate (most specific) annotation for each declaration.
+We added type annotations to every benchmark, aiming to use the most appropriate (\ie most specific) annotation for each declaration.
 %KJX: I changed it too because I though that was also misleading.
@@
 the non-determinism caused by \eg garbage collection and cache effects, we ran each benchmark for 1000 iterations in the same invocation of Moth, and discard the first 350 iterations to ignore warmup JIT compilation.
+Our previous work  identified this as a suitable cut off \cite{roberts-and-co-ecoop-2019}.
 
@@
 
-In previous work \cite{roberts-and-co-ecoop-2019}, we also compared the performance of untyped code on Moth against state-of-the-art VMs: Java, Node.js using the V8 JavaScript VM, and the Higgs JavaScript VM.
+In our previous work \cite{roberts-and-co-ecoop-2019}, we also compared the performance of untyped code on Moth against state-of-the-art VMs: Java, Node.js using the V8 JavaScript VM, and the Higgs JavaScript VM.
 Java was the fastest of these, and on average V8 was about 1.8x slower than Java, Moth was 2.3x slower, and Higgs was 10.4x slower.
@@
 \begin{figure*}
-	\input{graphs/full-scatter}
+	\input{full-scatter}
 	\caption{Graphs of (at most) 102 configurations in the typing lattices for each benchmark.
@@
 \begin{figure*}
-	\input{graphs/pattern}
+	\input{pattern}
 	\caption{Pairs of colour coded scatter and column graphs.
@@
 
-At the moment, our approach to identifying type annotations that case
+At the moment, our approach to identifying type annotations that cause
 performance anomalies is not integrated into a development environment.
@@
 likely to cause performance effects,
-and for instance are currently not optimised by the just-in-time compiler.
+and are currently not optimised away by the just-in-time compiler.
 However, detailed analysis will be needed in order to identify exactly what causes such performance effects: we previously undertook such an analysis for the List benchmark \cite{roberts-and-co-ecoop-2019}.

### 2019-09-12 arxiv.sh
Added stupid arxiv.org script

@@ -1,0 +1,19 @@
+#!/usr/bin/bash
+# Updates the paper to make it suitable for positing to arxiv.org
+T=$(mktemp)
+latexmk # for bbl file
+mkdir -p arxiv
+cp acmart.cls arxiv/ # files we need
+cp paper.bbl arxiv/ms.bbl
+
+latexpand --empty-comments paper.tex --output arxiv/ms.tex
+sed -i 's/]{acmart}/,authorversion=true]{acmart}\n\\pdfoutput=1\n\\settopmatter{printfolios=true}\n/' arxiv/ms.tex # Put appropriate copyright statement, force use of PDFLatex, and add page numbers
+
+# Copy the pgfplots library (this is needed since arxiv.org has an out of date version)
+find $(dirname $(kpsewhich pgfplots.sty)) -type f -exec cp {} arxiv/ \;
+find $(dirname $(kpsewhich pgfplots.code.tex)) -type f -exec cp {} arxiv/ \;
+
+# Now make the zip
+cd arxiv
+rm -f ../arxiv.zip &> /dev/null
+zip ../arxiv.zip ./*
